Silicone Wire Embolization-induced Acute Retinal Artery Ischemia and Reperfusion Model in Mouse: Gene Expression Provide Insight into Pathological Processes

Version 1

Abstract

Abstract

Acute retinal ischemia and ischemia-reperfusion injury are primary causes of retinal neural cell death and vision loss in retinal artery occlusion (RAO). The absence of an accurate mouse model simulating the retinal ischemic process has hampered progress in developing neuroprotective agents for RAO. A unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model was developed by employing silicone wire embolization combined with carotid artery ligation. The survival of retinal ganglion cells and visual function were evaluated to determine ischemia duration. Immunofluorescence staining, optical coherence tomography, and hematoxylin and eosin staining were utilized to assess changes in major classes of neural cells and retinal structure degeneration at two reperfusion durations. Transcriptomics was employed to investigate alterations in the pathological process of UPOAO following ischemia and reperfusion, highlighting transcriptomic differences between UPOAO and other retinal ischemia-reperfusion models. The UPOAO model successfully replicated the acute interruption of retinal blood supply seen in RAO. 60-minute ischemia was confirmed to lead the major retinal neural cells loss and visual function impairment. Notable thinning of the inner layer of the retina, especially the ganglion cell layer, was evident post-UPOAO. Temporal transcriptome analysis revealed various pathophysiological processes related to immune cell migration, oxidative stress, and immune inflammation during non-reperfusion and reperfusion periods. The resident microglia within the retina and peripheral leukocytes which access to the retina were pronounced increased on reperfusion periods. Comparison of differentially expressed genes between the UPOAO and high intraocular pressure models identified specific enrichments in lipid and steroid metabolism-related genes in the UPOAO model. The UPOAO model emerges as a novel tool for the screening of pathogenic genes, promoting further therapeutic research in RAO.

Introduction

Retinal artery occlusion (RAO) is a severe ophthalmic disease characterized by a sudden interruption of blood flow in the retinal artery, leading to retinal ischemia [1]. More than 60% of RAO patients experience impaired vision, ranging from finger counting to complete vision loss [2]. RAO patients face an elevated risk of cardiovascular and cerebrovascular events [3-5]. Retinal ischemia and hypoxia can cause irreversible damage to retinal cells in 90 minutes [6]. This damage is possibly attributed to the continued apoptosis of retinal ganglion cells (RGCs), triggered by inflammation and oxidative stress. Unfortunately, conservative treatments for RAO, such as eye massage and hyperbaric oxygen therapy, offer limited therapeutic benefits [7]. Thrombolysis, though effective, is also restricted by the narrow treatment time window (often within 4.5 hours) [8]. However, when blood flow is restored to the ischemic area after thrombolysis, it can cause a sudden elevation of tissue oxidative levels, resulting in ischemia-reperfusion injury (IRI) in the retina [9]. IRI is a common pathological condition that can trigger inflammation, retinal tissue injury and visual impairment [10]. Therefore, to comprehensively investigate the pathophysiological changes and explore potential neuroprotective therapeutics after retinal ischemia and reperfusion, it is of paramount importance to develop an animal model that can accurately simulate the pathological processes of RAO.

Various animal models have been employed to investigate the effects of retinal injury resulting from ischemia and subsequent reperfusion in RAO [6, 11, 12]. Based on the methods used for modeling, retinal ischemia-reperfusion models can be categorized into two groups: 1) Intravascular occlusion models, including photochemical-induced thrombosis model and vascular intervention model; 2) Extravascular occlusion models, including central retinal artery ligation (CRAL) model, unilateral common carotid artery occlusion (UCCAO) model, and high intraocular pressure (HIOP) model. The photochemical-induced thrombosis model and vascular intervention model have been reported as useful tools to assess retinal injury after ischemia and reperfusion. However, both models are limited by the need for trained interventional radiologists and high technique [13, 14]. CRAL, which directly clips or ligates the central retinal artery, also presents its drawbacks. This model may cause optic nerve damage, and it is difficult to apply in small animals like mice, which limits its use in experiments with larger sample sizes [6]. UCCAO, a retinal hypoperfusion model induced by occlusion of the unilateral common carotid artery (CCA), is better at simulating chronic ischemic retinal disease rather than acute retinal ischemia [15, 16]. The HIOP model is a widely utilized mouse model for studying ischemia-reperfusion in acute primary angle-closure glaucoma (APACG) [17]. However, it is worth noting that the mechanical compression of the retina caused by saline injection may also lead to retinal damage, potentially influencing IRI research. Therefore, it is critically important to develop a simple and low-skill-required mouse model that can simulate the acute retinal ischemia and reperfusion injury in RAO patients.

To better simulate the retinal ischemic process and possible IRI following RAO, we developed a novel vascular-associated mouse model called unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) model. In this model, we employed silicone wire embolization and carotid artery ligation to completely block the blood supply to the retina. We characterized the major classes of retinal neural cells and evaluated visual function following different durations of ischemia (30 minutes and 60 minutes) and reperfusion (3 days and 7 days) after UPOAO. Additionally, we utilized transcriptomics to investigate the transcriptional changes and elucidate changes in the pathophysiological process of the UPOAO model after ischemia and reperfusion. Lastly, we underscored the transcriptomic differences between the model and other retinal ischemic-reperfusion models, including HIOP and UCCAO, and revealed its unique pathological processes similar to that of retinal IRI in RAO. The UPOAO model offers new insights into the mechanisms and pathways involved in ischemia and reperfusion studies of RAO, providing a foundation for studying the protective strategy for ocular ischemic diseases.

Methods and materials

Animals

Eight-week-old male C57BL/6 mice were used in the experiments. Only male mice were used to exclude the potential influence of estrogenic hormones. The mice were provided with sufficient food and water and were maintained on a 12-hour dark-light cycle in a room with regulated temperature conditions. All experimental procedures were designed and conducted according to the ethical guidelines outlined in the ARVO Statement for the Use of Animals in Ophthalmic and Vision Research. The study protocol and methods received approval from the Experimental Animal Ethics Committee of Renmin Hospital of Wuhan University (approval number: WDRM-20220305A).

Preparation work and surgical procedure of UPOAO model

Eight-week-old male C57BL/6 mice weighing 20-25 grams were used in this study. An isoflurane-based anesthesia system was prepared to induce and maintain general anesthesia in mice during surgical procedures. The mice were anesthetized with a 1.5-2% concentration of isoflurane delivered in a mixture of nitrous oxide and oxygen via rubber tubing. Body temperature was maintained at 37±0.5°C throughout the surgery. Surgical instruments were sterilized with 75% ethanol before use to ensure sterility.

The mouse was gently positioned in a supine posture on a heating blanket and its neck was exposed. The preparation involved depilation of the neck area, followed by skin disinfection, and finally, a midline incision along the neck was made. Following the neck gland separation with two tweezers, a careful blunt dissection was employed to separate the left CCA, internal carotid artery (ICA) and external carotid artery (ECA), avoiding compression of nearby nerves and veins (Fig 1A) (Video 1). Then, the CCA and ICA were ligated with a 6-0 suture. Secure a knot using a 6-0 suture at the distal end of ECA and create a slipknot at the proximal end (Fig 1B). Utilize an ophthalmic scissor to create a small inverted “V”-shaped incision between the two suture knots on the ECA. Insert a specialized silicone wire embolus of which length is 7±0.1 mm and diameter is 0.19±0.1 mm through the incision, directing it into the ECA and further into the CCA (Fig 1C). Next, the ligation on the ICA was removed, and the ECA was cut at the point where the artery had been previously incised with scissors. Subsequently, the silicone wire embolus was retracted to the bifurcation of the CCA, rotated counterclockwise and inserted into the ICA and further into the pterygopalatine artery (PPA) (Fig 1D1). The silicone tail of the silicone wire embolus was near the bifurcation of CCA (Fig 1D2), effectively obstructing the ophthalmic artery (OA) (Video 2). Then fasten the slipknot and suture the skin. The mouse was free to move around during arterial embolization.

Modeling Procedure, Validation, and Cervical Artery Anatomy.

(A1-G1) Schematic illustration of unilateral pterygopalatine ophthalmic artery occlusion (UPOAO). (A2-G2) Practical operation of UPOAO. (A1, A2) Blunt separation and left cervical arteries exposure. (B1, B2) Arterial suture ligation. (C1, C2) Silicone wire embolus insertion. The artery was incised a hole and the silicone wire embolus was inserted. (D1, D2) Artery disconnection and movement of the silicone wire embolus. The artery was cut along the hole and the silicone wire embolus was retracted and reinserted. (E1, E2) Removal of the silicone wire embolus and reperfusion. After some time of ischemia, the silicone wire embolus was removed. (F1, F2) Removal of sutures. The sutures at both ends of the disconnected vessel were knotted, and other two sutures were removed. (G1, G2) Anatomic reduction and sutures of the skin. (H and I) Flat-mounted retina after perfusing rhodamine-labeled canavalin A into the heart of a UPOAO mouse. The sham eye was an unpracticed control eye, and the UPOAO lateral eye was the experimental eye. Retinal vessels in the sham eye (H) were filled with fluorescence while retinal vessels in the UPOAO lateral eye (I) were unfilled. Scale bar = 1 mm. (J and K) Fluorescein Fundus Angiography (FFA) of a UPOAO modeling mouse. The retina vessels (J) were perfused while UPOAO lateral retinal perfusion (K) was delayed. (L) The schematic illustration of cervical artery anatomy and ocular blood supply. Embolization of PPA led to ocular ischemia. The red arrow indicates the embolism site of the silicone wire embolus. The silicone wire embolus is the type 602156 wire embolus extended to 7mm and has a diameter of 0.19mm. The blue arrows indicate the molding locations of the HIOP model and the UCCAO model, respectively. CCA: common carotid artery; ICA: internal carotid artery; ECA: external carotid artery; PPA: pterygopalatine artery; MCA: middle cerebral artery; Post. Sup. Alveolar art.: Posterior superior alveolar artery; Infraorbital art.: infraorbital artery; OA: ophthalmic artery; SPCA: short posterior ciliary artery; LPCA: long posterior ciliary artery; CRA: central retinal artery.

Following a predetermined ischemia embolization time, the silicone wire embolus was carefully removed from the PPA without massive hemorrhage (Fig 1E). The CCA suture was removed and the arterial reperfusion was restored (Fig 1F) (Video 3). Then suture the skin wound and routinely feed mice during reperfusion (Fig 1G).

Staining of retinal vessels

To clearly visualize the retinal vessels, rhodamine-labeled canavalin A was used for mouse heart perfusion. The mouse was anesthetized by 1% sodium pentobarbital and placed in a supine position on a foam board resting on a tray. The mouse’s sternum was opened, exposing the heart. A needle attached to the perfusion tube was inserted into the left ventricle via the apex of the mouse heart and fixed. The right atrial appendage of the mouse was dissected with scissors to allow blood outflow. Saline solution was perfused through the left ventricle using a pump system to drain the blood from arteries. Subsequently, prepared solution of rhodamine labeled canavalin A was injected into the mice and circulated throughout the mouse vessels for 30 minutes. The mice were sacrificed and their eyes were extracted and immersed in PFA for one hour in the dark. Obtain a flat-mounted retina and capture retinal images using the Leica SP8 confocal microscope equipped with a 10× objective.

Quantification of RGCs and microglia

The mice were euthanized by cervical dislocation and their eyes were fixed in 4% paraformaldehyde (PFA) for 60 minutes at room temperature (RT). Subsequently, the cornea and lense were excised, and the retina was isolated intact to perform retinal flat-mounting. The retina was immersed in PBS with 5% Bovine Serum Albumin (BSA) and 0.5% Triton X-100 for overnight blocking and then incubated with Brn3a antibody (Santa Cruz Biotechnology, Dallas, Texas, USA) or Iba1 antibody (Wako, Japan). Following a two-day incubation with the primary antibody, the retina was gently washed three times with PBS. Subsequently, the retina was incubated with Alexa Fluor 594 (AntGene, Wuhan, China) in a light-protected cassette for two days. After three rinses, the retina was uniformly sectioned into a four-leaf clover morphology and flattened using a coverslip. Finally, retinal flat-mounts labeled with Brn3a were imaged using the Leica SP8 confocal microscope (Leica TCS SP8, Germany) equipped with 20× objective, while those labeled with Iba1 were imaged utilizing a fluorescence microscope (BX63; Olympus, Tokyo, Japan). Each quadrant was consistently subdivided into central, middle and peripheral fields at regular intervals. The surviving RGCs and activated microglia were quantified and averaged using Image-J software (National Institutes of Health, USA).

Electroretinogram (ERG)

Mice were dark-adapted overnight before ERG and all subsequent procedures were carried out in the dark environment. Before ERG, mice were anesthetized by 1% sodium pentobarbital through intraperitoneal injection and their pupils were dilated. A subcutaneous electrode was inserted into the posterior cervical skin, a tail electrode was affixed to the posterior end of the mouse tail, and the corneal electrode was gently placed on the central corneal surface. The RetiMINER-C visual electrophysiological system (3VMED Co., Ltd., Shanghai, China) was employed for recording electrical responses. The a-waves, b-waves, and Oscillatory Potentials (OPs)-waves after flash stimuli of 0.01, 0.03, 0.1, 0.3, 1.0, 3.0, and 10.0 cd.s/m2 in scotopic adaptation were recorded. The amplitudes and implicit times of a-waves, b-waves and OPs-waves in response to various flash stimuli were analyzed.

Optical Coherence Tomography Imaging (OCT) and Fluorescein Fundus Angiography (FFA)

The mice were placed on the platform of a Spectralis HRA+OCT device (Heidelberg Engineering, Heidelberg, Germany) for OCT imaging following anesthesia. Pupil dilation was performed, and normal saline was applied regularly to maintain corneal moisture in the mice. The focal length of the device was adjusted so that the mouse retina was clearly visible. Mouse’s head was gently repositioned to take pictures of the peripheral fundus. Multi-line mode was used to scan each layer of the retina and four-quadrant fields of view centered on the nipple on the upper left, lower left, upper right, and lower right were recorded. In this study, IMRL (inner middle retina layer) was defined as the combined thickness of the retinal nerve fiber layer (RNFL), ganglion cell layer (GCL), and the inner plexiform layer (IPL). MRL (middle retina layer) was defined as the combined thickness of the inner nuclear layer (INL) and the outer plexiform layer (OPL). ORL (outer retina layer) was defined as the combined thickness of the rest to the retinal pigment epithelium (RPE). Total retinal thickness (TRT) of the sensory retina was manually segmented and encompassed the entire thickness from the nerve fiber layer to the photoreceptor layer. The thickness of retina at 1.5 papillary diameters (PD), 3.0 PD, and 4.5 PD from the center of the optic disc were measured using the Heidelberg measuring tool.

For FFA, mice were anesthetized and their pupils were also dilated. Subsequently, fundus angiography was recorded immediately following intraperitoneal injection of sodium fluorescein (for no longer than 5 seconds), with photos taken alternately for both eyes.

Hematoxylin and eosin (HE)

Mouse eyes were fixed in FAS eyeball fixation solution (Service-bio, Wuhan, China) for 48 hours, followed by dehydration and subsequent embedding in paraffin. Paraffin was trimmed parallel to the optic nerve to obtain 3-4 μm thick sections where the optic nerve was located. Six paraffin sections from each eyeball were stained with HE. The Olympus fluorescence microscope was employed for taking pictures. Retinal thickness within 500 μm of the optic disc was measured using the Image-Pro Plus 6.0 software.

Staining of Retinal Sections

For immunofluorescence, mouse eyes were fixed in 4% PFA solution and subjected to gradient dehydration using 10%, 20%, and 30% sucrose solutions. The following day, the eyes were embedded in optimum cutting temperature (OCT.) compound (SAKURA, American) and frozen at -80°C. Several 14-μm-thick frozen sections were obtained from each eyeball through the frozen microtome (Leica, Wetzlar, Germany). The section surface was parallel to the optic nerve, and 3-6 frozen sections were mounted on a single slide. The slides were stored at -20°C until use.

All sections were blocked with 5% BSA and 0.5% Triton X-100 in PBS for 2 hours. Primary antibodies, as listed in Table 1, were diluted in PBS containing 5% BSA and 0.5% Triton X-100 and incubated at 4°C overnight. The following day, after three rinses, the sections were incubated with the second antibodies at RT for 2 hours in a cassette, followed by three washes. And then sections were stained with DAPI (Service-bio, Wuhan, China) for 15 minutes. After the final three rinses, the frozen sections were sealed. A Leica SP8 equipped with 40× objective was used to photograph frozen sections near the optic nerve head. The fluorescence intensity of all slices of each eye was mounted using Image-J.

Surgical procedure of UCCAO model and HIOP model

Animals for UCCAO were prepared and managed as described in the UPOAO model procedure. After blunt dissection, the left CCA was ligated by a 6-0 suture. After 60 minutesligation, the suture was removed and the skin wound was subsequently closed with sutures. Mice were routinely fed during reperfusion and mouse retinas of UCCAO were harvested 7 days post-surgery.

HIOP model was performed based on previous methods [18]. Briefly, mice were anesthetized by intraperitoneal injection of 1% sodium pentobarbital. Mice pupils were dilated using Tropicamide Phenylephrine Eye Drops (0.5% tropicamide and 0.5% deoxyadrenaline hydrochloride, Santen Pharmaceutical) 5 minutes in advance. An insulin hypodermic needle, attached to a silicone elastomer tube, was inserted into the anterior chamber of the unilateral experimental eye. Elevate intraocular pressure (IOP) for 60 minutes by raising the height of the saline solution storage bag to 162 cm and remove the needle to immediately restore IOP to baseline levels. Iris whitening and loss of retinal red-light reflexes demonstrated retinal ischemia. After surgery, levofloxacin hydrochloride ophthalmic gel was applied to the eyes of mice. Mouse retinas of HIOP were collected 7 days post-surgery.

Transcriptome sequence and analysis

Retinal samples from UPOAO, HIOP, and UCCAO mice were extracted and promptly frozen in liquid nitrogen within 10 minutes of cervical dislocation. Total RNA extraction followed the protocol outlined in the TRIzol Reagent manual (Life Technologies, CA, USA). RNA integrity and concentration were assessed using an Agilent 2,100 Bioanalyzer (Agilent Technologies, Inc, Santa Clara, CA, USA). The resulting RNA samples were then pooled and subjected to sequencing on the Illumina HiSeq3000 platform in a 150-bp paired-end read format. Raw RNA-sequencing (RNA-seq) reads underwent preprocessing and quantification using the featureCounts function in the SubReads package version 1.5.3, with default parameters.

RNA-seq analysis was repeated across 5 mice in the no-perfusion group, 4 mice in the 3-day perfusion group, and 5 mice in the 7-day perfusion group of the UPOAO model. Additionally, retinas from both the 7-day HIOP experimental eyes and UCCAO bilateral eyes were collected (n=5) for analysis.

Data normalization and subsequent processing utilized the ‘limma’ package in R software (version 4.2.3) [19]. For visualization purposes, volcano plots and Venn diagrams were generated to identify significant differential expression genes (DEGs) and perform Gene Ontology (GO) annotation analysis. DEGs with |log2 fold change (FC)| >1 and adjusted p-value <0.05 were deemed statistically significant. Enrichment analysis focused on DEGs with an adjusted p < 0.05 and an enriched gene count >5. To visualize the expression patterns of significant DEGs, a heatmap was generated using the ‘pheatmap’ package. Gene Set Enrichment Analysis (GSEA) was conducted using the GSEA program (version 4.3.2) [20], employing the default gene set m-subset (mh.all.v2023.1.mm.symbols.gmt) to explore significant functional and pathway differences. Enriched pathways were classified based on criteria such as adjusted p-value (<0.05), False Discovery Rate (FDR) q-value (<0.25), and normalized enrichment score (|NES| > 1). Furthermore, protein-protein interaction (PPI) analysis was performed using the Search Tool for the Retrieval of Interacting Genes (STRING) database (https://string-db.org/) [21]. Genes showing a higher degree of protein-level interactions with others were further analyzed using Cytoscape software (version 3.8.2) to generate a downstream PPI map. The resulting PPI pairings with an interaction score > 0.7 were extracted and visualized using Cytoscape 3.9.0 [22].

Real-time Quantitative polymerase chain reaction (RT-qPCR)

RNA was extracted from fresh mouse retinal samples and subjected to reverse transcription using HiScript lll RT SuperMix for qPCR (Vazyme, China) following the manufacturer’s instructions. RT-qPCR analysis was conducted on a CFX instrument (Bio-rad) using AceQ qPCR SYBR Green Master Mix (Vazyme, China). The PCR program comprised 40 cycles of 10 seconds at 95°C and 30 seconds at 60°C. Assays were performed in triplicate, and Ct values were normalized to beta actin levels. Relative quantification of target gene expression was performed using the 2-ΔΔCt method. Primer details are provided in Table S1.

Statistical Analysis

All data were analyzed using GraphPad Prism version 9.0 (GraphPad Software, San Diego, CA, USA). Each experimental group included a minimum of three biological replicates. T-tests were employed for comparing two groups, including RGCs and microglia counting, Ops-waves analysis, retinal thickness measurements in HE staining, and fluorescence intensity quantification. Two-way ANOVA (two-way analysis of variance) test was used for more than two different experimental groups, including ERG waves and retinal thickness measurements in OCT. All data were presented as the mean ± standard error of the mean (s.e.m) and the statistical graphs were depicted as scatter bar graphs and line charts. P < 0.05 was considered statistically significant.

Results

Silicone wire embolus insertion interrupts retinal blood flow

To reproduce the retinal ischemic process and potential IRI in RAO, we established a mouse model of UPOAO by combining silicone wire embolization and ligation of the carotid artery. Model validation was conducted through a series of experiments, with detailed descriptions provided in the Materials and Methods (Fig. 1A-G). To confirm blood flow disruption, fluorescently labeled lectin cardiac perfusion and FFA were performed. Rhodamine-labeled canavalin A was perfused with the silicone wire embolus in vivo, and bilateral eyes were evaluated. The sham eye exhibited normal blood perfusion in the retina, while the experimental eye showed no perfusion (Fig. 1H, I). FFA revealed delayed and limited perfusion in the experimental lateral retina, primarily near the optic disc (Fig. 1J, K). These findings indicate that insertion of the silicone wire embolus resulted in impaired blood flow to the retina.

60-minute ischemia in UPOAO damage retinal structure and function

To investigate the optimal ischemic duration, mice underwent two ischemic periods: 30 and 60 minutes, followed by reperfusion periods of 3 and 7 days. Retinal structure and visual function were assessed through flat-mounted retina analysis and flash electroretinography (ERG).

In the 30-minute ischemia group, no significant RGCs death was observed after 3 days (Fig. 2A, B) or 7 days of reperfusion (Fig. 2C, D). However, with 60 minutes of ischemia and 3 or 7 days of reperfusion, reductions in RGC density were evident (Fig. 2E-H). ERG results showed no statistical difference in amplitudes between bilateral eyes after 3 and 7 days of 30-minute reperfusion (Fig. 3A-C). However, the b-wave amplitude notably declined after 60 minutes of ischemia and 3 days of reperfusion (Fig. 3D). By 7 days of reperfusion, the b-wave amplitude had halved compared to sham eyes (Fig. 3E). The appearance time of the a-wave and b-wave in the experimental and control eyes was consistent within each of the four groups (Fig. S1). Additionally, the amplitudes of OPs in the experimental eyes, particularly in the 60-minute ischemia and 7-day reperfusion group, significantly decreased to less than 50% of those in the sham eyes (Fig. 3F, G). Based on RGCs survival and changes in ERG waves, we determined that a 60-minute ischemic duration is optimal for the UPOAO model.

Staining and quantification of retinal ganglion cells (RGCs) at different ischemia and reperfusion times.

Flat-mounted retina RGCs were labeled with Brn3a staining. (A, B) Representative pictures of peripheral retinal visual field (A), and quantification of surviving RGCs (B) in the 30-minute ischemia and 3-day reperfusion group. N = 3. (C, D) Representative pictures of peripheral retinal visual field (C), and quantification of surviving RGCs (D) in the 30-minute ischemia and 7-day reperfusion group. N = 3. (E, F) Representative pictures of peripheral retinal visual field (E), and quantification of surviving RGCs (F) in the 60-minute ischemia and 3-day reperfusion group. N = 5. (G, H) Representative pictures of peripheral retinal visual field (G), and quantification of surviving RGCs (H) in the 60-minute ischemia and 7-day reperfusion group. N = 5. The results showed the evident RGCs loss after 60-minute ischemia. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t-test. Scale bar = 100 μm.

Comparison of electroretinographic (ERG) dark-adapted responses at different ischemic and reperfusion times.

Following evaluating survival RGCs, visual function between sham vs. UPOAO experimental eyes at different ischemia and reperfusion times was evaluated using ERG. (A) Representative wave lines in four groups at the stimulus light intensity of 0.1, 1.0, and 10.0 cd.s/m2, respectively. (B) Quantification of amplitudes of a-waves and b-waves in the 30-minute ischemia and 3-day reperfusion group. N = 5. (C) Quantification of amplitudes of a-waves and b-waves in the 30-minute ischemia and 7-day reperfusion group. N = 5. (D) Quantification of amplitudes of a-waves, b-waves in the 60-minute ischemia and 3-day reperfusion group. N = 7. (E) Quantification of amplitudes of a-waves and b-waves in the 60-minute ischemia and 7-day reperfusion group. N = 8. Dark-adapted responses were similar in terms of a-wave amplitude but significantly decreased in b-wave amplitude in the 60-minute ischemia groups. The amplitudes of b-waves declined at 3 days and more intensely at 7 days. (F, G) Representative OPs-waves and quantification of amplitudes in the 60-minute ischemic groups. N = 7 in the 3-day reperfusion group; n = 8 in the 7-day reperfusion group. The OPs-wave amplitudes decreased significantly at 7-day reperfusion. The decline of amplitudes in b-waves, OPs-waves, and the loss of RGCs support the selection of a 60-minute ischemic duration as an appropriate choice. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test in a-waves and b-waves; paired t-test in OPs-waves.

Response time for a-waves and b-waves in ERG under different light intensities.

(A, B) Response times of a-waves and b-waves under different light intensities at 3 and 7 days post-UPOAO in the 30-minute ischemia group. (C, D) Response times of a-waves and b-waves under different light intensities at 3 and 7 days post-UPOAO in the 60-minute ischemia group. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test.

Evaluation of retinal thickness in UPOAO through OCT and HE

To assess retinal thickness non-invasively in the UPOAO model, we conducted OCT imaging during 3-day and 7-day reperfusion periods. Due to the challenge of distinguishing the RNFL, GCL, and IPL in OCT images, we utilized the IMRL to represent the cumulative thickness of these layers. Similarly, the MRL represented the INL and OPL, while the ORL encompassed the region from the inner segment/outer segment (IS/OS) to the RPE. Following a 3-day reperfusion period, no significant changes were observed in IMRL, MRL, and ORL at distances of 1.5, 3, and 4.5 PD from the optic disc (Fig. 4B, Fig. S2A, B). However, total retinal thickness decreased at a distance of 4.5 PD (Fig. 4C). After 7 days of reperfusion, total retinal thickness decreased at distances of 1.5 PD, 3 PD, and 4.5 PD, primarily due to inner retinal thinning (Fig. 4D-F, Fig. S2C, D).

Changes in retina morphology of the UPOAO animals.

(A) Representative OCT images of mouse retina at 3 days. Green lines indicated the OCT scan area from the optic disc. IMRL: inner middle retina layer, including RNFL, GCL and IPL layers. MRL: middle retina layer, including INL and OPL layers. ORL: outer retina layer, including IS, OS and RPE layers. TRT: total retina thickness, including all layers of the retina. (B and C) Quantification of IMRL and TRT thickness at 3 days. The thickness of IMRL and TRT in OCT was measured and compared at distances of 1.5 PD, 3.0 PD and 4.5 PD from the optic disc, respectively. N = 5. (D) Representative OCT images of mouse retina at 7 days. (E and F) Quantification of IMRL and TRT thickness at 7 days. N = 5. (G) Representative HE images of mouse retina at 3 days. (H, I and J) Quantification of NFL + GCL, IPL and INL thickness at 3 days. Retinal thickness in HE was measured near the optic nerve head and compared. N = 3. (K) Representative HE images of mouse retina at 7 days. (L, M and N) Quantification of NFL + GCL, IPL and INL thickness at 7 days. N = 3. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test in OCT and paired t-test in HE. PD: papillary diameters; RNFL: retinal nerve fiber layer; GCL: ganglion cell layer; IPL: inner plexiform layer; INL: inner nuclear layer; OPL: outer plexiform layer; IS: inner segment; OS: outer segment; RPE: retinal pigment epithelium. Scale bar = 50 μm.

Quantification of MRL and ORL thickness in OCT at 3-d and 7-d reperfusion of the UPOAO animals.

MRL and ORL thickness in OCT were measured and compared at distances of 1.5 PD, 3.0 PD and 4.5 PD from the optic disc, respectively. (A and B) Quantification of MRL (A) and ORL (B) thickness at 3 days. n = 5. No significant differences. (C and D) Quantification of MRL (C) and ORL (D) thickness at 7 days. N = 5. No significant differences. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test.

For a detailed analysis of inner retinal thickness, we extracted UPOAO mouse eyes and stained them with HE. The thickness of RNFL + GCL increased at 3 days of reperfusion (Fig. 4H), followed by a decrease at 7 days of reperfusion (Fig. 4L). IPL thickness decreased at both 3 days and 7 days of reperfusion (Fig. 4I, M), while INL thickness decreased only at 7 days of reperfusion (Fig. 4N). We hypothesized that the initial swelling of GCL in response to acute ischemia and hypoxia led to early thickening, followed by gradual thinning due to tissue dysfunction. This hypothesis was supported by the observation of RGC loss and reduced IMRL thickness evaluated by OCT at 7 days (Fig. 4E).

We observed that changes in retinal thickness became noticeable at 3 days post-UPOAO and exhibited a significant decrease at 7 days, as observed in OCT. HE results further elucidated that alterations in inner retinal thickness constituted a substantial portion of the total retinal thickness during the early reperfusion stage after UPOAO.

Survival of retinal neural cells in UPOAO

Given that mouse retinal function heavily relies on rod cells, and scotopic (low light) vision is primarily governed by rod photoreceptors, the observed reductions in b-wave and OPs-wave amplitudes indicate impaired synaptic transmission within the inner retina [23]. Our OCT and HE findings suggest structural injuries within the inner retina. Notably, a-waves in ERG and the thickness of the outer retinal layers in both OCT and HE remained unchanged. To delve deeper into these observations, we conducted immunofluorescence staining to investigate alterations in major retinal cell types, primarily focusing on Bipolar cells (BCs), photoreceptor cells, Horizontal cells (HCs), cholinergic amacrine cells.

BCs loss and photoreceptor cells survival

PKCα is a marker that delineates BCs, with their cell bodies primarily located in the outermost part of the INL, axonal terminals extending into the innermost part of the IPL, and dendrites confined to the OPL (Fig 5A, B). In the experimental UPOAO eyes, BCs did not show significant changes at the 3-day reperfusion mark but exhibited dramatic alterations by the 7-day reperfusion period (Fig 5A2, B2). Particularly, the immunostaining density of BCs’ dendritic and axonal arbors notably decreased at the 7-day reperfusion mark. BCs account for approximately 40% of INL cells in mice, and their somata and axonal processes form a substantial part of the INL and IPL, consistent with the thinning of the INL observed in both OCT and HE at 7 days (Fig 4E, N). It’s noteworthy that the decline in PKCα+ BCs at the 7-day reperfusion point coincided with the decrease in b-wave amplitude in ERG (Fig 3E), indicating the functional interplay between BCs and other cells.

Changes of bipolar cells, Horizontal cells and cholinergic amacrine cells in the UPOAO mice.

(A) Representative images of DAPI and PKCα co-staining of mouse retina at 3 days. (B) Representative images of DAPI and PKCα co-staining of mouse retina at 7 days. (C) Quantification of PKCα fluorescence density at 3 days. (D) Quantification of PKCα fluorescence density at 7 days. (E) Representative images of DAPI and ChAT co-staining of mouse retina at 3 days. (F) Representative images of DAPI and ChAT co-staining of mouse retina at 7 days. (G) Quantification of ChAT fluorescence density at 3 days. (H) Quantification of ChAT fluorescence density at 7 days. Data points were from sections of four animals. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t- test. Scale bar = 50 μm.

Recoverin marks the somata and outer segments of photoreceptor cells and is localized within the outer nuclear layer (ONL) (Fig S3A, B). Interestingly, recoverin-positive photoreceptors remained relatively stable throughout the reperfusion periods, which aligns with the a-waves observed in ERG (Fig 3D, E) and the changes seen in the outer retinal layers in OCT (Fig S2C, D).

Changes of photoreceptor cells in the UPOAO mice.

(A) Representative images of DAPI and Recoverin co-staining of mouse retina at 3 days. (B) Representative images of DAPI and Reconerin co-staining of mouse retina at 7 days. Data points were from sections of four animals. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t- test. Scale bar = 50 μm.

HCs and cholinergic amacrine cells loss

We proceeded to evaluate HCs and cholinergic amacrine cells using calbindin and ChAT immunostaining, respectively. Calbindin immunostaining delineated the somata of HCs within the INL, with their terminal axon connections linearly positioned within the OPL (Fig 5E, F). In the UPOAO experimental eyes, a notable reduction in HCs number was observed at both 3-day and 7-day reperfusion periods. By 7 days, a significant decrease in cell body numbers, a considerable reduction in axon density, and disrupted linear connections were evident (Fig 5F2).

ChAT+ cell bodies located in the GCL and INL, with dendrites forming two narrow stratified bands within the IPL. The immunofluorescence density of ChAT+ amacrine cells decreased notably after 3 days and even more prominently after 7 days (Fig 5I, J). The fluorescence intensity of the two bands within the IPL markedly weakened and was nearly invisible by 7 days (Fig 5J2).

Time course transcriptome analysis revealed features of different reperfusion periods in UPOAO

To explore the pathophysiological processes of UPOAO, we extracted retinas of bilateral eyes for transcriptome sequencing. The samples encompassed the 60-minute ischemic eyes without reperfusion, 60-minute ischemia followed by 3-day reperfusion, 60-minute ischemia followed by 7-day reperfusion, which were compared to the sham eye as controls.

In the no-perfusion group, 215 genes were upregulated and 204 genes were downregulated (Fig 6A). GO enrichment analysis revealed that DEGs were related to leukocyte migration, epidermis development, myeloid leukocyte migration and other cell migration pathways (Fig 6B). The heatmap showed the upregulated and downregulated genes of immune cells migration-related pathways at the no-perfusion period in UPOAO (Fig 6C) and high-connectivity DEGs (‘hub genes’) were also enriched in these pathways, such as Dusp1 and Fos (Fig S4A). GSEA also showed similar results (Fig S4B-D). These results suggested early ischemic processions such as cell migration and potential collateral vessel formation.

Transcriptomic features at different times of reperfusion.

RNA-seq evaluation of 0, 3 and 7 days in UPOAO and mostly enriched in the pathways of immune cells migration, oxidative stress and immune inflammation. (A, D, G) Volcano plots showing differential expression genes (DEGs) in the UPOAO and sham eyes in the no-perfusion group (A), 3-day group (D) and 7-day group (G), respectively. Red dots: significant upregulated genes, green and blue dots: significant downregulated genes, grey dots: stable expressed genes, adjusted P < 0.05. log2FC = 1. (B, E, H) Gene ontology (GO) analysis of differential genes in the no-perfusion group (B), 3-day group (E) and 7-day group (H). The overlapping DEGs between UPOAO and Sham eyes were enriched in the pathways of immune cells migration (0d), oxidative stress (3d) and immune inflammation (7d), respectively. (C, F, I) Heatmap showing TOP 150 overlapping DEGs related with pathways of immune cells migration (C), oxidative stress (F) and immune inflammation (I) based on the DEGs between UPOAO and Sham at no-perfusion, 3-day and 7-day reperfusion, respectively. Ranking was determined by the magnitude of fold change. In each heatmap, upper box showed the top 10 up-regulated genes, and the below one showed the top 10 down-regulated genes.

Hub genes in PPI analysis and GSEA analysis at no-reperfusion in UPOAO.

(A) Hub genes in PPI analysis of DEGs at no-reperfusion group. (B-D) GSEA analysis of the pathway of the cell killing (B), leukocyte mediated-immunity (C) and lymphocyte mediated-immunity (D) at no-reperfusion group.

In the 3-day reperfusion group, 372 genes were upregulated and 170 genes were downregulated (Fig 6D). GO enrichment analysis revealed that 3-day reperfusion DEGs were related to energy metabolism, mitochondrial regulation, and oxidative stress pathways (Fig 6E). The heatmap showed the upregulated and downregulated genes of oxidative stress-related pathways at 3-day reperfusion in UPOAO and some hub genes, such as Mrpl18 and Mrps23, were included in the heatmap (Fig 6F, Fig S5A). GSEA showed the pathway of bounding membrane of organelle, cell surface and plasma membrane protein complex changed (Fig S5B). Furthermore, 44 overlapping genes, between mitochondrial genes and DEGs of 3 days post-UPOAO, mainly related to mitochondrial transport and the mitochondrial tricarboxylic acid cycle (Fig S5C, D). These results suggested mitochondria and metabolism were involved in IRI.

Hub genes, GSEA analysis at 3-day reperfusion in UPOAO and relation with mitochondrial genes.

(A) Hub genes in PPI analysis of DEGs at 3-day reperfusion group. (B) GSEA analysis of the pathway of plasma membrane protein complex (top left), bounding membrane of organelle (top right) and cell surface (bottom). (C) Venn diagram of mitochondrial genes and DEGs at 3-day reperfusion (44 genes). (D) GO analysis of 44 overlapping DEGs.

In the 7-day reperfusion group, 429 genes were upregulated and 310 genes were downregulated (Fig 6G). The 7-day reperfusion DEGs were enriched in regulation of immune effector process, positive regulation of response to external stimulus and inflammation (Fig 6H). The heatmap showed the upregulated and downregulated genes of immune inflammation-related pathways at 7-day reperfusion in UPOAO and most hub genes were also enriched in this pathway, such as Cd86, Cd48, Tlr4, Tlr6 (Fig 6I, Fig S6A). GSEA also showed similar results (Fig S6B). What’s else, RT-qPCR confirmed the immune inflammation-related gene expression was upregulated (Fig S8). Significantly, 112 overlapping genes, between the immune-related genes and DEGs at 7 days post-UPOAO, mainly pertain to immune receptor activity, phagocytic vesicle, etc. (Fig S6C, D). These results underscore the predominant role of immune responses during this stage.

To explore the association between 3-d reperfusion and 7-d reperfusion changes, we analyzed the DEGs and found 17 overlapping genes (Fig S7A), mainly associated with the nitric oxide biosynthetic process (Fig S7B). Our finding indicated that the immune inflammatory response observed after 7 days of reperfusion may be the cumulative effect of the acute oxidative stress response during the initial 3-day reperfusion period.

Hub genes, GSEA analysis at 7-day reperfusion in UPOAO and relation with immune genes.

(A) Hub genes in PPI analysis of DEGs at 7-day reperfusion group. (B) GSEA analysis of the pathway of reactome innate immune system (top left), reactome immunoregulatory interactions between a lymphoid and a non-lymphoid cell (top right), reactome immune system (bottom left) and reactome adaptive immune system (bottom right). (C) Venn diagram of immune genes and DEGs at 7-day reperfusion (112 genes). (D) GO analysis of 112 overlapping DEGs.

The co-expressed genes for 3-day and 7-day reperfusion.

(A) Venn diagram of DEGs at 3-day and 7-day reperfusion and list of 17 overlapping DEGs. (B) GO analysis of 17 overlapping DEGs.

Immune inflammation-related gene expression was upregulated in the UPOAO-7d group.

ABCA1, CD86, CD48, TLR4, TLR6, and TSPO genes were significantly upregulated in the UPOAO-7d group, while other immune-inflammatory and chemotaxis-related genes showed no significant differences. The data points are from retina of four animals. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, Paired t-test.

Leukocytes Infiltration and Microglial Activation

Time course transcriptome analysis underscore the predominant role of immune inflammatory responses during the reperfusion stage. Since the inflammatory response in RIRI has been reported was orchestrated by peripheral immune cells and resident immune cells within the retina [24, 25], we demonstrated the presence of leukocyte infiltration and microglial activation in UPOAO model. Immunofluorescent staining for CD45 was performed on retinas at different time points post-modeling to visualize the distribution and quantity of white blood cells (Fig 7A-D). Laser confocal Z-plane projections confirmed minimal leukocyte infiltration in the retinas of sham eyes, which showed CD45+ cells at the vascular lumen likely representing patrolling cells (Fig 7A). A significant increase in CD45+ cells was observed in UPOAO model retinas at 1 day (Fig 7E), with a progressive increase in quantity over time (Fig 7F, G). Notably, the majority of CD45+ cells in UPOAO model retinas at 1 day displayed a morphology similar to that of vascular leukocytes (Fig 7B), suggesting a large influx of peripheral white blood cells into the retinal tissue. In contrast, CD45+ cells at 3-day and 7-day reperfusion exhibited a combination of amoeboid and branched morphologies (Fig 7C, D), indicative of activated states.

Peripheral leukocyte infiltration and retinal resident microglial activation.

Rhodamine-labeled canavalin A was immediately used for cardiac perfusion to mark blood vessels, followed by CD45 immunofluorescent staining to observe the relationship between blood vessels and CD45+ cells in Sham (A), UPOAO-1d (B), 3d (C), and 7d (D). The white arrows indicate the presence of CD45+ cells in the blood vessels. UPOAO-1d (E), 3d (F), 7d (G) whole-retinal CD45+ cell counts were performed. The cellular morphology and distribution of microglial cells in the superficial retina of UPOAO-3d (H) and 7d (I). And cell counting of microglial cells in the superficial retina of UPOAO-3d (J) and 7d (K). Data points are from retinal sections of four animals. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t- test. Scale bar = 50 μm.

Microglia, known to undergo activation and morphological changes in response to various insults, were examined in the superficial retina of sham and UPOAO eyes. Immunofluorescent staining for Iba1 was conducted on flat-mounted retinas at 3-day and 7-day post-reperfusion. In sham eyes, Iba1+ microglia displayed small soma and elongated dendrites, evenly distributed within the retina (Fig 7H1, I1). Conversely, UPOAO retinas exhibited numerous activated Iba1+ microglia with enlarged soma, shortened dendrites, and an ameboid appearance (Fig 7H2, I2). The number of Iba1+ microglia in UPOAO retinas was approximately five times higher than that in sham eyes at 3 days (Fig 7J), with a further increase observed at 7 days (Fig 7K), albeit less prominent compared to the 3-day time point (Fig 7J). The results indicated a potential reduction in immune-mediated inflammation over time.

RNA-seq comparison between UPOAO and extravascular occlusion models

The transcriptome of two extravascular occlusion models HIOP and UCCAO were analyzed for a further study of the characteristics of UPOAO. The HIOP model induces ischemia through anterior chamber perfusion with normal saline (Fig 8A), while the UCCAO model involves ligation of the unilateral common carotid artery (Fig 8H). RNA-seq analysis of retinas exposed to 60 minutes of ischemia followed by 7-day reperfusion in the HIOP model revealed enrichment in immune-related pathways, specifically highlighting increased activity in leukocyte-mediated immunity and regulation of immune effector processes (Fig 8B). This finding was consistent with GO analysis, which showed associations with toll-like receptor binding, T cell receptor binding processes, and immune-related functions in the overlapping DEGs between HIOP and UPOAO models (Fig 8C, D). To delve deeper, we explored the remaining DEGs in the HIOP model in addition to the overlapping genes. GO analysis revealed that the remaining DEGs in the HIOP model were also enriched in immune responses, such as the T cell receptor complex pathway, adaptive immune response, and B cell-mediated immunity function (Fig 8E). Remarkably, upon examining the remaining DEGs specific to the UPOAO model, GO analysis showed DEGs distinctly related to lipid and steroid metabolic processes (Fig 8F, G). This observation indicates that the UPOAO model, closely related to RAO, not only exhibits conventional immune responses but also likely involves unique regulation of lipid and steroid metabolism.

Transcriptomic results comparison between UPOAO model, HIOP model and UCCAO model.

(A) Schematic illustration of HIOP model. RNA-seq for HIOP retina samples at 7 days. (B) GO analysis of differential genes in the HIOP model. (C) Venn diagram indicating the overlapping DEGs (146 genes) between HIOP and UPOAO and their respective remaining DEGs (both 593 genes in each group). (D) GO analysis of the overlapping DEGs between HIOP and UPOAO. (E) GO analysis of remaining DEGs in the HIOP_7d model except for overlapping DEGs. (F) GO analysis of remaining DEGs in the UPOAO_7d model. (G) Heatmap showing the inter-sample distribution of lipid and steroid-related DEGs from the analysis of remaining DEGs in the UPOAO model. (H) Schematic illustration of UCCAO model. (I) GO analysis of DEGs in the UCCAO model. (J) Hub genes in PPI network analysis of the DEGs in the UCCAO model were shown. (K) Venn diagram indicating the overlapping DEGs between UCCAO and UPOAO (15 genes). (L) Venn diagram indicated there were two overlapping DEGs between UCCAO DEGs and m6A related genes.

Similarly, the UCCAO model exhibited associations with negative regulation of B cell proliferation, lymphocyte activation, and related processes (Fig 8I). Additionally, hub genes identified through protein-protein interaction (PPI) analysis were presented (Fig 8J). However, the UCCAO model displayed minimal overlap with the UPOAO model, with only 15 overlapping DEGs identified in the Venn diagram analysis (Fig 8K). Further analysis of the UCCAO DEGs highlighted two genes (Mettl14 and YTHDF2) related to m6A modifications (Fig 8L). Although both UPOAO and UCCAO models occludes arterial blood flow, our results revealed that the pathophysiological processes of two models were particularly different in 7-day reperfusion.

Discussion

This study aimed to develop a more appropriate mouse model for investigating the ischemia-reperfusion process in RAO and unravelling the underlying pathophysiological mechanisms. We combined silicone wire embolization with carotid artery ligation and effectively blocked the blood supply to the retina artery, which closely mimics the characteristics of the acute interruption of blood supply in RAO patients [26]. In the UPOAO model, a 60-minute period of ischemia time could simulate the injury of the major retinal neural cells like RGCs, BCs, HCs, cholinergic amacrine cells, and the resultant visual impairment. Moreover, histologic examination demonstrated thinning of the inner layer of the retina, especially the GCL layer. Time course transcriptome analysis revealed various pathophysiological processes related to immune cell migration, oxidative stress, and immune inflammation during non-reperfusion and reperfusion periods. The resident microglia within the retina and peripheral leukocytes which access to the retina were pronounced increased on reperfusion periods. Additionally, we compared the transcriptomic signatures of the UPOAO model with two commonly used ischemia-reperfusion mouse models (HIOP and UCCAO), shedding light on the potential relevance of the UPOAO model to ocular vascular occlusive diseases.

The UPOAO mouse model can effectively simulate the characteristic features of RAO as observed in patients. RAO patients often exhibit distinctive patterns in dark ERGs, where b-waves are reduced while a-waves remain stable [27-29]. In our study, we observed the same phenomenon in ERG amplitude after 60-minute ischemia in the UPOAO model. This was further confirmed by immunofluorescence staining of bipolar cells and photoreceptor cells. Importantly, a previous study has linked the thinning of the inner retina with alterations of ERG in patients with RAO [30]. Our result showed that ischemic damage predominantly affected the inner layer of the retina, while the outer layer was almost unaffected. This distinction is attributed to the inner layer’s reliance on the retinal artery for blood supply while the outer layer is supplied by the choroid. Additionally, histological examination revealed increased thickness of retinal NFL and GCL layers, as well as decreased thickness of the IPL at 3-day reperfusion, indicating that ischemia-induced edema has not completely subsided.

This study comprehensively explored the transcriptomic signature after ischemia and revealed resident and peripheral Immune cells may play a major role in pathological processes. We uncovered that in the non-reperfusion group, the transcriptional changes are primarily involved in immune cells such as leukocytes, neutrophils migration from the blood vessels into the ischemic retinal tissue. This process is closely linked to the injury during ischemia and reperfusion, where leukocytes and neutrophils infiltrate into the neural tissue through the vascular endothelium [31, 32]. Our results showed increased CD45+ leukocytes in retina and retinal microvascular, suggesting that the leukocytes were recruited form the bloodstream to the damaged retinal tissue following reperfusion period. During the process of leukocytes access to the retina, chemokines, chemokine receptors, adhesion molecules and components of the cytoskeleton play an important role in regulating endothelial permeability and facilitating the adhesion of leukocytes to endothelial cells [33, 34], which is consistent with our findings of related genes.

Moving on to the 3-day reperfusion period, we observed an increased number of DEGs significantly involved in the critical pathological response of oxidative stress in the context of IR tissues. The generation of reactive oxygen species (ROS) in mitochondria and subsequent oxidative stress are widely recognized as major causes of retinal cell damage induced by IR [35, 36]. This suggests that mitochondrial function, which is involved in oxidative stress processes, may play an important role in the pathophysiology of IRI during the early reperfusion stage. Oxidative stress can influence the permeability of the inner blood-retina barrier (iBRB), allowing leukocytes access to retinal tissues[37]. The infiltration and activation of immune cells are recognized as the underlying cause of many retinal diseases including ischemic ophthalmia [37, 38]. In UPOAO model, the increase of CD45+ leukocytes and Iba1+ microglia in retina after 3-day reperfusion periods confirmed this viewpoint.

In the later stage of reperfusion (7 days), the DEGs are mainly enriched in immune regulation and inflammation. With more upregulated DEGs than downregulated ones, we believe that the activation of immune inflammatory response contributes to further IRI in retinal tissue. Previous studies found that the retina can elicit immunological responses during ischemia-reperfusion injury and immune inflammation is an important phenomenon in the progression of this injury [25]. The excessive ROS generated by mitochondria in RGCs during the activation of inflammatory responses can damage cell structure and visual function [39]. Microglia, which are considered as primary resident immune cells, contribute to the inflammatory responses and consequent neural damage [40]. Infiltrating leukocytes can activate resident microglia, forming a feedback loop that exacerbates inflammation [41]. The complex interactions between immune cells and retina neurons after RIRI have been reported [42]. Interestingly, in our result of UPOAO model, the death of major retinal neural cells like RGCs, BCs, HCs correlate with the phase of increased infiltrating leukocytes and resident microglia. Therefore, we considered that the immune response and neuroinflammatory observed after 7 days of reperfusion may be the cumulative effect of the acute oxidative stress and resident and peripheral immune cells responses (Fig 9). In a word, we described the pathological processes at different time points and highlight the important role of resident and peripheral Immune cells responses in the UPOAO model. This offers valuable insights for subsequent screening of pathogenic genes and potential immunotherapy approaches.

The characteristic features of UPOAO model in ischemia-reperfusion periods.

Compared with HIOP and UCCAO models, the UPOAO model stands out as a more appropriate choice for studying retinal IRI in RAO. The HIOP model, widely used for investigating the pathogenesis of APACG, exhibits retinal degeneration features similar to the UPOAO model during two reperfusion periods [43]. In our UPOAO model, histologic and immunohistochemical analysis demonstrates that 60 minutes of ischemia followed by 3 days or 7 days reperfusion can cause irreversible damage to the retinal structure and visual function specifically, we observed distinct alterations such as the thinning of the inner retina, substantial apoptosis of RGCs, decreased b-wave. Multiple studies on the HIOP model reported consistent results [40, 44, 45]. However, some typical phenotypes in HIOP model such as thinning of the outer nuclear layer and the entire neuroretina and decreased a-wave were not observed in UPOAO model. moreover, our transcriptome sequence revealed a specific subset of DEGs unique to the UPOAO model, including Apoe, Abca1, Ldlr, Cyp39a1, Bmp6 etc. Notably, these DEGs are significantly enriched in pathways associated with lipid and steroid synthesis. Steroids and lipid metabolism homeostasis is disrupted under conditions of oxidative stress and inflammation [46]. In the UPOAO model, these lipid and steroid biological processes may result from the cumulative pathological responses triggered by IRI, setting it apart from the HIOP model. In the UCCAO model, RNA-seq results showed that DEGs were mainly enriched in two pathways: 1) epigenetic modification-related pathways, including nucleobase-containing compound catabolic process, RNA catabolic process, mRNA catabolic process and N6-methyladenosine (m6A) modification; 2) cell death pathways, including regulation of autophagosome assembly, negative regulation of neuron death, negative regulation of neuron apoptotic process. Epigenetic mechanisms may play a key role in the pathophysiology of ocular disease [47]. m6A, one of the most common RNA modifications, has been reported to regulate the cell death processes including apoptosis and autophagy in the pathological process of IRI [48]. The result suggests that epigenetic mechanisms may significantly influence cell death during UCCAO reperfusion for 7 days. Lee et al. evaluated the characteristics of UCCAO without reperfusion using visual, histologic and immunohistochemical approaches and their results showed delayed perfusion of the ipsilateral retina, thinning of the inner retinal layer 10 weeks after surgery, and dramatic decrease in the amplitudes of b-waves on day 14 after UCCAO [15, 49]. This suggested that UCCAO primarily represents a model of retinal hypoperfusion injury and may not effectively reflect acute ischemic-induced structural and functional damage in RAO (Fig 9). The results have led us to consider UPOAO model as an effective experimental model to study pathological processes underlying acute ischemia and IRI.

Our research has certain limitations that should be acknowledged. Firstly, our study focused on the changes occurred within 60 minutes of ischemia and the first 7 days of reperfusion in the UPOAO model. Further exploration is needed to understand changes brought by longer reperfusion time. Additionally, we proposed the possible pathological mechanisms, including iBRB damage, oxidative stress and immune inflammation mainly based on the enrichment results of DEGs at three reperfusion time points. More molecular experimental validation is needed.

In summary, by thoroughly exploring the injury to major retinal neural cells, visual impairment and pathophysiological changes in the UPOAO model, we confirmed that the this model can effectively simulate the acute ischemia-reperfusion processes in RAO and serve as an ideal mouse model for investigating the underlying ischemia and reperfusion pathological processes. Furthermore, our UPOAO model holds great promise as a novel model for studies of pathogenic genes and potential therapeutic interventions for RAO.

eLife Assessment

The manuscript establishes a sophisticated mouse model for acute retinal artery occlusion (RAO) by combining unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) with a silicone wire embolus and carotid artery ligation, generating ischemia-reperfusion injury upon removal of the embolus. This clinically relevant model is useful for studying the cellular and molecular mechanisms of RAO. The data overall are solid, presenting a novel tool for screening pathogenic genes and promoting further therapeutic research in RAO.

Reviewer #1 (Public Review):

Summary:

Wang, Y. et al. used a silicone wire embolus to definitively and acutely clot the pterygopalatine ophthalmic artery in addition to carotid artery ligation to completely block the blood supply to the mouse inner retina, which mimics clinical acute retinal artery occlusion. A detailed characterization of this mouse model determined the time course of inner retina degeneration and associated functional deficits, which closely mimic human patients. Whole retina transcriptome profiling and comparison revealed distinct features associated with ischemia, reperfusion, and different model mechanisms. Interestingly and importantly, this team found a sequential event including reperfusion-induced leukocyte infiltration from blood vessels, residual microglial activation, and neuroinflammation that may lead to neuronal cell death.

Strengths:

Clear demonstration of the surgery procedure with informative illustrations, images, and superb surgical videos.

Two-time points of ischemia and reperfusion were studied with convincing histological and in vivo data to demonstrate the time course of various changes in retinal neuronal cell survivals, ERG functions, and inner/outer retina thickness.

The transcriptome comparison among different retinal artery occlusion models provides informative evidence to differentiate these models.

The potential applications of the in vivo retinal ischemia-reperfusion model and relevant readouts demonstrated by this study will certainly inspire further investigation of the dynamic morphological and functional changes of retinal neurons and glial cell responses during disease progression and before and after treatments.

Weaknesses:

It would be beneficial to the manuscript and the readers if the authors could improve the English of this manuscript by correcting obvious grammar errors, eliminating many of the acronyms that are not commonly used by the field, and providing a reason why this complicated but clever surgery procedure was designed and a summary table with the time course of all the morphological, functional, cellular, and transcriptome changes associated with this model.

Reviewer #2 (Public Review):

Summary:

The authors of this manuscript aim to develop a novel animal model to accurately simulate the retinal ischemic process in retinal artery occlusion (RAO). A unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model was established using silicone wire embolization combined with carotid artery ligation. This manuscript provided data to show the changes in major classes of retinal neural cells and visual dysfunction following various durations of ischemia (30 minutes and 60 minutes) and reperfusion (3 days and 7 days) after UPOAO. Additionally, transcriptomics was utilized to investigate the transcriptional changes and elucidate changes in the pathophysiological process in the UPOAO model post-ischemia and reperfusion. Furthermore, the authors compared transcriptomic differences between the UPOAO model and other retinal ischemic-reperfusion models, including HIOP and UCCAO, and revealed unique pathological processes.

Strengths:

The UPOAO model represents a novel approach to studying retinal artery occlusion. The study is very comprehensive.

Weaknesses:

Some statements are incorrect and confusing. It would be helpful to review and clarify these to ensure accuracy and improve readability.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment:

The manuscript establishes a sophisticated mouse model for acute retinal artery occlusion (RAO) by combining unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) with a silicone wire embolus and carotid artery ligation, generating ischemia-reperfusion injury upon removal of the embolus. This clinically relevant model is useful for studying the cellular and molecular mechanisms of RAO. The data overall are solid, presenting a novel tool for screening pathogenic genes and promoting further therapeutic research in RAO.

Thank you for recognizing the sophistication and clinical relevance of our mouse model for acute retinal artery occlusion. We are grateful for your supportive feedback.

Public reviews:

(1) Response to Reviewer #1:

Summary:

Wang, Y. et al. used a silicone wire embolus to definitively and acutely clot the pterygopalatine ophthalmic artery in addition to carotid artery ligation to completely block the blood supply to the mouse inner retina, which mimics clinical acute retinal artery occlusion. A detailed characterization of this mouse model determined the time course of inner retina degeneration and associated functional deficits, which closely mimic human patients. Whole retina transcriptome profiling and comparison revealed distinct features associated with ischemia, reperfusion, and different model mechanisms. Interestingly and importantly, this team found a sequential event including reperfusion-induced leukocyte infiltration from blood vessels, residual microglial activation, and neuroinflammation that may lead to neuronal cell death.

Strengths:

Clear demonstration of the surgery procedure with informative illustrations, images, and superb surgical videos.

Two-time points of ischemia and reperfusion were studied with convincing histological and in vivo data to demonstrate the time course of various changes in retinal neuronal cell survivals, ERG functions, and inner/outer retina thickness.

The transcriptome comparison among different retinal artery occlusion models provides informative evidence to differentiate these models.

The potential applications of the in vivo retinal ischemia-reperfusion model and relevant readouts demonstrated by this study will certainly inspire further investigation of the dynamic morphological and functional changes of retinal neurons and glial cell responses during disease progression and before and after treatments.

We sincerely appreciate your detailed and positive feedback. These evaluations are invaluable in highlighting the significance and impact of our work. Thank you for your thoughtful and supportive review.

Weaknesses:

It would be beneficial to the manuscript and the readers if the authors could improve the English of this manuscript by correcting obvious grammar errors, eliminating many of the acronyms that are not commonly used by the field, and providing a reason why this complicated but clever surgery procedure was designed and a summary table with the time course of all the morphological, functional, cellular, and transcriptome changes associated with this model.

Thank you for your thorough review of the manuscript. We sincerely apologize for any grammatical errors resulting from our English language proficiency and have taken the necessary steps to polish the article. Additionally, we have heeded your advice and reduced the use of field-specific acronyms to enhance readability for both the manuscript and its readers.

Regarding the rationale behind the design of the UPOAO model, we have provided a description in Introduction section. Our group focuses on the research of pathogenesis and clinical treatment for RAO. The absence of an accurate mouse model simulating the retinal ischemic process has hampered progress in developing neuroprotective agents for RAO. To better simulate the retinal ischemic process and possible ischemia-reperfusion injury following RAO, we developed a novel vascular-associated mouse model called the unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) model. We drew inspiration from the widely employed middle cerebral artery occlusion (MCAO) model, commonly used in cerebral ischemic injury research, which guided the development of the UPOAO model.

We appreciate your valuable suggestion regarding the inclusion of a summary table outlining the time course of morphological, functional, cellular, and transcriptome changes associated with this model. To address this, we intend to include a supplementary table at the end of the article (Table. S2 Summary Table), which will offer a comprehensive overview of the experimental results, thereby aiding in clarity and interpretation.

Once again, we thank you for your insightful comments and suggestions, which have greatly contributed to the improvement of our manuscript.

(2) Response to Reviewer #2:

Summary:

The authors of this manuscript aim to develop a novel animal model to accurately simulate the retinal ischemic process in retinal artery occlusion (RAO). A unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model was established using silicone wire embolization combined with carotid artery ligation. This manuscript provided data to show the changes in major classes of retinal neural cells and visual dysfunction following various durations of ischemia (30 minutes and 60 minutes) and reperfusion (3 days and 7 days) after UPOAO. Additionally, transcriptomics was utilized to investigate the transcriptional changes and elucidate changes in the pathophysiological process in the UPOAO model post-ischemia and reperfusion. Furthermore, the authors compared transcriptomic differences between the UPOAO model and other retinal ischemic-reperfusion models, including HIOP and UCCAO, and revealed unique pathological processes.

Strengths:

The UPOAO model represents a novel approach to studying retinal artery occlusion. The study is very comprehensive.

We greatly appreciate your positive assessment of our work and are encouraged by your recognition of its significance.

Weaknesses:

Some statements are incorrect and confusing. It would be helpful to review and clarify these to ensure accuracy and improve readability.

We sincerely appreciate your meticulous review of the manuscript. Taking into account your valuable feedback, we will thoroughly address the inaccuracies identified in the revised version. Additionally, we will commit to polishing the article to ensure improved readability. We apologize for any confusion caused by these inaccuracies and genuinely thank you for bringing them to our attention.

Recommendations For The Authors:

Reviewer #1:

(1) Response to comment:

The conclusions of this paper are mostly well supported by clear images and convincing data analysis, but some aspects of image presentation and additional data analysis may be needed to strengthen the manuscript.

We sincerely appreciate your positive assessment of our work and your recognition of the clear images and convincing data analysis supporting our conclusions. Your constructive feedback on enhancing the clarity of our manuscript's image presentation and additional data analysis is highly valued. In response to your suggestions, we have taken steps to improve readability by removing or correcting uncommon acronyms from certain images. We have also conducted further data analysis to provide more comprehensive insights. Thank you for your guidance in improving the quality of our manuscript.

(2) Response to recommendation (1):

In Results 3.1 or in Method 2.2: please explain why this combination of silicone wire embolization and carotid artery ligation was chosen to replace previous models such as UCCAO? What are the advantages? And why the silicone wire embolus was inserted through ECA instead of inserting into CCA directly? The cleverly designed surgical procedure is very impressive but the reasoning behind it is not obvious and needs more explanation.

Thank you for your valuable feedback.

In the introduction, we briefly describe the rationale for developing the UPOAO model to simulate acute ischemia-reperfusion of retinal artery occlusion (RAO). Previous common retinal ischemia model had certain shortcomings. For example, in the HIOP model, which is often used for simulating glaucoma, the ischemic factor of interrupted retinal blood flow may be amplified due to the dual effects of IOP-induced mechanical stress [1, 2] and vascular ischemia due to normal saline perfusion in the anterior chamber. In the UCCAO model, recanalization is performed after ligation of the carotid blood vessels, and the retina communicates with the blood vessels in the brain, resulting in retinal hypoperfusion. The retina ischemia in UCCAO is a chronical process, for example, the retina became thinner at week 10 and week 15 [3], while RAO is an acute total retinal ischemic disease. Therefore, it is critically important to develop a simple mouse model that can simulate acute retinal ischemia and reperfusion injury in RAO patients.

Various models have been developed for ischemic stroke research, with the endoluminal suture model being the most employed method for middle cerebral artery occlusion (MCAO). In this model, filaments are introduced through either the external or internal carotid artery and advanced into the middle cerebral artery, causing temporary blood flow blockage for a specific duration. This method has been extensively employed in studies involving transient occlusion [4]. Among the MCAO models, the Koizumi method (occlusion from the common carotid artery (CCA) to the middle cerebral artery (MCA)) and the Longa method (occlusion from the external carotid artery (ECA) to the MCA) are frequently used. Among these two methods, the Longa method is more widely utilized in research studies. The Longa method has a much lower mortality rate post-surgery (26%) than that of the Koizumi (44%) [5]. The MCAO model induces substantial infarct areas and significantly contributes to advancements in stroke research, including investigations into blood-brain barrier disruption and inflammatory responses to ischemia.

RAO is considered a form of ocular stroke. Inspired by the MCAO model, we have employed a silicone wire embolus to induce acute interruption of blood flow to the retina. This approach enables the investigation of pathophysiological processes associated with RAO, providing valuable insights into the understanding of this condition. We have clarified these points in the revised manuscript (line 129).

The reasoning behind inserting the silicone wire embolus through the ECA instead of directly into the CCA is twofold:

(1) Convenience and avoidance of heavy bleeding and mortality. Inserting the silicone wire embolus requires creating an opening in the artery, which then needs to be ligated at both ends after the silicone wire embolus is removed to prevent excessive bleeding. The ECA's ability to form a straight line with the ICA after folding makes it more convenient for the entry and removal of the silicone wire embolus. This procedure is more convenient to perform on the ECA. The blood flow to the CCA can be restored after the plug is removed from ECA, ensuring that the blood supply to the brain through the CCA is not affected.

(2) Preservation of reperfusion process. If the silicone wire embolus were inserted directly into the CCA, the ends of the CCA opening would need to be ligated after the silicone wire embolus is removed. This would result in a lack of reperfusion process after retinal ischemia. To enable the reperfusion process, the decision was made to open the ECA instead.

We have clarified these points in the revised manuscript to better explain the rationale behind our methodology (line 139). Thank you for prompting this important clarification, which we believe will enhance the understanding of our readers.

(3) Response to recommendation (2):

Did the UPOPA actually block OA, including both the retinal (CRA) and choroidal (SPCA and LPCA) blood supply? If so, why does it seem only the inner retina was affected but not the outer retina?

Thank you for your question. We agree with you that the UPOAO model blocks OA, which includes retinal and choroidal vessels. Our experimental results primarily indicate damage to the inner retinal layer within 7 days of reperfusion. For example, OCT and HE staining showed significant thinning of the inner retina after 60 minutes of ischemia followed by 7 days of reperfusion (Figure 4). At the same time, the b-wave amplitudes were decreases, usually indicating damage to the inner layer of the retina. However, the outer retina was seemed not affected by 60 minutes of ischemia based on the results of OCT, HE and immunofluorescence.

Inner layer of the retina was known to show the highest sensitivity to hypoxic challenges [6], whereas the outer retinal layer was more resistant to hypoxic stress [7]. The possible reason for these results was that the outer layer like photoreceptors is more tolerant against ischemia than inner layer of the retina. Previous studies of retinal ischemia-reperfusion models supported this assumption. In the UCCAO model, the b-wave was more affected than the a-wave. Decreases in the amplitudes of OPs, scotopic b-wave, and photopic b-wave were consistently observed on week 4 after UCCAO, while the amplitude of scotopic a-wave did not dramatically change [8]. Prolonged ischemia, such as permanent ischemia, led to photoreceptor cell degradation, as seen in Stevens et al.'s report of photoreceptors loss 3 months after permanent ligation of both common carotid arteries in bilateral common carotid artery occlusion (BCCAO) [9]. In the HIOP model, the GCL and INL reacted sensitively to ischemic processes. A significant thinning of the GCL as early as 6 hours after 60 minutes of ischemia [10]. Horizontal cells and photoreceptors remained mostly unaffected, while most RGCs and several amacrine cell subtypes disappear [11, 12].

Our study revealed the changes that occurred within 60 minutes of ischemia and the first 7 days of reperfusion in the UPOAO model. One possibility was that the ischemia duration in our model was not long enough to affect the outer retinal cells. Furthermore, the observation time point for reperfusion was not long enough to see the structure damage and visual dysfunctions in the outer retinal layer. As we have explained in the manuscript, further exploration is needed to understand changes induced by longer ischemia duration and reperfusion periods. Revealing the damage to retinal structure and function during longer ischemia time will be an emphasis direction for our further research.

(4) Response to recommendation (3):

Better to only use well-accepted acronyms and remove those that are rarely seen in other publications, such as IMRL, MRL, HIOP, TRT, etc.

Thank you for your valuable feedback. In our manuscript, we utilized the Spectralis HRA+OCT device (Heidelberg) to capture the retinal images. However, the resulting image layering did not adequately distinguish each retinal layer clearly. To address this limitation, we referred to a clinical OCT stratification approach in RVO and divided the retina into the inner, middle, and outer layers [16]. We acknowledge that this hierarchical description is not commonly used and have therefore followed your recommendation to remove these rare acronyms and instead employ the layer structure abbreviation along with the plus sign. The methods and results have been revised accordingly (line 213, line 368, Figure 4 and Figure S2).

In addition, for the HIOP model, it is also known as the IR or RIRI model [17-19], and the pathophysiological process of retinal ischemia-reperfusion injury (IRI) is usually used to represent this type of anterior chamber perfusion model. To avoid confusion between the pathophysiological process of ischemia-reperfusion studied in this paper and the common model of high intraocular pressure, we have consistently referred to it as the HIOP model, an abbreviation that is cited in many references [20-22].

Thanks again for the suggestion. We apologize for any confusion caused by the use of abbreviations and have made the necessary corrections in the manuscript. We have also strengthened the details of OCT layering in the images to enhance readability for our audience.

(5) Response to recommendation (4):

Figure 3F, G: What do the OP changes mean? What retina cell dysfunction leads to OP changes? Is there RGC-relevant visual function readout to correlate with RGC death?

Oscillatory potentials (OPs) are important components of the electroretinogram (ERG). While the precise origin of OPs remains unclear, they are generally believed to be generated from the inner retinal layer, specifically involving bipolar cells, amacrine cells and ganglion cells [23]. OPs are sensitive indicators of retinal ischemic effects and can detect dysfunction before alterations in the b-waves occur [24-26] (We have added these statements at line 358). In this research, the reduction of OPs indicated dysfunction in the inner retinal layer and retinal ischemia.

The function of RGCs can be non-invasively assessed by using various ERG technique that emphasize the activity of inner retina neurons, including OPs of multifocal ERG (mfERG), photopic negative response (PhNR) in mfERG, pattern electroretinogram (PERG), negative Scotopic Threshold Response (nSTR) [27]. Among these indicators, the PERG appears to be more specifically related to the presence of functional RGCs. However, the complexity of electrophysiological sources and species-specific differences in RGCs characteristics should also be considered. In addition, visual evoked potentials (VEP) can assess the function of visual signaling in the whole visual pathway from RGC axons to the visual cortex of the brain [28, 29]. Unfortunately, due to the unavailability of specific equipment required for evaluating RGCs function, we encountered limitations in conducting a comprehensive assessment in this study. This limitation emphasizes the importance of future studies incorporating RGCs evaluation to provide a more comprehensive understanding of visual pathway functionality and its implications, considering indicators such as PERG and PhNR.

Thank you for your careful review and insightful questions.

(6) Response to recommendation (5):

Figure 4B: RNFL/GCL/IPL normally called GCC (ganglion cell complex).

We appreciate your helpful recommendation regarding the abbreviation GCC (ganglion cell complex) for the combination of RNFL, GCL, and IPL. We have updated this terminology in the revised manuscript (line 213 and Figure 4).

(7) Response to recommendation (6):

Figure 4 A-F: Normally a circular OCT image surrounding the optic nerve head is preferred to measure retina thickness. If in these figures, all the OCT images are from the same location, it may be acceptable, but need to provide imaging details on how these OCT planes are selected and what has been done to make sure the same locations were selected for comparison.

We agree with your comment on OCT imaging that the retina is usually captured OCT images surrounding the optic nerve head. In this study, our goal was to assess both the thickness of the peripheral retina and the retina near the optic nerve head. To achieve this, we considered the optic nerve head as the apex of the selected field of view (left upper region of panel A in Figure 4). For each mouse, we obtained OCT images of the superior nasal (SN), superior temporal (ST), inferior nasal (IN), and inferior temporal (IT) fields of the optic nerve. We then averaged the thicknesses from these four fields. In each field, we measured and statistically evaluated the retinal thickness at distances of 1.5, 3, and 4.5 papillae diameters (PD) from the optic nerve head.

This approach allowed us to ensure that the same locations were selected for comparison and provided a comprehensive assessment of retinal thickness across different regions. We have detailed this methodology in the revised manuscript to clarify the imaging process and the consistency of the selected locations.

Thank you for your insightful feedback.

Reviewer #2:

Addressing the following concerns is necessary to improve the manuscript.

(1) Response to recommendation (1):

The manuscript contains many grammatical errors and should be carefully reviewed for corrections. For example: In the title, "Silicone Wire Embolization-induced Acute Retinal Artery Ischemia and Reperfusion Model in Mouse: Gene Expression Provide Insight into Pathological Processes". It should be "Provides" instead of "Provide". In the Abstract, "The resident microglia within the retina and peripheral leukocytes which access to the retina were pronounced increased on reperfusion periods." It should be "pronouncedly" or "markedly" instead of " pronounced".

Thank you for your careful reading and pointing out the grammatical errors in the manuscript. We apologize for these mistakes and have since revised and polished the article with the assistance of native English speakers. Ensuring accurate and clear language usage in scientific writing is crucial, and we appreciate your help in improving the quality of our manuscript. Thank you for bringing these errors to our attention.

(2) Response to recommendation (2):

Video 2: the video content from "30s-47s" and "50s-67s" is repeatedly shown.

Thank you for your careful review of the video. In the process of preparing the external carotid artery for silicone wire embolus insertion, we first ligated the distal end with a square knot and then tied a loose knot at the proximal end. In the video content from "30s-47s" and "50s-67s", we are tying a square knot. We apologize for any confusion caused by these repeated video clips.

(3) Response to recommendation (3):

Figure 1: The ConA staining (H-I) and FFA (J-K) were performed before the removal of silicone wire embolus. It would be beneficial to clarify this in the figure legend too. Additionally, the label 'Post. Sup. Alveolar art.: Posterior superior alveolar artery' is not present in Figure 1L."

Thank you for your thorough review of the manuscript and the valuable suggestions regarding Figure 1. We have updated the figure legend of Figure 1 to clarify that ConA staining (H-I) and FFA (J-K) were performed before the removal of the silicone wire embolus (line 868 and line 873). Additionally, we have included the label 'Post. Sup. Alveolar art' in Figure 1L as you pointed out. We appreciate your careful attention to detail, and we have ensured that these omissions have been rectified in the revised version of the manuscript.

(4) Response to recommendation (4):

Figure 2: only representative images of RGCs at the peripheral retina were shown. It is not clear if only RGCs in the peripheral retina were quantified. Is there RGC loss in the central and middle retina in the UPOAO model as well? How many fields of RGCs were quantified for each retina?

Thank you for your meticulous review of the manuscript. The quantification method of RGCs is described in detail as follows:

Four radial incisions were made in the retina and flattened on a glass slide to create a "four-leaf clover" shape. Retina was photographed using a fluorescence microscope (BX63, Olympus, Japan). We captured images from three different regions of each retinal quadrant: 0.1 mm-0.5 mm (central region, field numbers: 1, 4, 7, 10), 0.9 mm-1.3 mm (middle region, field numbers: 2, 5, 8, 11), and 1.7 mm-2.1 mm (peripheral region, field numbers: 3, 6, 9, 12) from the optic nerve head, respectively, as shown in Author response image 1.

Of these, the peripheral field changes were the most noticeable, so we used the Leica SP8 confocal microscope (20X) to capture peripheral field RGCs as a demonstration (Figure 2A, C, E, G). RGC counts of twelve fields of each retina were quantified and the average density of RGCs in twelve fields per retina was shown in Figure 2B, D, F, K. RGC counts in the central (field number: 1, 4, 7, 10), middle (field number: 2, 5, 8, 11), and peripheral (field number: 3, 6, 9, 12) visual fields were shown in Author response table 1-4.We have included this detailed methodology in the revised manuscript to clarify the quantification process and to address the presence of RGCs loss in both the central and middle retina in the UPOAO model. Thank you for pointing out the need for this clarification.

Schematic diagram of field selection.

Scale bar=1.4 mm. Each retinal petal has three distinct visual fields (the area circled by the green line) that radiate from the optic nerve head to the periphery, in that order, the central, middle, and peripheral visual fields.

(5) Response to recommendation (5):

Figure 3: The representative wave lines in panels A (60min_3d, 60min_7d) and F do not reflect the statistical analysis presented in panels D, E, and G, especially for the amplitudes of b waves and OPs.

Thank you for your careful review of the manuscript. We've added labels for a-waves, b-waves, and improved the presentation of OPs to make the details of the amplitude more visible (Figure 3). In the previous version, due to incorrect settings, we did not adjust the ordinate spacing when fitting curves of representative wave lines in four groups, resulting in the curves being compressed vertically to the same height. We have now adjusted the curves to be fitted under the same scale bar (shown in the bottom right corner of Figure. 3A). What’s else, we removed the baseline wave of the OPs wave and adjusted the abscissa scale to highlight the N waves and P waves for easy reading (Figure 3F).

(6) Response to recommendation (6):

There are two different Supplementary Figure 1 and no Supplementary Figure 3, resulting in misaligned references to Supplementary Figures 1, 2, and 3 in the text.

Thank you for your careful review of the manuscript. We have reviewed the manuscript again and identified errors in uploading the supplementary figures, which resulted in duplicate Supplementary Figure 1 and the absence of Supplementary Figure 3. We have corrected these issues and realigned the references to Supplementary Figures 1, 2, and 3 in the text to ensure consistency. We appreciate your attention to detail and your reminder to address this issue.

(7) Response to recommendation (7):

There is confusion about the definition of ORL (outer retina layer). In Lines 208-209, ORL was defined as the combined thickness of the rest to the retinal pigment epithelium (RPE). It seems the ONL is included in ORL. But in lines 358-359, 907-908, "the ORL encompassed the region from the inner segment/outer segment (IS/OS) to the RPE". Please make the definition consistent. In addition, it is hard to distinguish the regions marked by the green lines in Fig. 4A (sham image) after Line 902.

Thank you for your careful review of the manuscript. We have addressed the confusion regarding the definition of the outer retinal layer (ORL). The Heidelberg OCT device does not distinguish the layers of the mouse retina well, so we divided it into three broader layers:

(1) Ganglion Cell Complex (GCC) layer, which encompasses RNFL+GCL+IPL.

(2) Middle Retinal Layer, which includes INL+OPL.

(3) Outer Retinal Layer (ORL), which includes ONL+IS/OS+RPE.

We apologize for the inconsistency and have revised the errors in the manuscript and figure legends accordingly. Additionally, we have removed rare domain-specific acronyms and replaced them with more commonly understood abbreviations, as suggested, to avoid confusion.

Furthermore, we have enlarged parts of the OCT images to better display the layers, hoping to meet the readers' requirements and improve clarity. Thank you for your valuable feedback.

(8) Response to recommendation (8):

Figure 4 (Panels H-J, L-M) incorporated with the text (Line 902) differs from the high-resolution version of Figure 4 included later in the manuscript. In Figure 4 (Panels H-J, L-M) merged with the text (Line 902), the quantification of the IPL and INL thickness is incorrect, and the scale bar is inaccurate. However in the high-resolution version of Figure 4 provided later, the thickness of the RNFL+GCL is incorrect.

Thank you for your careful review of the manuscript. The quantification of the IPL and INL thickness in Figure 4 (Panels H-J, L-M) incorporated with the text has been revised to ensure accurate measurements and scale bars (Figure 4 and line 924). The high-resolution version of Figure 4 provided later has been updated to correct the thickness measurements of the RNFL+GCL. We have ensured that the ordinate in the high-resolution version of Figure 4 now correctly represents length units, consistent with the equal proportional conversion used in the integrated text figures.

Thank you for your valuable feedback and for pointing out these errors. We have made the necessary corrections to align the figures accurately with the manuscript.

(9) Response to recommendation (9):

Line 384-386: the statement "Notably, a-waves in ERG and the thickness of the outer retinal layers in both OCT and HE remained unchanged." is not accurate, since a-waves in ERG is not changed in 3 days but changed in 7 days, and the thickness of the outer retinal layers in HE is either not measured or not shown in Figure 4.

Thank you for your careful review of the manuscript. We apologize for this error and have revised it.

We aimed to convey that the amplitude of the a-waves, which represent the function of the photoreceptors, does not show significant variation, which is consistent with the thickness of the outer retinal layer observed in OCT and HE images. Our results indicated that at 7 days post-injury, the amplitude of the a-waves in ERG was statistically different only at stimulus light intensity of 0.3, 3.0 and 10.0 cd.s/m2. In contrast, the b-wave amplitude was reduced by half compared to sham eyes at almost all stimulus light intensities. At the same time, the immunofluorescence staining results of photoreceptor cells showed no significant change at 7-days. Therefore, we consider the change in a-wave amplitudes were not significant compared to the significant decrease in b-wave amplitude. We have clarified this in the revised manuscript.

We also analyzed the thickness of the outer retinal layers in HE and found it to be consistent with OCT results, showing no significant changes (shown in below Author response image 2).

Thank you for your valuable feedback, which has helped improve the accuracy and clarity of our manuscript.

Thickness of OPL, ONL, IS/OS+RPE in HE staining.

n=3; ns: no significance (p>0.05).

(10) Response to recommendation (10):

Figure 5 and Figure S3: Quantification data from different sections of the same retina should be averaged to represent one single sample (one data point) for statistical analysis. * in images of Fig. 5E, F, I, J is not defined in the figure legend. It would be easier for readers to follow if the GCL, IPL, INL, and OPL were labeled in retinal sections.

Thank you for your careful review of the manuscript and recommendation. We have reperformed the statistical analysis and updated the results in Figure 5 and Figure S3. In the UPOAO experimental eyes, no no significant change in the number of HCs (Calbindin) was observed during the 3-days reperfusion period, while a notable reduction was observed after 7 days (Figure 5). Additionally, we have added the definition of the asterisks (*) in the figure legend to clarify their significance. We have also labeled the retinal layers, including the GCL, IPL, INL, OPL, and ONL, in the images to make it easier for readers to follow and understand the data.

Thank you for helping us improve the clarity and accuracy of our manuscript.

(11) Response to recommendation (11):

Lines 407-409, the statement "which aligns with the a-waves observed in ERG (Figure 3D, E) and the changes seen in the outer retinal layers in OCT (Fig S2C, D)" is confusing. No changes were observed by OCT in Fig S2D.

Thank you for your review and we are sorry about the confusion. The overall trend of the amplitude of the a-wave in ERG at 7-days did not change significantly, which is consistent with the immunofluorescence staining results of the photoreceptor cells. Based on these observations, we consider that the change in the amplitude of the a-wave was not significant. As you pointed out in recommendation 9，since a-waves in ERG were changed in 7-days at the stimulus light intensity of 0.3, 3.0 and 10.0 cd.s/m2, our description on the a-waves in 7-days was not accurate. We have clarified this point in the revised manuscript to ensure it accurately reflects the data presented.

(12) Response to recommendation (12):

In Figure S4, panel C shows lymphocyte-mediated immunity, and panel D shows leukocyte-mediated immunity. Please adjust the figure legend accordingly to reflect the figures.

Thank you for your careful review of the manuscript. We have modified the figure legend of Figure S4.

(13) Response to recommendation (13):

Lines 440-442 state "These results suggested early ischemic processions such as cell migration and potential collateral vessel formation." It is not clear why and how "potential collateral vessel formation" is suggested by Figure 6 and Figure S4. Please clarify this in the text.

Thank you for your careful review of the manuscript and we have deleted this sentence due to insufficient evidence. We have corrected this sentence: "These results suggested that in the early stage of retinal ischemic injury, leukocytes from the microvasculature may infiltrate retinal tissue. More experimental validation will be performed to confirm this hypothesis."(line 448). We will be more cautious in drawing conclusions in the future. Thank you for your reminder.

(14) Response to recommendation (14):

For the figure legend of Figure 6 "In each heatmap, upper box showed the top 10 up-regulated genes, and the below one showed the top 10 down-regulated genes." Is this correct? It appears that the upper box shows the top 10 down-regulated genes, and the lower box shows the top 10 up-regulated genes.

Thank you for your careful review of the manuscript and we have modified the figure legend of Figure 6. In the heatmaps, the upper box showed the top 10 down-regulated genes, and the below one showed the top 10 up-regulated genes (line 977).

(15) Response to recommendation (15):

For the figure legend of Figure 7, the statement 'Data points are from retinal sections of four animals' is incorrect, as these data were obtained from whole retinas instead of retinal sections. Please revise the legend to reflect this accurately. The scale bar was absent in the images of Figure 7. Asterisk in Figure 7H and 7I was not defined.

Thank you for your careful review of the manuscript and we have revised the errors. We have added the scale bar (Figure 7D). The white asterisks in Figure 7H and 7I indicate the activated microglial cells and we have added this definition in the legend of Figure7 (line 981).

(16) Response to recommendation (16):

It would be better to switch the order of Figure S7 and Figure S8 to align with their descriptions in the text.

Thank you for your recommendation and we have switched the order of Figure S7 and Figure S8.

(17) Response to recommendation (17):

The gene names in Figure S8 should be written consistently with those listed in Table S1.

Thank you for your recommendation and we have corrected the gene names.

(18) Response to recommendation (18):

In Figure 9, it is not clear why amacrine cells were not included in the UPOAO model, as amacrine cells were also injured as shown in Figure 5I-L.

Thank you for your careful review of the manuscript and we have added amacrine cells in Figure 9.

References

(1) Yang, H., et al., The connective tissue phenotype of glaucomatous cupping in the monkey eye - Clinical and research implications. Prog Retin Eye Res, 2017. 59: p. 1-52.

(2) Pavlatos, E., et al., Regional Deformation of the Optic Nerve Head and Peripapillary Sclera During IOP Elevation. Invest Ophthalmol Vis Sci, 2018. 59(8): p. 3779-3788.

(3) Lee, D., et al., A mouse model of retinal hypoperfusion injury induced by unilateral common carotid artery occlusion. Experimental Eye Research, 2020. 201: p. 108275.

(4) Barthels, D. and H. Das, Current advances in ischemic stroke research and therapies. Biochim Biophys Acta Mol Basis Dis, 2020. 1866(4): p. 165260.

(5) Smith, H.K., et al., Critical differences between two classical surgical approaches for middle cerebral artery occlusion-induced stroke in mice. J Neurosci Methods, 2015. 249: p. 99-105.

(6) Janáky, M., et al., Hypobaric hypoxia reduces the amplitude of oscillatory potentials in the human ERG. Doc Ophthalmol, 2007. 114(1): p. 45-51.

(7) Tinjust, D., H. Kergoat, and J.V. Lovasik, Neuroretinal function during mild systemic hypoxia. Aviat Space Environ Med, 2002. 73(12): p. 1189-94.

(8) Lee, D., et al., Retinal Degeneration in a Murine Model of Retinal Ischemia by Unilateral Common Carotid Artery Occlusion. Biomed Res Int, 2021. 2021: p. 7727648.

(9) Yamamoto, H., et al., Complex neurodegeneration in retina following moderate ischemia induced by bilateral common carotid artery occlusion in Wistar rats. Exp Eye Res, 2006. 82(5): p. 767-79.

(10) Palmhof, M., et al., From Ganglion Cell to Photoreceptor Layer: Timeline of Deterioration in a Rat Ischemia/Reperfusion Model. Front Cell Neurosci, 2019. 13: p. 174.

(11) Adachi, M., et al., High intraocular pressure-induced ischemia and reperfusion injury in the optic nerve and retina in rats. Graefes Arch Clin Exp Ophthalmol, 1996. 234(7): p. 445-51.

(12) Jehle, T., et al., Quantification of ischemic damage in the rat retina: a comparative study using evoked potentials, electroretinography, and histology. Invest Ophthalmol Vis Sci, 2008. 49(3): p. 1056-64.

(13) Hayreh, S.S., H.E. Kolder, and T.A. Weingeist, Central retinal artery occlusion and retinal tolerance time. Ophthalmology, 1980. 87(1): p. 75-8.

(14) Luo, X., et al., Hypoglycemia induces general neuronal death, whereas hypoxia and glutamate transport blockade lead to selective retinal ganglion cell death in vitro. Invest Ophthalmol Vis Sci, 2001. 42(11): p. 2695-705.

(15) Schmid, H., et al., Loss of inner retinal neurons after retinal ischemia in rats. Invest Ophthalmol Vis Sci, 2014. 55(4): p. 2777-87.

(16) Furashova, O. and E. Matthè, Hyperreflectivity of Inner Retinal Layers as a Quantitative Parameter of Ischemic Damage in Acute Retinal Vein Occlusion (RVO): An Optical Coherence Tomography Study. Clin Ophthalmol, 2020. 14: p. 2453-2462.

(17) Pang, Y., et al., CD38 Deficiency Protects Mouse Retinal Ganglion Cells Through Activating the NAD+/Sirt1 Pathway in Ischemia-Reperfusion and Optic Nerve Crush Models. Invest Ophthalmol Vis Sci, 2024. 65(5): p. 36.

(18) Feng, Y., et al., GSK840 Alleviates Retinal Neuronal Injury by Inhibiting RIPK3/MLKL-Mediated RGC Necroptosis After Ischemia/Reperfusion. Invest Ophthalmol Vis Sci, 2023. 64(14): p. 42.

(19) Zeng, S., et al., CREG Protects Retinal Ganglion Cells loss and Retinal Function Impairment Against ischemia-reperfusion Injury in mice via Akt Signaling Pathway. Mol Neurobiol, 2023. 60(10): p. 6018-6028.

(20) Rosenbaum, D.M., et al., The role of the p53 protein in the selective vulnerability of the inner retina to transient ischemia. Invest Ophthalmol Vis Sci, 1998. 39(11): p. 2132-9.

(21) Zhang, Y., et al., Melatonin Alleviates Pyroptosis of Retinal Neurons Following Acute Intraocular Hypertension. CNS Neurol Disord Drug Targets, 2021. 20(3): p. 285-297.

(22) Zhu, J., et al., Protective effects of Erigeron breviscapus Hand.- Mazz. (EBHM) extract in retinal neurodegeneration models. Mol Vis, 2018. 24: p. 315-325.

(23) Wachtmeister, L., Oscillatory potentials in the retina: what do they reveal. Prog Retin Eye Res, 1998. 17(4): p. 485-521.

(24) Cao, W., et al., Dextromethorphan attenuates the effects of ischemia on rabbit electroretinographic oscillatory potentials. Documenta Ophthalmologica, 1993. 84(3): p. 247-256.

(25) Xu, J., et al., Pregabalin Mediates Retinal Ganglion Cell Survival From Retinal Ischemia/Reperfusion Injury Via the Akt/GSK3β/β-Catenin Signaling Pathway. Invest Ophthalmol Vis Sci, 2022. 63(12): p. 7.

(26)Takács, B., et al., Electroretinographical Analysis of the Effect of BGP-15 in Eyedrops for Compensating Global Ischemia-Reperfusion in the Eyes of Sprague Dawley Rats. Biomedicines, 2024. 12(3).

(27) Porciatti, V., Electrophysiological assessment of retinal ganglion cell function. Exp Eye Res, 2015. 141: p. 164-70.

(28) Ridder, W.H. and S. Nusinowitz, The visual evoked potential in the mouse—Origins and response characteristics. Vision Research, 2006. 46(6): p. 902-913.

(29) Liu, S., et al., An optimized procedure to record visual evoked potential in mice. Exp Eye Res, 2022. 218: p. 109011.

Version 2

Abstract

Abstract

Acute retinal ischemia and ischemia-reperfusion injury are the primary causes of retinal neural cell death and vision loss in retinal artery occlusion (RAO). The absence of an accurate mouse model for simulating the retinal ischemic process has hindered progress in developing neuroprotective agents for RAO. We developed a unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model using silicone wire embolization combined with carotid artery ligation. The survival of retinal ganglion cells and visual function were evaluated to determine the duration of ischemia. Immunofluorescence staining, optical coherence tomography, and haematoxylin and eosin staining were utilized to assess changes in major neural cell classes and retinal structure degeneration at two reperfusion durations. Transcriptomics was employed to investigate alterations in the pathological process of UPOAO following ischemia and reperfusion, highlighting transcriptomic differences between UPOAO and other retinal ischemia-reperfusion models. The UPOAO model successfully replicated the acute interruption of retinal blood supply observed in RAO. 60-minutes of Ischemia led to significant loss of major retinal neural cells and visual function impairment. Notable thinning of the inner retinal layer, especially the ganglion cell layer, was evident post-UPOAO. Temporal transcriptome analysis revealed various pathophysiological processes related to immune cell migration, oxidative stress, and immune inflammation during the non-reperfusion and reperfusion periods. A pronounced increase in microglia within the retina and peripheral leukocytes accessing the retina was observed during reperfusion periods. Comparison of differentially expressed genes (DEGs) between the UPOAO and high intraocular pressure models revealed specific enrichments in lipid and steroid metabolism-related genes in the UPOAO model. The UPOAO model emerges as a novel tool for screening pathogenic genes and promoting further therapeutic research in RAO.

Introduction

Retinal artery occlusion (RAO) is a severe ophthalmic disease characterized by a sudden interruption of blood flow in the retinal artery, leading to retinal ischemia [1]. Over 60% of RAO patients suffer from impaired vision, ranging from finger counting to complete vision loss [2]. Additionally, RAO patients face increased risk of cardiovascular and cerebrovascular events [3-5]. Retinal ischemia and hypoxia can lead to irreversible damage to retinal cells within 90 minutes [6]. This damage is primarily due to the apoptosis of retinal ganglion cells (RGCs), which is driven by inflammation and oxidative stress. Unfortunately, conservative treatments for RAO, such as ocular massage and hyperbaric oxygen therapy, provide limited therapeutic benefits [7]. Although thrombolysis is effective, its application is restricted by a narrow therapeutic window (typically within 4.5 hours) [8]. Furthermore, the restoration of blood flow to the ischemic area following thrombolysis can cause a sudden increase in tissue oxidative levels, resulting in ischemia-reperfusion injury (IRI) in the retina [9]. IRI is a common pathological condition that can induce inflammation, retinal tissue damage and visual impairment [10]. Therefore, it is crucial to develop an animal model that accurately simulates the pathological processes of RAO to thoroughly investigate the pathophysiological changes and explore potential neuroprotective treatments following retinal ischemia and reperfusion.

Various animal models have been employed to investigate the effects of retinal injury resulting from ischemia and subsequent reperfusion in RAO [6, 11, 12]. Based on the methods used for modelling, retinal ischemia-reperfusion models can be categorized into two groups: 1) intravascular occlusion models, including photochemical-induced thrombosis model and vascular intervention model; and 2) extravascular occlusion models, including central retinal artery ligation (CRAL) model, unilateral common carotid artery occlusion (UCCAO) model, and high intraocular pressure (HIOP) model. The photochemical-induced thrombosis model and vascular intervention model have been reported to be useful tools for assessing retinal injury after ischemia and reperfusion. However, both models are limited by the need for trained interventional radiologists and the need for advanced techniques [13, 14]. The CRAL model, which directly clips or ligates the central retinal artery, also presents drawbacks, such as potential optic nerve damage and difficulty in application to small animals such as mice, limiting its use in experiments with larger sample sizes [6]. The UCCAO model, which induces retinal hypoperfusion by occluding the unilateral common carotid artery (CCA), is more suitable for simulating chronic ischemic retinal disease than acute retinal ischemia [15, 16]. The HIOP model is widely utilized in mice for studying ischemia-reperfusion in acute primary angle-closure glaucoma (APACG) [17]. However, the mechanical compression of the retina caused by saline injection in this model may also lead to retinal damage, potentially influencing IRI research. Therefore, it is critically important to develop a simple and low-skill-required mouse model that can simulate acute retinal ischemia and reperfusion injury in RAO patients.

To better simulate the retinal ischemic process and possible IRI following RAO, we developed a novel vascular-associated mouse model called the unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) model. In this model, we employed silicone wire embolization and carotid artery ligation to completely block the blood supply to the retina. We characterized the major classes of retinal neural cells and evaluated visual function following different durations of ischemia (30 minutes and 60 minutes) and reperfusion (3-days and 7-days) after UPOAO. Additionally, we utilized transcriptomics to investigate the transcriptional changes and elucidate the pathophysiological process of the UPOAO model after ischemia and reperfusion. Finally, we highlighted the transcriptomic differences between this model and other retinal ischemia-reperfusion models, including HIOP and UCCAO, revealing the unique pathological processes that closely resemble retinal IRI in RAO. The UPOAO model offers new insights into the mechanisms and pathways involved in ischemia and reperfusion studies of RAO, providing a foundation for studying protective strategies for ocular ischemic diseases.

Methods and materials

Animals

Eight-week-old male C57BL/6 mice were used in the experiments. Only male mice were used to exclude the potential influence of oestrogenic hormones. The mice were provided sufficient food and water and were maintained on a 12-hour dark-light cycle in a room with regulated temperature conditions. All the experimental procedures were designed and conducted according to the ethical guidelines outlined in the Association for Research in Vision and Ophthalmology (ARVO) Statement for the Use of Animals in Ophthalmic and Vision Research. The study protocol and methods received approval from the Experimental Animal Ethics Committee of Renmin Hospital of Wuhan University (approval number: WDRM-20220305A).

Preparation and surgical procedure of the UPOAO model

Eight-week-old male C57BL/6 mice weighing 20-25 grams were used in this study. An isoflurane-based anaesthesia system was used to induce and maintain general anaesthesia in the mice during surgical procedures. The mice were anaesthetized with a 1.5-2% concentration of isoflurane delivered in a mixture of nitrous oxide and oxygen via rubber tubing. Body temperature was maintained at 37 ± 0.5 °C throughout the surgery. Surgical instruments were sterilized with 75% ethanol before use to ensure sterility.

We drew inspiration from the widely employed middle cerebral artery occlusion (MCAO) model [18, 19], commonly used in cerebral ischemic injury research, which guided the development of the UPOAO model. The mouse was gently positioned in a supine posture on a heating blanket, and its neck was exposed. The preparation involved depilation of the neck area, followed by skin disinfection, and finally, a midline incision along the neck was made. After separating the neck gland using two tweezers, careful blunt dissection was employed to separate the left CCA, internal carotid artery (ICA) and external carotid artery (ECA), avoiding the compression of nearby nerves and veins (Fig. 1A) (Video 1). Then, the CCA and ICA were ligated with a 6-0 suture. A knot was secured using a 6-0 suture at the distal end of the ECA, and a slipknot was created at the proximal end (Fig. 1B). To preserve the reperfusion process and avoid bleeding and mortality during surgery, a silicone wire embolus was inserted through the ECA instead of through the CCA. Ophthalmic scissors were used to make a small inverted "V"-shaped incision between the two suture knots on the ECA. A specialized silicone wire embolus, measuring 7 ± 0.1 mm in length and 0.21 ± 0.1 mm in diameter, was inserted through the incision, directing it into the ECA and further into the CCA (Fig. 1C). Next, the ligation on the ICA was removed, and the ECA was cut at the point where the artery had been previously incised with scissors. Subsequently, the silicone wire embolus was retracted to the bifurcation of the CCA, rotated counterclockwise, and inserted into the ICA and further into the pterygopalatine artery (PPA) (Fig. 1D1). The silicone tail of the silicone wire embolus was positioned near the bifurcation of the CCA (Fig. 1D2), effectively obstructing the ophthalmic artery (OA) (Video 2). Then, the slipknot was fastened, and the skin was sutured. The mouse was free to move around during arterial embolization.

Modeling Procedure, Validation, and Cervical Artery Anatomy

(A1-G1) Schematic illustration of unilateral pterygopalatine ophthalmic artery occlusion (UPOAO). (A2-G2) Practical operation of UPOAO. (A1, A2) Blunt separation and exposure of the left cervical arteries. (B1, B2) Arterial suture ligation. (C1, C2) Insertion of the silicone wire embolus. The artery was incised to create a hole, and the silicone wire embolus was inserted. (D1, D2) Artery disconnection and movement of the silicone wire embolus. The artery was cut along the incision, and the silicone wire embolus was retracted and reinserted. (E1, E2) Removal of the silicone wire embolus and reperfusion. (F1, F2) Suture removal. The sutures at both ends of the disconnected vessel were knotted, and the other two sutures were removed. (G1, G2) Anatomic reduction and suturing of the skin. (H, I) Canavalin A label vasculature of the UPOAO mouse retina. The silicone wire embolus was inserted into artery before perfusing rhodamine-labeled canavalin A into the heart of UPOAO mouse. The sham eye served as an unpracticed control eye, while the UPOAO lateral eye represented the experimental eye. Retinal vessels in the sham eye (H) exhibited fluorescence filling, while retinal vessels in the UPOAO lateral eye (I) remained unfilled. Scale bar = 1 mm. (J and K) Fluorescein Fundus Angiography (FFA) performed before removing the silicone wire embolus from the UPOAO mouse. The vessels in the sham lateral retina (J) were perfused, while the lateral retinal perfusion in UPOAO (K) was delayed. (L) Schematic illustration of cervical artery anatomy and ocular blood supply. Embolization of the PPA resulted in ocular ischemia. The red arrow indicates the site of the silicone wire embolus occlusion. The silicone wire embolus used a type 602156 wire, extended to 7mm with a diameter of 0.21mm. The blue arrows indicate the modelling locations of the HIOP model and the UCCAO model, respectively. CCA: common carotid artery; ICA: internal carotid artery; ECA: external carotid artery; PPA: pterygopalatine artery; MCA: middle cerebral artery; Infraorbital art.: infraorbital artery; OA: ophthalmic artery; SPCA: short posterior ciliary artery; LPCA: long posterior ciliary artery; CRA: central retinal artery.

Following a predetermined ischemia embolization time, the silicone wire embolus was carefully removed from the PPA without massive haemorrhage (Fig. 1E). The CCA suture was removed, restoring arterial reperfusion (Fig. 1F) (Video 3). Then, the skin wound was sutured, and the mice were routinely fed during reperfusion (Fig. 1G).

2.3 Staining of retinal vessels

To clearly visualize the retinal vessels, rhodamine-labelled canavalin A was used for mouse heart perfusion. Each mouse was anaesthetized with 1% sodium pentobarbital and placed in the supine position on a foam board resting on a tray. The sternum was opened to expose the heart. A needle attached to the perfusion tube was inserted into the left ventricle via the apex of the mouse heart and secured. The right atrial appendage was dissected with scissors to allow blood outflow. Saline solution was perfused through the left ventricle using a pump system to drain blood from the arteries. Subsequently, the prepared solution of rhodamine-labelled canavalin A was injected into the mice and circulated throughout the mouse vessels for 30 minutes. The mice were then sacrificed, and their eyes were removed and immersed in paraformaldehyde (PFA) for one hour in the dark. A flat-mounted retina was obtained, and retinal images were captured using a Leica SP8 confocal microscope equipped with a 10× objective.

Quantification of RGCs and microglia

The mice were euthanized by cervical dislocation, and their eyes were fixed in 4% PFA for 60 minutes at room temperature (RT). Subsequently, the cornea and lens were excised, and the intact retina was isolated for retinal flat mounting. The retina was immersed in PBS supplemented with 5% bovine serum albumin (BSA) and 0.5% Triton X-100 for overnight blocking and then incubated with Brn3a antibody (Santa Cruz Biotechnology, Dallas, Texas, USA) or Iba1 antibody (Wako, Japan). Following a two-day incubation with the primary antibody, the retina was gently washed three times with PBS. Subsequently, the retina was incubated with Alexa Fluor 594 (AntGene, Wuhan, China) in a light-protected cassette for two days. After three rinses, the retina was uniformly sectioned into a four-leaf clover morphology and flattened using a coverslip. Finally, retinal flat mounts labelled with Brn3a were imaged using a Leica SP8 confocal microscope (Leica TCS SP8, Germany) equipped with a 20 × objective, while those labelled with Iba1 were imaged utilizing a fluorescence microscope (BX63; Olympus, Tokyo, Japan). Each quadrant was consistently subdivided into central, middle and peripheral fields at regular intervals. The surviving RGCs and activated microglia were quantified and averaged using ImageJ software (National Institutes of Health, USA).

Electroretinogram (ERG)

Mice were dark-adapted overnight before ERG, and all subsequent procedures were carried out in the dark. Before ERG, the mice were anaesthetized with 1% sodium pentobarbital via intraperitoneal injection, and their pupils were dilated. A subcutaneous electrode was inserted into the posterior cervical skin, a tail electrode was affixed to the posterior end of the mouse tail, and the corneal electrode was gently placed on the central corneal surface. A RetiMINER-C visual electrophysiological system (3VMED Co., Ltd., Shanghai, China) was used for recording electrical responses. The a-waves, b-waves, and oscillatory potentials (OPs) after flash stimuli of 0.01, 0.03, 0.1, 0.3, 1.0, 3.0, and 10.0 cd.s/m2 in scotopic adaptation were recorded. The amplitudes and implicit times of a-waves, b-waves, and OPs in response to various flash stimuli were analysed.

Optical Coherence Tomography Imaging (OCT) and Fluorescein Fundus Angiography (FFA)

The mice were placed on the platform of a Spectralis HRA + OCT device (Heidelberg Engineering, Heidelberg, Germany) for OCT imaging following anaesthesia. Pupil dilation was performed, and normal saline was applied regularly to maintain corneal moisture. The focal length of the device was adjusted until the mouse retina was clearly visible. The head of the mouse was gently repositioned to capture images of the peripheral fundus. The multiline mode was used to scan each layer of the retina, and four quadrants of view centred on the nipple in the upper left, lower left, upper right, and lower right were recorded. In this study, the total retinal thickness was manually segmented into three parts and encompassed the entire thickness from the nerve fibre layer to the photoreceptor layer. The ganglion cell complex (GCC) was defined as the combined thickness of the retinal nerve fibre layer (RNFL), ganglion cell layer (GCL), and inner plexiform layer (IPL). The inner nuclear layer (INL) and the outer plexiform layer (OPL) were combined for thickness analysis due to the difficulty in distinguishing between these layers. The remaining retinal layers, including the outer nuclear layer (ONL), inner segment/outer segment (IS/OS), and retinal pigment epithelium (RPE) layers, were measured together. The thickness of the retina at 1.5 papillary diameters (PD), 3.0 PD, and 4.5 PD from the centre of the optic disc was measured using the Heidelberg measuring tool.

For FFA, the mice were anaesthetized, and their pupils were dilated. Subsequently, fundus angiography was performed immediately following an intraperitoneal injection of sodium fluorescein. Images were acquired alternately for both eyes within 5 seconds of the injection.

Hematoxylin and eosin (HE)

Mouse eyes were fixed in FAS eyeball fixation solution (Service-bio, Wuhan, China) for 48 hours, followed by dehydration and subsequent embedding in paraffin. The paraffin blocks were trimmed parallel to the optic nerve to obtain 3-4 μm thick sections where the optic nerve was located. Six paraffin sections from each eyeball were stained with HE. Images were captured using an Olympus fluorescence microscope. The retinal thickness within 500 μm of the optic disc was measured using Image-Pro Plus 6.0 software.

Staining of Retinal Sections

For immunofluorescence, mouse eyes were fixed in 4% PFA solution and subjected to gradient dehydration using 10%, 20%, and 30% sucrose solutions. The following day, the eyes were embedded in optimum cutting temperature (OCT) compound (SAKURA, USA) and frozen at -80 ° C. Several 14-μ m-thick frozen sections were obtained from each eyeball through the use of a freezing microtome (Leica, Wetzlar, Germany). The section surface was parallel to the optic nerve, and 3-6 frozen sections were mounted on a single slide. The slides were stored at -20°C until use.

All sections were blocked with 5% BSA and 0.5% Triton X-100 in PBS for 2 hours. Primary antibodies, as listed in Table 1, were diluted in PBS containing 5% BSA and 0.5% Triton X-100 and incubated at 4 ° C overnight. The following day, after three rinses, the sections were incubated with secondary antibodies at RT for 2 hours in a cassette, followed by three washes. The sections were then stained with DAPI (Service-bio, Wuhan, China) for 15 minutes. After the final three rinses, the frozen sections were sealed. A Leica SP8 instrument equipped with a 40× objective was used to photograph frozen sections near the optic nerve head. The fluorescence intensity of all slices of each eye was determined using ImageJ software.

Surgical procedure for the UCCAO model and the HIOP model

Animals for the UCCAO model were prepared and managed according to the procedure described for the UPOAO model. After blunt dissection, the left CCA was ligated with a 6-0 suture. After 60 minutes of ligation, the suture was removed, and the skin wound was subsequently closed with sutures. Mice were routinely fed during the reperfusion period, and the retinas of UCCAO mice were harvested 7-days post-surgery.

The HIOP model was generated based on previously established methods [20]. Briefly, mice were anaesthetized by intraperitoneal injection of 1% sodium pentobarbital. Pupils were dilated using tropicamide phenylephrine eye drops (0.5% tropicamide and 0.5% deoxyadrenaline hydrochloride, Santen Pharmaceutical) applied 5 minutes in advance. An insulin hypodermic needle attached to a silicone elastomer tube was inserted into the anterior chamber of the unilateral experimental eye. The IOP was elevated for 60 minutes by raising the height of the saline solution storage bag to 162 cm, after which the needle was removed to immediately restore the IOP to baseline levels. Iris whitening and loss of retinal red-light reflexes indicated retinal ischemia. After surgery, levofloxacin hydrochloride ophthalmic gel was applied to the eyes of the mice. Retinas from HIOP mice were collected 7-days post-surgery.

Transcriptome sequence and analysis

Retinal samples from UPOAO, HIOP, and UCCAO mice were extracted and promptly frozen in liquid nitrogen within 10 minutes of cervical dislocation. Total RNA extraction was performed according to the protocol outlined in the TRIzol Reagent manual (Life Technologies, CA, USA). RNA integrity and concentration were assessed using an Agilent 2,100 Bioanalyzer (Agilent Technologies, Inc., Santa Clara, CA, USA). The resulting RNA samples were then pooled and subjected to sequencing on the Illumina HiSeq3000 platform in a 150-bp paired-end read format. The raw RNA-sequencing (RNA-seq) reads were preprocessed and quantified using the featureCounts function in the SubReads package version 1.5.3, with default parameters.

RNA-seq analysis was conducted on samples from 5 mice in the non-perfusion group, 4 mice in the 3-days perfusion group, and 5 mice in the 7-days perfusion group of the UPOAO model. Additionally, retinas from both the 7-days HIOP experimental eyes and bilateral UCCAO eyes were collected (n=5) for analysis.

Data normalization and subsequent processing utilized the ‘limma’ package in R software (version 4.2.3) [21]. Volcano plots and Venn diagrams were generated for visualization to identify significantly differentially expressed genes (DEGs) and perform Gene Ontology (GO) annotation analysis. DEGs with a |log2-fold change (FC)| > 1 and adjusted p value < 0.05 were significantly differentially expressed. Enrichment analysis focused on DEGs with an adjusted p < 0.05 and an enriched gene count > 5. To visualize the expression patterns of significant DEGs, a heatmap was generated using the ‘pheatmap’ package. Gene Set Enrichment Analysis (GSEA) was conducted using the GSEA program (version 4.3.2) [22], employing the default gene set m-subset (mh.all.v2023.1.mm.symbols.gmt) to explore significant functional and pathway differences. Enriched pathways were classified based on criteria such as adjusted p value (< 0.05), false discovery rate (FDR) q value (< 0.25), and normalized enrichment score (|NES| > 1). Furthermore, protein ‒ protein interaction (PPI) analysis was performed using the Search Tool for the Retrieval of Interacting Genes (STRING) database (https://string-db.org/) [23]. Genes with a greater degree of protein-level interactions with others were further analysed using Cytoscape software (version 3.8.2) to generate a downstream PPI map. PPI pairings with an interaction score > 0.7 were extracted and visualized using Cytoscape 3.9.0 [24].

Real-time Quantitative polymerase chain reaction (RT-qPCR)

RNA was extracted from fresh mouse retinal samples and subjected to reverse transcription using HiScript lll RT SuperMix for qPCR (Vazyme, China) following the manufacturer’s instructions. RT‒ qPCR analysis was conducted on a CFX instrument (Bio-Rad) using AceQ qPCR SYBR Green Master Mix (Vazyme, China). The PCR program comprised 40 cycles of 10 seconds at 95°C and 30 seconds at 60°C. Assays were performed in triplicate, and Ct values were normalized to beta-actin levels. Relative quantification of target gene expression was calculated using the 2-ΔΔCt method. The primer sequences are provided in Table S1.

Statistical Analysis

All the data were analysed using GraphPad Prism version 9.0 (GraphPad Software, San Diego, CA, USA). Each experimental group included a minimum of three biological replicates. T tests, including RGCs and microglia counting, OPs analysis, retinal thickness measurements via HE staining, and fluorescence intensity quantification, were used to compare the two groups. Two-way ANOVA (two-way analysis of variance) test was used for more than two different experimental groups, including ERG waves and retinal thickness measurements in OCT. All data are presented as the mean ± standard error of the mean (SEM), and the statistical graphs are shown as scatter bar graphs and line charts. P < 0.05 was considered to indicate statistical significance.

Results

Silicone wire embolus insertion interrupts retinal blood flow

To reproduce the retinal ischemic process and potential IRI in RAO, we established a mouse model of UPOAO by combining silicone wire embolization with carotid artery ligation. Model validation was conducted through a series of experiments, with detailed descriptions provided in the Materials and Methods section (Fig. 1A-G). To confirm blood flow disruption, we performed cardiac perfusion with fluorescently labelled lectin and FFA. Rhodamine-labelled canavalin A was used during in vivo perfusion with silicone wire embolus, and both eyes were evaluated. The sham eye exhibited normal blood perfusion in the retina, while the experimental eye showed no perfusion (Fig. 1H, I). FFA revealed delayed and limited perfusion in the experimental lateral retina, primarily near the optic disc (Fig. 1J, K). These findings indicate that the insertion of the silicone wire embolus effectively impaired blood flow to the retina.

60-minute ischemia in UPOAO damage retinal structure and function

To investigate the optimal ischemic duration, the mice were subjected to ischemic periods of 30 and 60 minutes, followed by reperfusion periods of 3-days and 7-days. Retinal structure and visual function were assessed through flat-mounted retina analysis and flash ERG.

In the 30-minute ischemia group, no significant RGCs death was observed after 3-days (Fig. 2A, B) or 7-days of reperfusion (Fig. 2C, D). However, after 60 minutes of ischemia followed by either 3-days or 7-days of reperfusion, reductions in RGCs density were evident (Fig. 2E-H). ERG results showed no statistical difference in amplitudes between bilateral eyes after 3- and 7-days of 30-minutes reperfusion (Fig. 3A-C). However, the b-wave amplitude notably decreased after 60 minutes of ischemia and 3-days of reperfusion (Fig. 3D). By 7-days of reperfusion, the b-wave amplitude had halved compared to that of the sham eyes (Fig. 3E). The appearance times of the a-waves and b-waves in the experimental and control eyes remained consistent across all four groups (Fig. S1). Additionally, OPs have also been used to sensitively monitor retinal ischemic effects by detecting changes before any alterations in b-waves occur [25-27]. The amplitudes of OPs in the experimental eyes, particularly in the 60-minute ischemia and 7-days reperfusion group, significantly decreased to less than 50% of those in the sham eyes (Fig. 3F, G). Based on RGCs survival and changes in ERG waves, we determined that a 60-minute ischemic duration is optimal for the UPOAO model.

Staining and Quantification of Retinal Ganglion Cells (RGCs) at Different Ischemia and Reperfusion Times.

Flat-mounted retina RGCs were labeled with Brn3a staining. (A, B) Representative pictures of the peripheral retinal visual field (A), and quantification of surviving RGCs (B) in the 30-minute ischemia and 3-days reperfusion group. n = 3. (C, D) Representative pictures of the peripheral retinal visual field (C), and quantification of surviving RGCs (D) in the 30-minute ischemia and 7-days reperfusion group. n = 3. (E, F) Representative pictures of the peripheral retinal visual field (E), and quantification of surviving RGCs (F) in the 60-minute ischemia and 3-days reperfusion group. n = 5. (G, H) Representative pictures of the peripheral retinal visual field (G), and quantification of surviving RGCs (H) in the 60-minute ischemia and 7-days reperfusion group. n = 5. The results showed a significant loss of RGCs after 60 minutes of ischemia. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t-test. Scale bar = 100 μm.

Comparison of Electroretinographic (ERG) Dark-Adapted Responses at Different Ischemia and Reperfusion Times.

Following the evaluation of surviving RGCs, visual function in sham and UPOAO experimental eyes at various ischemia and reperfusion times was assessed using ERG. (A) Representative waveforms in the four groups at the stimulus light intensities of 0.1, 1.0, and 10.0 cd.s/m2, respectively. (B) Quantification of a-wave and b-wave amplitudes in the 30-minute ischemia and 3-days reperfusion group. n = 5. (C) Quantification of a-wave and b-wave amplitudes in the 30-minute ischemia and 7-days reperfusion group. n = 5. (D) Quantification of a-wave and b-wave amplitudes in the 60-minute ischemia and 3-days reperfusion group. n = 7. (E) Quantification of a-wave and b-wave amplitudes in the 60-minute ischemia and 7-days reperfusion group. n = 8. Dark-adapted responses showed almost similar a-wave amplitudes but significantly decreased b-wave amplitudes in the 60-minute ischemia groups. The amplitudes of b-waves declined at 3-days and even more prominently at 7-days. (F, G) Representative OPs and quantification of amplitudes in the 60-minute ischemic groups. n = 7 in the 3-days reperfusion group; n = 8 in the 7-days reperfusion group. The amplitudes of OPs decreased significantly at 7-days reperfusion. The decline in b-waves and OPs along with the loss of RGCs, supports the selection of a 60-minute ischemic duration as an appropriate choice. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test for a-waves and b-waves; paired t-test for OPs.

Evaluation of retinal thickness in UPOAO through OCT and HE

To assess retinal thickness non-invasively in the UPOAO model, we conducted OCT imaging during 3-days and 7-days reperfusion periods. Following a 3-days reperfusion period, no significant changes were observed in the GCC, INL + OPL, and ONL + IS/OS + RPE layers at 1.5, 3, and 4.5 PD from the optic disc (Fig. 4B, Fig. S2A, B). However, the total retinal thickness decreased at 4.5 PD (Fig. 4C). After 7-days of reperfusion, the total retinal thickness decreased at 1.5 PD, 3 PD, and 4.5 PD, primarily due to inner retinal thinning (Fig. 4D-F, Fig. S2C, D).

Changes in Retina Morphology in the UPOAO Model.

(A) Representative OCT images of the mouse retina at 3-days. The Green lines indicate the OCT scan area starting from the optic disc. Local magnification and layering are annotated. GCC: ganglion cell complex, including RNFL, GCL and IPL layers. (B and C) Quantification of GCC and total retinal thickness at 3-days. The thickness of the GCC and the entire retina in OCT was measured and compared at distances of 1.5 PD, 3.0 PD and 4.5 PD from the optic disc, respectively. n = 5. (D) Representative OCT images of the mouse retina at 7-days. (E and F) Quantification of GCC and total retinal thickness at 7-days. n = 5. (G) Representative HE images of the mouse retina at 3-days. (H, I and J) Quantification of NFL + GCL, IPL, and INL thickness at 3-days. Retinal thickness in HE was measured near the optic nerve head and compared. n = 3. (K) Representative HE images of the mouse retina at 7-days. (L, M and N) Quantification of NFL + GCL, IPL, and INL thickness at 7-days. n = 3. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test in OCT and paired t-test in HE. PD: papillary diameters; RNFL: retinal nerve fiber layer; GCL: ganglion cell layer; IPL: inner plexiform layer; INL: inner nuclear layer; OPL: outer plexiform layer; ONL: outer nuclear layer; IS: inner segment; OS outer segment; RPE: retinal pigment epithelium. Scale bar = 50 μm.

For a detailed analysis of inner retinal thickness, we extracted UPOAO mouse eyes and stained them with HE. The thickness of the RNFL + GCL increased at 3-days of reperfusion (Fig. 4H), followed by a decrease at 7-days of reperfusion (Fig. 4L). IPL thickness decreased at both 3-days and 7-days of reperfusion (Fig. 4I, M), while INL thickness decreased only at 7-days of reperfusion (Fig. 4N). We hypothesized that the initial swelling of the GCL in response to acute ischemia and hypoxia led to early thickening, followed by gradual thinning due to tissue dysfunction. This hypothesis was supported by the observation of RGCs loss and reduced GCC thickness evaluated by OCT at 7-days (Fig. 4E).

We observed by OCT that changes in retinal thickness became noticeable at 3-days post-UPOAO and exhibited a significant decrease at 7-days. HE results further revealed that alterations in inner retinal thickness constituted a substantial portion of the total retinal thickness during the early reperfusion stage after UPOAO.

Survival of retinal neural cells in UPOAO

Given that mouse retinal function heavily relies on rod cells and that scotopic (low light) vision is primarily governed by rod photoreceptors, the observed reductions in b-wave and OPs-wave amplitudes indicate impaired synaptic transmission within the inner retina [28]. Our OCT and HE findings suggested structural injuries within the inner retina. Notably, the thickness of the outer retinal layers in OCT remained unchanged. To further explore these observations, we conducted immunofluorescence staining to investigate alterations in major retinal cell types, primarily focusing on bipolar cells (BCs), photoreceptor cells, horizontal cells (HCs), and cholinergic amacrine cells.

BCs loss and photoreceptor cells survival

PKC α serves as a marker delineating BCs, with their cell bodies primarily located in the outermost part of the INL, axonal terminals extending into the innermost part of the IPL, and dendrites confined to the OPL (Fig. 5A, B). In the experimental UPOAO eyes, BCs did not show significant changes at the 3-days reperfusion mark but exhibited dramatic alterations by the 7-days reperfusion period (Fig. 5A2, B2). In particular, the immunostaining density of BCs dendritic and axonal arbors notably decreased at the 7-days reperfusion mark. BCs account for approximately 40% of INL cells in mice, and their somata and axonal processes form a substantial part of the INL and IPL, consistent with the thinning of the INL observed in both OCT and HE at 7-days (Fig. 4E, N). Notably, the decrease in the number of PKCα+ BCs at the 7-days reperfusion point coincided with the decrease in the b-waves amplitude in ERG (Fig. 3E), indicating functional interplay between BCs and other cells.

Changes in Bipolar Cells, Horizontal Cells, and Cholinergic Amacrine Cells in UPOAO Mice.

(A) Representative images of mouse retina co-stained with DAPI and PKC α at 3-days (B) Representative images of mouse retina co-stained with DAPI and PKC α at 7-days. (C) Quantification of PKC α fluorescence density at 3-days. (D) Quantification of PKC α fluorescence density at 7-days. n = 4. (E) Representative images of mouse retina co-stained with DAPI and Calbindin at 3-days. (F) Representative images of mouse retina co-stained with DAPI and Calbindin at 7-days. Horizontal cell somata are indicated by white asterisks. (G) Quantification of Calbindin fluorescence density at 3-days. (H) Quantification of Calbindin fluorescence density at 7-days. n = 3. (I) Representative images of mouse retina co-stained with DAPI and ChAT at 3-days. (J) Representative images of mouse retina co-stained with DAPI and ChAT at 7-days. Cholinergic amacrine cell somata are indicated by white asterisks. (K) Quantification of ChAT fluorescence density at 3-days. (L) Quantification of ChAT fluorescence density at 7-days. n = 3. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001,t-test. Scale bar = 50 μm.

Recoverin marks the somata and outer segments of photoreceptor cells and is localized within the ONL (Fig. S3A, B). Interestingly, recoverin-positive photoreceptors remained relatively stable throughout the reperfusion periods, which aligns with the unchanged thickness of the outer retinal layers observed in OCT (Fig. S2C, D).

HCs and cholinergic amacrine cells loss

We evaluated HCs and cholinergic amacrine cells using calbindin and ChAT immunostaining, respectively. Calbindin immunostaining highlighted the somata of HCs within the INL, with terminal axon connections linearly positioned within the OPL (Fig. 5E, F). In the UPOAO experimental eyes, no significant change in the number of HCs was observed during the 3-days reperfusion period, while a notable reduction was observed after 7 days. By 7-days, a significant decrease in cell body numbers, a considerable reduction in axon density, and disrupted linear connections were observed (Fig. 5F2).

ChAT+ cell bodies were located in the GCL and INL, with dendrites forming two narrow stratified bands within the IPL. The immunofluorescence density of ChAT+ amacrine cells decreased notably after 3 days and even more prominently after 7 days (Fig. 5I, J). The fluorescence intensity of the two bands within the IPL markedly decreased and was nearly invisible after 7 days (Fig. 5J2).

Time course transcriptome analysis revealed features of different reperfusion periods in UPOAO

To explore the pathophysiological processes of UPOAO, we extracted retinas from both eyes for transcriptome sequencing. The samples included retinas from eyes subjected to 60 minutes of ischemia without reperfusion, 60 minutes of ischemia followed by 3-days of reperfusion, and 60 minutes of ischemia followed by 7-days of reperfusion, with the sham-treated eyes serving as controls.

In the non-perfusion group, 215 genes were upregulated and 204 genes were downregulated (Fig. 6A). GO enrichment analysis revealed that the DEGs were related to leukocyte migration, epidermis development, myeloid leukocyte migration and other cell migration pathways (Fig. 6B). The heatmap and box showed the upregulated and downregulated genes involved in immune cell migration-related pathways during the non-perfusion period in UPOAO (Fig. 6C). High-connectivity DEGs (‘hub genes’), such as Dusp1 and Fos, were also enriched in these pathways (Fig. S4A). GSEA also showed similar results (Fig. S4B-D). These results suggested that in the early stage of retinal ischemic injury, leukocytes from the microvasculature may infiltrate retinal tissue. More experimental validation will be performed to confirm this hypothesis.

Transcriptomic Features at Different Times of Reperfusion.

RNA-seq evaluation was performed at 0-day, 3-days, and 7-days reperfusion periods in UPOAO revealing enrichment in pathways related to immune cells migration, oxidative stress, and immune inflammation. (A, D, G) Volcano plots display differential expression genes (DEGs) between UPOAO and sham eyes in the no-perfusion group (A), 3-days perfusion group (D), and 7-days perfusion group (G), respectively. Red dots: significant upregulated genes, green and blue dots: significant downregulated genes, grey dots: stable expressed genes, adjusted P < 0.05. log2FC = 1. (B, E, H) Gene ontology (GO) analysis of differential genes in the non-perfusion group (B), 3-days group (E), and 7-days group (H). The DEGs between UPOAO and sham eyes were enriched in pathways associated with immune cells migration (0d), oxidative stress (3d), and immune inflammation (7d), respectively. (C, F, I) Heatmap displaying the top 100 downregulated and top 100 upregulated DEGs between UPOAO and sham at non-perfusion, 3-days, and 7-days reperfusion, respectively. The box represents the genes related to immune cells migration (C), oxidative stress (F), and immune inflammation (I). The ranking was determined by the magnitude of fold change. In each heatmap, the upper box represents the top 10 downregulated genes, while the lower box represents the top 10 upregulated genes.

In the 3-days reperfusion group, 372 genes were upregulated, and 170 genes were downregulated (Fig. 6D). GO enrichment analysis revealed that the DEGs at 3-days after reperfusion were related to energy metabolism, mitochondrial regulation, and oxidative stress pathways (Fig. 6E). The heatmap and box displayed the upregulated and downregulated genes involved in oxidative stress-related pathways at the 3-days reperfusion stage in UPOAO, including hub genes such as Mrpl18 and Mrps23 (Fig. 6F, Fig. S5A). GSEA revealed changes in pathways related to the binding membrane of organelles, the cell surface and the plasma membrane protein complex (Fig. S5B). Furthermore, 44 overlapping genes between the mitochondrial genes and DEGs at 3-days post-UPOAO were mainly related to mitochondrial transport and the mitochondrial tricarboxylic acid cycle (Fig. S5C, D). These results suggested that mitochondria and metabolism play significant roles in IRI.

In the 7-days reperfusion group, 429 genes were upregulated, and 310 genes were downregulated (Fig 6G). The DEGs at this stage were enriched in pathways related to immune effector processes, positive regulation of responses to external stimuli and inflammation (Fig. 6H). A heatmap was generated to show the upregulated and downregulated genes of immune inflammation-related pathways at 7-days after reperfusion in UPOAO, and most hub genes, such as Cd86, Cd48, Tlr4, and Tlr6 were also enriched in these pathways (Fig. 6I, Fig. S6A). GSEA also showed similar results (Fig S6B). Moreover, RT ‒ qPCR confirmed the upregulation of immune inflammation-related gene expression (Fig. S7). Significantly, 112 genes that overlapped between immune-related genes and DEGs at 7-days post-UPOAO were primarily associated with immune receptor activity and phagocytic vesicles (Fig. S6C, D). These results underscore the predominant role of immune responses during this stage.

To explore the associations between changes observed at 3-days and 7-days after reperfusion, we analyzed the DEGs and identified 17 overlapping genes (Fig. S8A), which were mainly associated with the nitric oxide biosynthetic process (Fig. S8B). Our findings suggest that the immune inflammatory response observed after 7-days of reperfusion may represent the cumulative effect of the acute oxidative stress response during the initial 3-days reperfusion period.

Leukocyte Infiltration and Microglial Activation

Time course transcriptome analysis underscored the predominant role of immune inflammatory responses during the reperfusion stage. Previous reports have indicated that the inflammatory response in retinal ischemia-reperfusion injury (RIRI) is orchestrated by peripheral immune cells and resident immune cells within the retina [29, 30]. In our study, we demonstrated the presence of leukocyte infiltration and microglial activation in the UPOAO model. Immunofluorescence staining for CD45 was performed on retinas at different time points post modelling to visualize the distribution and quantity of white blood cells (Fig. 7A-D). Laser confocal Z-plane projections confirmed minimal leukocyte infiltration in the retinas of sham eyes, which showed that CD45+ cells in the vascular lumen likely represented patrolling cells (Fig. 7A). A significant increase in CD45+ cells was observed in UPOAO model retinas after 1 day (Fig. 7E), with a progressive increase in quantity over time (Fig. 7F, G). Notably, the majority of CD45+ cells in UPOAO model retinas at 1-day displayed a morphology similar to that of vascular leukocytes (Fig. 7B), suggesting a large influx of peripheral white blood cells into the retinal tissue. In contrast, CD45+ cells at 3-days and 7-days of reperfusion exhibited a combination of amoeboid and branched morphologies (Fig. 7C, D), indicative of activated states.

Peripheral Leukocyte Infiltration and Retinal Resident Microglial Activation.

Rhodamine-labeled canavalin A was used for immediate cardiac perfusion to visualize blood vessels, followed by CD45 immunofluorescent staining to observe the relationship between blood vessels and CD45+ cells in sham (A), 1-day perfusion group (B), 3-days perfusion group (C), and 7-days perfusion group (D). The presence of CD45+ cells within blood vessels is indicated by white arrows. Scale bar = 50 μ m. CD45+ cell counts were performed in whole retinas of 1-day (E), 3-days (F), 7-days (G). The cellular morphology and distribution of microglial cells in the superficial retina were assessed in 3-days (H) and 7-days (I). Activated microglial cells are indicated by white asterisks. Microglial cell counting was conducted in the superficial retina of 3-days (J) and 7-days (K). Data points for CD45+ cells were derived from four flat-mounted retinas, and data points for microglial cells were from five flat-mounted retinas. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t-test. Scale bar = 50 μm.

Microglia, which are known to undergo activation and morphological changes in response to various insults, were examined in the superficial retinas of sham and UPOAO eyes. Immunofluorescence staining for Iba1 was conducted on flat-mounted retinas at 3-days and 7-days post-reperfusion. In sham eyes, Iba1+ microglia displayed small somas and elongated dendrites that were evenly distributed within the retina (Fig. 7H1, I1). Conversely, UPOAO retinas exhibited numerous activated Iba1+ microglia with enlarged somas, shortened dendrites, and an ameboid appearance (Fig. 7H2, I2). The number of Iba1+ microglia in UPOAO retinas was approximately five times greater than that in sham eyes at 3-days (Fig. 7J), with a further increase observed at 7-days (Fig. 7K), although this increase was less pronounced than that at the 3-days time point (Fig. 7J). These results indicated a potential reduction in immune-mediated inflammation over time.

RNA-seq comparison between UPOAO and extravascular occlusion models

To further study the characteristics of UPOAO, we analyzed the transcriptomes of two extravascular occlusion models: HIOP and UCCAO. The HIOP model induces ischemia through anterior chamber perfusion with normal saline (Fig. 8A), while the UCCAO model involves ligation of the unilateral common carotid artery (Fig. 8H). RNA-seq analysis of retinas subjected to 60 minutes of ischemia followed by 7-days of reperfusion in the HIOP model revealed enrichment of immune-related pathways, specifically highlighting increased activity in leukocyte-mediated immunity and regulation of immune effector processes (Fig. 8B). This finding was consistent with the results of GO analysis, which revealed associations of the overlapping DEGs between the HIOP and UPOAO models with toll-like receptor binding, T-cell receptor binding, and immune-related functions (Fig. 8C, D). To further explore these DEGs, we explored the remaining DEGs in the HIOP model in addition to the overlapping genes. GO analysis revealed that the remaining DEGs in the HIOP model were also enriched in immune responses, such as the T-cell receptor complex pathway, adaptive immune response, and B-cell-mediated immunity function (Fig. 8E). Remarkably, upon examining the remaining DEGs specific to the UPOAO model, GO analysis revealed DEGs distinctly related to lipid and steroid metabolic processes (Fig. 8F, G). This observation indicates that the UPOAO model, which is closely related to RAO, involves not only conventional immune responses but also unique regulation of lipid and steroid metabolism.

Transcriptomic Results Comparison between UPOAO, HIOP, and UCCAO Models.

RNA-seq comparison between UPOAO and extravascular occlusion models: HIOP and UCCAO. (A) Schematic illustration of the HIOP model. (B) GO analysis of differential expression genes(DEGs) in the HIOP model at 7-days perfusion. (C) Venn diagram indicating the overlapping DEGs (146 genes) between HIOP and UPOAO models, as well as the remaining DEGs specific to each group (593 genes in each group). (D) GO analysis of the overlapping DEGs between HIOP and UPOAO models. (E) GO analysis of the remaining DEGs in the HIOP model at 7-days perfusion, excluding the overlapping DEGs. (F) GO analysis of the remaining DEGs, excluding the overlapping DEGs. (G) Heatmap showing the inter-sample distribution of lipid and steroid-related DEGs from the analysis of the remaining DEGs in the UPOAO model. (H) Schematic illustration of the UCCAO model. (I) GO analysis of DEGs in the UCCAO model. (J) Hub genes identified through protein-protein interaction (PPI) network analysis of the DEGs in the UCCAO model. (K) Venn diagram indicating the overlapping DEGs between UCCAO and UPOAO models (15 genes). (L) Venn diagram indicating the presence of two overlapping DEGs between UCCAO DEGs and m6A-related genes.

Similarly, the UCCAO model exhibited associations with negative regulation of B-cell proliferation, lymphocyte activation, and related processes (Fig. 8I). Additionally, the hub genes identified through protein‒ protein interaction (PPI) analysis are presented (Fig. 8J). However, the UCCAO model displayed minimal overlap with the UPOAO model, with only 15 overlapping DEGs identified in the Venn diagram analysis (Fig. 8K). Further analysis of the UCCAO DEGs revealed two genes (Mettl14 and YTHDF2) related to m6A modifications (Fig. 8L). Although arterial blood flow is occluded in both the UPOAO and UCCAO models, our results revealed that the pathophysiological processes of the two models were particularly different after 7-days of reperfusion.

Discussion

This study aimed to develop a more appropriate mouse model for investigating the ischemia-reperfusion process in RAO and unravelling the underlying pathophysiological mechanisms. We combined silicone wire embolization with carotid artery ligation to effectively block the blood supply to the retinal artery, closely mimicking the characteristics of the acute interruption of blood supply in RAO patients [31]. In the UPOAO model, a 60-minute ischemia period was sufficient to simulate the injury of major retinal neural cells, such as RGCs, BCs, HCs, and cholinergic amacrine cells, resulting in visual impairment. Moreover, histologic examination demonstrated thinning of the inner layer of the retina, especially the GCL. Time course transcriptome analysis revealed various pathophysiological processes related to immune cell migration, oxidative stress, and immune inflammation during non-reperfusion and reperfusion periods. The resident microglia within the retina and peripheral leukocytes that access the retina were markedly increased during reperfusion periods. Additionally, we compared the transcriptomic signatures of the UPOAO model with those of two commonly used ischemia-reperfusion mouse models (HIOP and UCCAO), highlighting the potential relevance of the UPOAO model to ocular vascular occlusive diseases.

The UPOAO mouse model effectively simulates the characteristic features of RAO observed in patients. RAO patients often exhibit distinctive patterns in dark-adapted ERGs, where b-waves are reduced while a-waves remain stable [32-34]. In our study, we observed almost the same phenomenon in ERG amplitude after 60 minutes of ischemia in the UPOAO model (Table S2). This observation was further confirmed by immunofluorescence staining of bipolar cells and photoreceptor cells. Importantly, a previous study linked the thinning of the inner retina with alterations in ERG in patients with RAO [35]. Our results showed that ischemic damage predominantly affected the inner retinal layers, while the outer layer was almost unaffected (Table S2). This distinction is attributed to the reliance of the inner layers on the retinal artery for blood supply, whereas the outer layers are supplied by the choroid. Additionally, histological examination revealed increased retinal NFL and GCL thickness and decreased IPL thickness after 3-days of reperfusion, indicating that ischemia-induced oedema had not completely subsided.

This study comprehensively explored the transcriptomic signature after ischemia and revealed that resident and peripheral immune cells may play major roles in pathological processes. In the non-reperfusion group, the transcriptional changes were primarily involved in the migration of immune cells such as leukocytes and neutrophils from blood vessels into ischemic retinal tissue. This process is closely linked to injury during ischemia and reperfusion, where leukocytes and neutrophils infiltrate neural tissue through the vascular endothelium [36, 37]. Our results revealed an increase in CD45+ leukocytes in the retina and retinal microvasculature, suggesting that these leukocytes were recruited from the bloodstream to the damaged retinal tissue following reperfusion. During leukocyte access to the retina, chemokines, chemokine receptors, adhesion molecules and cytoskeletal components play important roles in regulating endothelial permeability and facilitating leukocyte adhesion to endothelial cells [38, 39]. Our findings of related gene expression are consistent with this mechanism.

During the 3-days reperfusion period, we observed that an increased number of DEGs were significantly involved in the critical pathological response to oxidative stress in the context of IR tissues. The generation of reactive oxygen species (ROS) in mitochondria and subsequent oxidative stress are widely recognized as major causes of retinal cell damage induced by IR [40, 41]. This finding suggested that mitochondrial function, which is involved in oxidative stress processes, may play an important role in the pathophysiology of IRI during the early reperfusion stage. Oxidative stress can influence the permeability of the inner blood retina barrier (iBRB), allowing leukocytes to access retinal tissues [42]. The infiltration and activation of immune cells are recognized as the underlying causes of many retinal diseases, including ischemic ophthalmia [42, 43]. In the UPOAO model, the increase in CD45+ leukocytes and Iba1+ microglia in the retina after 3-days of reperfusion supports this viewpoint.

In the later stage of reperfusion (7-days), the DEGs were mainly enriched in immune regulation and inflammation. With more upregulated DEGs than downregulated ones, we hypothesize that the activation of the immune inflammatory response contributes to further IRI in retinal tissue. Previous studies have shown that the retina can elicit immunological responses during ischemia-reperfusion injury and that immune inflammation is an important phenomenon in the progression of this injury [30]. The excessive ROS generated by mitochondria in RGCs during the activation of inflammatory responses can damage cell structure and visual function [44]. Microglia, considered the primary resident immune cells, contribute to inflammatory responses and consequent neural damage [45]. Infiltrating leukocytes can activate resident microglia, forming a feedback loop that exacerbates inflammation [46]. Complex interactions between immune cells and retinal neurons after RIRI have been reported [47]. Interestingly, in our UPOAO model, the death of major retinal neural cells, such as RGCs, BCs, and HCs, was correlated with an increase in infiltrating leukocytes and resident microglia. Therefore, we considered that the immune response and neuroinflammation observed after 7-days of reperfusion may result from the cumulative effects of acute oxidative stress and the responses of resident and peripheral immune cells (Fig. 9, Table S2). In summary, we described the pathological processes at different time points and highlighted the important role of resident and peripheral immune cell responses in the UPOAO model. This finding offers valuable insights for subsequent screening of pathogenic genes and potential immunotherapy approaches.

The Characteristic Features of UPOAO Model during Ischemia-Reperfusion Periods.

Compared with the HIOP and UCCAO models, the UPOAO model is a more appropriate choice for studying retinal IRI in RAO. The HIOP model, widely used for investigating the pathogenesis of APACG, exhibits retinal degeneration features similar to those of the UPOAO model during both reperfusion periods [48]. In our UPOAO model, histologic and immunohistochemical analysis demonstrated that 60 minutes of ischemia followed by 3-days or 7-days of reperfusion can cause irreversible damage to retinal structure and visual function. Specifically, we observed distinct alterations, such as thinning of the inner retina, substantial RGCs apoptosis, and decreased b-wave amplitude. Multiple studies on the HIOP model have reported consistent results [45, 49, 50]. However, some typical phenotypes in the HIOP model, such as thinning of the outer nuclear layer and the entire neuroretina and decreased a-wave amplitude, were not observed in the UPOAO model. Moreover, our transcriptome sequencing revealed a specific subset of DEGs unique to the UPOAO model, including Apoe, Abca1, Ldlr, Cyp39a1, and Bmp6. Notably, these DEGs were significantly enriched in pathways associated with lipid and steroid synthesis. Steroid and lipid metabolism homeostasis is disrupted under conditions of oxidative stress and inflammation [51]. In the UPOAO model, the observed lipid and steroid biological processes may result from the cumulative pathological responses triggered by IRI, as observed in the HIOP model. In the UCCAO model, RNA-seq revealed that DEGs were enriched in two main pathways: 1) epigenetic modification-related pathways, including nucleobase-containing compound catabolic process, RNA catabolic process, mRNA catabolic process and N6-methyladenosine (m6A) modification; and 2) cell death pathways, including regulation of autophagosome assembly, negative regulation of neuron death, and negative regulation of the neuronal apoptotic process. Epigenetic mechanisms may play a key role in the pathophysiology of ocular disease [52]. m6A, one of the most common RNA modifications, has been reported to regulate cell death processes, including apoptosis and autophagy, in the pathological process of IRI [53]. These results suggest that epigenetic mechanisms may significantly influence cell death during the 7-days reperfusion period in the UCCAO model. Lee et al. evaluated the characteristics of UCCAO without reperfusion using visual, histological and immunohistochemical approaches. Their results revealed delayed perfusion of the ipsilateral retina, thinning of the inner retinal layer 10 weeks after surgery, and a dramatic decrease in the amplitudes of b-waves on day 14 after UCCAO [15, 54]. This finding suggested that UCCAO primarily represents a model of retinal hypoperfusion injury and may not effectively reflect acute ischemic-induced structural and functional damage in RAO (Fig. 9). These results led us to consider the UPOAO model an effective experimental model for studying the pathological processes underlying acute ischemia and IRI.

Our research has certain limitations that should be acknowledged. First, our study focused on the changes that occurred within 60 minutes of ischemia and within the first 7-days of reperfusion in the UPOAO model. Further exploration is needed to understand the changes induced by longer reperfusion periods. Additionally, we proposed possible pathological mechanisms, including iBRB damage, oxidative stress and immune inflammation, mainly based on the enrichment results of DEGs at three reperfusion time points. More molecular experimental validation is required to substantiate these proposed mechanisms.

In summary, by thoroughly exploring the injury to major retinal neural cells, visual impairment and pathophysiological changes in the UPOAO model, we confirmed that this model can effectively simulate the acute ischemia-reperfusion processes in RAO. It serves as an ideal mouse model for investigating the underlying pathological mechanisms of ischemia and reperfusion. Furthermore, our UPOAO model holds great promise as a novel model for studies on pathogenic genes and potential therapeutic interventions for RAO.

Supporting information

eLife Assessment

The manuscript establishes a sophisticated mouse model for acute retinal artery occlusion (RAO) by combining unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) with a silicone wire embolus and carotid artery ligation, generating ischemia-reperfusion injury upon removal of the embolus. This clinically relevant model is useful for studying the cellular and molecular mechanisms of RAO. The data overall are solid, presenting a novel tool for screening pathogenic genes and promoting further therapeutic research in RAO.

Reviewer #1 (Public Review):

Summary:

Wang, Y. et al. used a silicone wire embolus to definitively and acutely clot the pterygopalatine ophthalmic artery in addition to carotid artery ligation to completely block blood supply to the mouse inner retina, which mimic clinical acute retinal artery occlusion. A detailed characterization of this mouse model determined the time course of inner retina degeneration and associated functional deficits, which closely mimic human patients. Whole retina transcriptome profiling and comparison revealed distinct features associated with ischemia, reperfusion, and different model mechanisms. Interestingly and importantly, this team found a sequential event including reperfusion-induced leukocyte infiltration from blood vessels, residual microglial activation, and neuroinflammation that may lead to neuronal cell death.

Strengths:

Clear demonstration of the surgery procedure with informative illustrations, images, and superb surgical videos.

Two time points of ischemia and reperfusion were studied with convincing histological and in vivo data to demonstrate the time course of various changes in retinal neuronal cell survivals, ERG functions, and inner/outer retina thickness.

The transcriptome comparison among different retinal artery occlusion models provides informative evidence to differentiate these models.

The potential applications of the in vivo retinal ischemia-reperfusion model and relevant readouts demonstrated by this study will certainly inspire further investigation of the dynamic morphological and functional changes of retinal neurons and glial cell responses during disease progression and before and after treatments.

Weaknesses:

It would be beneficial to the manuscript and the readers if the authors could improve the English of this manuscript by correcting obvious grammar errors, eliminating many of the acronyms that are not commonly used by the field, and providing a reason why this complicated but clever surgery procedure was designed and a summary table with time course of all the morphological, functional, cellular, and transcriptome changes associated with this model.

Reviewer #2 (Public Review):

Summary:

The authors of this manuscript aim to develop a novel animal model to accurately simulate the retinal ischemic process in retinal artery occlusion (RAO). A unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model was established using silicone wire embolization combined with carotid artery ligation. This manuscript provided data to show the changes of major classes of retinal neural cells and visual dysfunction following various durations of ischemia (30 minutes and 60 minutes) and reperfusion (3 days and 7 days) after UPOAO. Additionally, transcriptomics was utilized to investigate the transcriptional changes and elucidate changes in the pathophysiological process in the UPOAO model post-ischemia and reperfusion. Furthermore, the authors compared transcriptomic differences between the UPOAO model and other retinal ischemic-reperfusion models, including HIOP and UCCAO, and revealed unique pathological processes.

Strengths:

The UPOAO model represents a novel approach for studying retinal artery occlusion. The study is very comprehensive.

Weaknesses:

Originally, some statements were incorrect and confusing. However, the authors have made clarifications in the revised manuscript to avoid confusion.

Author response:

The following is the authors’ response to the current reviews.

eLife assessment:

The manuscript establishes a sophisticated mouse model for acute retinal artery occlusion (RAO) by combining unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) with a silicone wire embolus and carotid artery ligation, generating ischemia-reperfusion injury upon removal of the embolus. This clinically relevant model is useful for studying the cellular and molecular mechanisms of RAO. The data overall are solid, presenting a novel tool for screening pathogenic genes and promoting further therapeutic research in RAO.

Thank you for your thorough evaluation. We are pleased that you find our mouse model for acute retinal artery occlusion to be sophisticated and clinically relevant. Your recognition of the model’s utility in studying the cellular and molecular mechanisms of RAO, as well as its potential for advancing therapeutic research, is highly encouraging and underscores the significance of our work. We are grateful for your supportive feedback.

Public Reviews:

Reviewer #1:

Summary:

Wang, Y. et al. used a silicone wire embolus to definitively and acutely clot the pterygopalatine ophthalmic artery in addition to carotid artery ligation to completely block blood supply to the mouse inner retina, which mimic clinical acute retinal artery occlusion. A detailed characterization of this mouse model determined the time course of inner retina degeneration and associated functional deficits, which closely mimic human patients. Whole retina transcriptome profiling and comparison revealed distinct features associated with ischemia, reperfusion, and different model mechanisms. Interestingly and importantly, this team found a sequential event including reperfusion-induced leukocyte infiltration from blood vessels, residual microglial activation, and neuroinflammation that may lead to neuronal cell death.

Strengths:

Clear demonstration of the surgery procedure with informative illustrations, images, and superb surgical videos.

Two time points of ischemia and reperfusion were studied with convincing histological and in vivo data to demonstrate the time course of various changes in retinal neuronal cell survivals, ERG functions, and inner/outer retina thickness.

The transcriptome comparison among different retinal artery occlusion models provides informative evidence to differentiate these models.

The potential applications of the in vivo retinal ischemia-reperfusion model and relevant readouts demonstrated by this study will certainly inspire further investigation of the dynamic morphological and functional changes of retinal neurons and glial cell responses during disease progression and before and after treatments.

We sincerely appreciate your detailed and positive feedback. These evaluations are invaluable in highlighting the significance and impact of our work. Thank you for your thoughtful and supportive review.

Weaknesses:

The revised manuscript has been significantly improved in clarity and readability. It has addressed all my questions convincingly.

Thank you for your positive feedback. We are pleased to hear that the revisions have significantly improved the manuscript's clarity and readability, and that we have convincingly addressed all your questions. Your encouraging words are of great importance to us.

Reviewer #2 (Public Review):

Summary:

The authors of this manuscript aim to develop a novel animal model to accurately simulate the retinal ischemic process in retinal artery occlusion (RAO). A unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model was established using silicone wire embolization combined with carotid artery ligation. This manuscript provided data to show the changes of major classes of retinal neural cells and visual dysfunction following various durations of ischemia (30 minutes and 60 minutes) and reperfusion (3 days and 7 days) after UPOAO. Additionally, transcriptomics was utilized to investigate the transcriptional changes and elucidate changes in the pathophysiological process in the UPOAO model post-ischemia and reperfusion. Furthermore, the authors compared transcriptomic differences between the UPOAO model and other retinal ischemic-reperfusion models, including HIOP and UCCAO, and revealed unique pathological processes.

Strengths:

The UPOAO model represents a novel approach for studying retinal artery occlusion. The study is very comprehensive.

Thank you for your positive feedback. We are delighted that you find the UPOAO model to be a novel and comprehensive approach to studying retinal artery occlusion. Your recognition of the depth and significance of our study is highly valuable and encourages us in our ongoing research.

Weaknesses:

Originally, some statements were incorrect and confusing. However, the authors have made clarifications in the revised manuscript to avoid confusion.

We sincerely appreciate your meticulous review of the manuscript. We have thoroughly addressed the inaccuracies identified in the revised version. Additionally, we have polished the article to ensure improved readability. We apologize for any confusion caused by these inaccuracies and genuinely. We appreciate your careful attention to detail, and your patience and meticulous suggestions have significantly improved the clarity and readability of our manuscript.

The following is the authors’ response to the original reviews.

Recommendations for the authors:

Reviewer #1:

The revised manuscript has been significantly improved in clarity and readability. It has addressed all my questions convincingly.

Thank you for your positive feedback. We are pleased to hear that the revisions have significantly improved the manuscript's clarity and readability, and that we have convincingly addressed all your questions. Your encouraging words are of great importance to us.

Reviewer #2:

The authors have revised the manuscript and/or provided answers to the majority of prior comments, which have helped to strengthen the work. However, addressing the following concerns is still necessary to further improve the manuscript.

Thank you for acknowledging our revisions and the improvements made to the manuscript. We appreciate your continued feedback and will address the remaining concerns to further enhance the quality of our work.

The quantification method of RGCs is described in detail in the response letter, but this detailed methodology was not included in the revised manuscript to clarify the quantification process.

Thank you for your helpful recommendations. We have added detailed methodology in the revised manuscript to clarify the quantification process (line 180-188).

The graphs in Fig. 3D b-wave and Fig. 3E-b wave are duplicated.

We apologize for the error in our figures. We have corrected the mistake by replacing the duplicated image in Fig. 3E-b wave with the correct one (line 880). Your careful observation has been very helpful in improving our manuscript. Thank you for bringing this to our attention.

The quantifications of the thickness of retinal layers in HE-stained sections in Figure 4 (IPL) and Response Figure 2 are incorrect. For mice retina, the thickness of the IPL is approximately 50 µm.

Thank you for your meticulous review of the manuscript. We have rectified the inaccuracies in the quantification of retinal layer thickness in HE-stained sections in Figure 4, addressing the initial issue with the scale bar.

We consulted with a microscope engineer and used a microscope microscale to calibrate the scale of the fluorescence microscope (BX63; Olympus, Tokyo, Japan) at the suggestion of the engineer.

We recount the thickness of all layers of the HE-stained retinal section (line 902). The inner retina thickness in Figure 4 has been adjusted under a new scale bar, and the thickness of the outer retinal layers is now displayed in

Author response image 1. However, the IPL thickness of the sham eye in the UPOAO model is still not aligned with the common thickness of 50 µm. Therefore we review the literature within our laboratory, focusing on C57BL/6 mice from the same source, revealed that the inner retina thickness (GCC+INL) in the HE-stained sections of the sham eye in the UPOAO model (around 80 µm) is consistent with previous findings (see Author response image 2) conducted by Kaibao Ji and published in Experimental Eye Research in 2021 [1].

We captured and analyzed the average retinal thickness of each layer over a long range of 200-1100 μm from the optic nerve head (see Author response image 3, highlighted by the green line). The field region has been corrected in the revised manuscript (line 232). Considering the significant variation in retinal thickness from the optic nerve to the periphery, we consulted literature on multi-point measurements of HE-stained retinas. The average thickness of the GCC layer in the control group was approximately 57 µm at 600 µm from the optic nerve head and about 48 µm at 1200 µm from the optic nerve head in the literature [2] (see Author response image 4). The GCC layer thickness of the sham eye in the UPOAO model is around 50 µm, in alignment with existing literature. In future studies, we will pay more attention to the issue of thickness averaging.

We appreciate your thorough review and valuable feedback, which has enabled us to correct errors and enhance the accuracy of our research.

Thickness of OPL, ONL, IS/OS+RPE in HE staining.

n=3; ns: no significance (p>0.05).

Cited from Ji, K.

, et al., Resveratrol attenuates retinal ganglion cell loss in a mouse model of retinal ischemia reperfusion injury via multiple pathways. Experimental Eye Research, 2021. 209: p. 108683.

Schematic diagram illustrating the selection of regions.

The figure was captured using a fluorescence microscope (BX63; Olympus, Tokyo, Japan) under a 4X objective. Scale bar=500 µm.

Cited from Feng, L.

, et al., Ripa-56 protects retinal ganglion cells in glutamate-induced retinal excitotoxic model of glaucoma. Sci Rep, 2024. 14(1): p. 3834.

There are some typos in the summary table. For example: 'Amplitudes of a-wave (0.3, 2.0, and 10.0 cd.s/m²)' should be 'Amplitudes of a-wave (0.3, 3.0, and 10.0 cd.s/m²)'; and 'IINL thickness' in HE' should be 'INL thickness'.

Thank you for pointing out the typos in the summary table (line 1073). We have corrected 'Amplitudes of a-wave (0.3, 2.0, and 10.0 cd.s/m²)' to 'Amplitudes of a-wave (0.3, 3.0, and 10.0 cd.s/m²)' and 'IINL thickness' to 'INL thickness'. Your attention to detail is greatly appreciated and has been very helpful in improving our manuscript.

References

(1) Ji, K., et al., Resveratrol attenuates retinal ganglion cell loss in a mouse model of retinal ischemia reperfusion injury via multiple pathways. Experimental Eye Research, 2021. 209: p. 108683.

(2) Feng, L., et al., Ripa-56 protects retinal ganglion cells in glutamate-induced retinal excitotoxic model of glaucoma. Sci Rep, 2024. 14(1): p. 3834.

Version 3

Abstract

Abstract

Acute retinal ischemia and ischemia-reperfusion injury are the primary causes of retinal neural cell death and vision loss in retinal artery occlusion (RAO). The absence of an accurate mouse model for simulating the retinal ischemic process has hindered progress in developing neuroprotective agents for RAO. We developed a unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) mouse model using silicone wire embolization combined with carotid artery ligation. The survival of retinal ganglion cells and visual function were evaluated to determine the duration of ischemia. Immunofluorescence staining, optical coherence tomography, and haematoxylin and eosin staining were utilized to assess changes in major neural cell classes and retinal structure degeneration at two reperfusion durations. Transcriptomics was employed to investigate alterations in the pathological process of UPOAO following ischemia and reperfusion, highlighting transcriptomic differences between UPOAO and other retinal ischemia-reperfusion models. The UPOAO model successfully replicated the acute interruption of retinal blood supply observed in RAO. 60-minutes of Ischemia led to significant loss of major retinal neural cells and visual function impairment. Notable thinning of the inner retinal layer, especially the ganglion cell layer, was evident post-UPOAO. Temporal transcriptome analysis revealed various pathophysiological processes related to immune cell migration, oxidative stress, and immune inflammation during the non-reperfusion and reperfusion periods. A pronounced increase in microglia within the retina and peripheral leukocytes accessing the retina was observed during reperfusion periods. Comparison of differentially expressed genes (DEGs) between the UPOAO and high intraocular pressure models revealed specific enrichments in lipid and steroid metabolism-related genes in the UPOAO model. The UPOAO model emerges as a novel tool for screening pathogenic genes and promoting further therapeutic research in RAO.

Introduction

Retinal artery occlusion (RAO) is a severe ophthalmic disease characterized by a sudden interruption of blood flow in the retinal artery, leading to retinal ischemia [1]. Over 60% of RAO patients suffer from impaired vision, ranging from finger counting to complete vision loss [2]. Additionally, RAO patients face increased risk of cardiovascular and cerebrovascular events [3–5]. Retinal ischemia and hypoxia can lead to irreversible damage to retinal cells within 90 minutes [6]. This damage is primarily due to the apoptosis of retinal ganglion cells (RGCs), which is driven by inflammation and oxidative stress. Unfortunately, conservative treatments for RAO, such as ocular massage and hyperbaric oxygen therapy, provide limited therapeutic benefits [7]. Although thrombolysis is effective, its application is restricted by a narrow therapeutic window (typically within 4.5 hours) [8]. Furthermore, the restoration of blood flow to the ischemic area following thrombolysis can cause a sudden increase in tissue oxidative levels, resulting in ischemia-reperfusion injury (IRI) in the retina [9]. IRI is a common pathological condition that can induce inflammation, retinal tissue damage and visual impairment [10]. Therefore, it is crucial to develop an animal model that accurately simulates the pathological processes of RAO to thoroughly investigate the pathophysiological changes and explore potential neuroprotective treatments following retinal ischemia and reperfusion.

Various animal models have been employed to investigate the effects of retinal injury resulting from ischemia and subsequent reperfusion in RAO [6, 11, 12]. Based on the methods used for modelling, retinal ischemia-reperfusion models can be categorized into two groups: 1) intravascular occlusion models, including photochemical-induced thrombosis model and vascular intervention model; and 2) extravascular occlusion models, including central retinal artery ligation (CRAL) model, unilateral common carotid artery occlusion (UCCAO) model, and high intraocular pressure (HIOP) model. The photochemical-induced thrombosis model and vascular intervention model have been reported to be useful tools for assessing retinal injury after ischemia and reperfusion. However, both models are limited by the need for trained interventional radiologists and the need for advanced techniques [13, 14]. The CRAL model, which directly clips or ligates the central retinal artery, also presents drawbacks, such as potential optic nerve damage and difficulty in application to small animals such as mice, limiting its use in experiments with larger sample sizes [6]. The UCCAO model, which induces retinal hypoperfusion by occluding the unilateral common carotid artery (CCA), is more suitable for simulating chronic ischemic retinal disease than acute retinal ischemia [15, 16]. The HIOP model is widely utilized in mice for studying ischemia-reperfusion in acute primary angle-closure glaucoma (APACG) [17]. However, the mechanical compression of the retina caused by saline injection in this model may also lead to retinal damage, potentially influencing IRI research. Therefore, it is critically important to develop a simple and low-skill-required mouse model that can simulate acute retinal ischemia and reperfusion injury in RAO patients.

To better simulate the retinal ischemic process and possible IRI following RAO, we developed a novel vascular-associated mouse model called the unilateral pterygopalatine ophthalmic artery occlusion (UPOAO) model. In this model, we employed silicone wire embolization and carotid artery ligation to completely block the blood supply to the retina. We characterized the major classes of retinal neural cells and evaluated visual function following different durations of ischemia (30 minutes and 60 minutes) and reperfusion (3-days and 7-days) after UPOAO. Additionally, we utilized transcriptomics to investigate the transcriptional changes and elucidate the pathophysiological process of the UPOAO model after ischemia and reperfusion. Finally, we highlighted the transcriptomic differences between this model and other retinal ischemia-reperfusion models, including HIOP and UCCAO, revealing the unique pathological processes that closely resemble retinal IRI in RAO. The UPOAO model offers new insights into the mechanisms and pathways involved in ischemia and reperfusion studies of RAO, providing a foundation for studying protective strategies for ocular ischemic diseases.

Methods and materials

Animals

Eight-week-old male C57BL/6 mice were used in the experiments. Only male mice were used to exclude the potential influence of oestrogenic hormones. The mice were provided sufficient food and water and were maintained on a 12-hour dark-light cycle in a room with regulated temperature conditions. All the experimental procedures were designed and conducted according to the ethical guidelines outlined in the Association for Research in Vision and Ophthalmology (ARVO) Statement for the Use of Animals in Ophthalmic and Vision Research. The study protocol and methods received approval from the Experimental Animal Ethics Committee of Renmin Hospital of Wuhan University (approval number: WDRM-20220305A).

Preparation and surgical procedure of the UPOAO model

Eight-week-old male C57BL/6 mice weighing 20-25 grams were used in this study. An isoflurane-based anaesthesia system was used to induce and maintain general anaesthesia in the mice during surgical procedures. The mice were anaesthetized with a 1.5-2% concentration of isoflurane delivered in a mixture of nitrous oxide and oxygen via rubber tubing. Body temperature was maintained at 37 ± 0.5°C throughout the surgery. Surgical instruments were sterilized with 75% ethanol before use to ensure sterility.

We drew inspiration from the widely employed middle cerebral artery occlusion (MCAO) model [18, 19], commonly used in cerebral ischemic injury research, which guided the development of the UPOAO model. The mouse was gently positioned in a supine posture on a heating blanket, and its neck was exposed. The preparation involved depilation of the neck area, followed by skin disinfection, and finally, a midline incision along the neck was made. After separating the neck gland using two tweezers, careful blunt dissection was employed to separate the left CCA, internal carotid artery (ICA) and external carotid artery (ECA), avoiding the compression of nearby nerves and veins (Fig. 1A) (Video 1).

Modeling Procedure, Validation, and Cervical Artery Anatomy

(A1-G1) Schematic illustration of unilateral pterygopalatine ophthalmic artery occlusion (UPOAO). (A2-G2) Practical operation of UPOAO. (A1, A2) Blunt separation and exposure of the left cervical arteries. (B1, B2) Arterial suture ligation. (C1, C2) Insertion of the silicone wire embolus. The artery was incised to create a hole, and the silicone wire embolus was inserted. (D1, D2) Artery disconnection and movement of the silicone wire embolus. The artery was cut along the incision, and the silicone wire embolus was retracted and reinserted. (E1, E2) Removal of the silicone wire embolus and reperfusion. (F1, F2) Suture removal. The sutures at both ends of the disconnected vessel were knotted, and the other two sutures were removed. (G1, G2) Anatomic reduction and suturing of the skin. (H, I) Canavalin A label vasculature of the UPOAO mouse retina. The silicone wire embolus was inserted into artery before perfusing rhodamine-labeled canavalin A into the heart of UPOAO mouse. The sham eye served as an unpracticed control eye, while the UPOAO lateral eye represented the experimental eye. Retinal vessels in the sham eye (H) exhibited fluorescence filling, while retinal vessels in the UPOAO lateral eye (I) remained unfilled. Scale bar = 1 mm. (J and K) Fluorescein Fundus Angiography (FFA) performed before removing the silicone wire embolus from the UPOAO mouse. The vessels in the sham lateral retina (J) were perfused, while the lateral retinal perfusion in UPOAO (K) was delayed. (L) Schematic illustration of cervical artery anatomy and ocular blood supply. Embolization of the PPA resulted in ocular ischemia. The red arrow indicates the site of the silicone wire embolus occlusion. The silicone wire embolus used a type 602156 wire, extended to 7mm with a diameter of 0.21mm. The blue arrows indicate the modelling locations of the HIOP model and the UCCAO model, respectively. CCA: common carotid artery; ICA: internal carotid artery; ECA: external carotid artery; PPA: pterygopalatine artery; MCA: middle cerebral artery; Infraorbital art.: infraorbital artery; OA: ophthalmic artery; SPCA: short posterior ciliary artery; LPCA: long posterior ciliary artery; CRA: central retinal artery.

Then, the CCA and ICA were ligated with a 6-0 suture. A knot was secured using a 6-0 suture at the distal end of the ECA, and a slipknot was created at the proximal end (Fig. 1B). To preserve the reperfusion process and avoid bleeding and mortality during surgery, a silicone wire embolus was inserted through the ECA instead of through the CCA. Ophthalmic scissors were used to make a small inverted “V”-shaped incision between the two suture knots on the ECA. A specialized silicone wire embolus, measuring 7 ± 0.1 mm in length and 0.21 ± 0.1 mm in diameter, was inserted through the incision, directing it into the ECA and further into the CCA (Fig. 1C). Next, the ligation on the ICA was removed, and the ECA was cut at the point where the artery had been previously incised with scissors. Subsequently, the silicone wire embolus was retracted to the bifurcation of the CCA, rotated counterclockwise, and inserted into the ICA and further into the pterygopalatine artery (PPA) (Fig. 1D1). The silicone tail of the silicone wire embolus was positioned near the bifurcation of the CCA (Fig. 1D2), effectively obstructing the ophthalmic artery (OA) (Video 2). Then, the slipknot was fastened, and the skin was sutured. The mouse was free to move around during arterial embolization.

Following a predetermined ischemia embolization time, the silicone wire embolus was carefully removed from the PPA without massive haemorrhage (Fig. 1E). The CCA suture was removed, restoring arterial reperfusion (Fig. 1F) (Video 3). Then, the skin wound was sutured, and the mice were routinely fed during reperfusion (Fig. 1G).

Staining of retinal vessels

To clearly visualize the retinal vessels, rhodamine-labelled canavalin A was used for mouse heart perfusion. Each mouse was anaesthetized with 1% sodium pentobarbital and placed in the supine position on a foam board resting on a tray. The sternum was opened to expose the heart. A needle attached to the perfusion tube was inserted into the left ventricle via the apex of the mouse heart and secured. The right atrial appendage was dissected with scissors to allow blood outflow. Saline solution was perfused through the left ventricle using a pump system to drain blood from the arteries. Subsequently, the prepared solution of rhodamine-labelled canavalin A was injected into the mice and circulated throughout the mouse vessels for 30 minutes. The mice were then sacrificed, and their eyes were removed and immersed in paraformaldehyde (PFA) for one hour in the dark. A flat-mounted retina was obtained, and retinal images were captured using a Leica SP8 confocal microscope equipped with a 10× objective.

Quantification of RGCs and microglia

The mice were euthanized by cervical dislocation, and their eyes were fixed in 4% PFA for 60 minutes at room temperature (RT). Subsequently, the cornea and lens were excised, and the intact retina was isolated for retinal flat mounting. The retina was immersed in PBS supplemented with 5% bovine serum albumin (BSA) and 0.5% Triton X-100 for overnight blocking and then incubated with Brn3a antibody (Synaptic Systems, Germany) or Iba1 antibody (Wako, Japan). Following a two-day incubation with the primary antibody, the retina was gently washed three times with PBS. Subsequently, the retina was incubated with Alexa Fluor 594 (AntGene, Wuhan, China) in a light-protected cassette for two days. After three rinses, the retina was uniformly sectioned into a four-leaf clover morphology and flattened using a coverslip. Retinal flat mounts labelled with Brn3a were photographed to count using a fluorescence microscope (BX51, Olympus, Japan) and representative pictures were imaged using a Leica SP8 confocal microscope (Leica TCS SP8, Germany). Retinas labelled with Iba1 were imaged utilizing a fluorescence microscope (BX63; Olympus, Tokyo, Japan). Each quadrant of the retina was systematically subdivided into central, middle, and peripheral fields (distance from the optic nerve head: central field: 0.1 mm to 0.5 mm, middle field: 0.9 mm to 1.3 mm, and peripheral field: 1.7 mm to 2.1 mm). The surviving RGCs and activated microglia of 12 fields in each retina were quantified and averaged using ImageJ software (National Institutes of Health, USA).

Electroretinogram (ERG)

Mice were dark-adapted overnight before ERG, and all subsequent procedures were carried out in the dark. Before ERG, the mice were anaesthetized with 1% sodium pentobarbital via intraperitoneal injection, and their pupils were dilated. A subcutaneous electrode was inserted into the posterior cervical skin, a tail electrode was affixed to the posterior end of the mouse tail, and the corneal electrode was gently placed on the central corneal surface. A RetiMINER-C visual electrophysiological system (3VMED Co., Ltd., Shanghai, China) was used for recording electrical responses. The a-waves, b-waves, and oscillatory potentials (OPs) after flash stimuli of 0.01, 0.03, 0.1, 0.3, 1.0, 3.0, and 10.0 cd.s/m2 in scotopic adaptation were recorded. The amplitudes and implicit times of a-waves, b-waves, and OPs in response to various flash stimuli were analysed.

Optical Coherence Tomography Imaging (OCT) and Fluorescein Fundus Angiography (FFA)

The mice were placed on the platform of a Spectralis HRA + OCT device (Heidelberg Engineering, Heidelberg, Germany) for OCT imaging following anaesthesia. Pupil dilation was performed, and normal saline was applied regularly to maintain corneal moisture. The focal length of the device was adjusted until the mouse retina was clearly visible. The head of the mouse was gently repositioned to capture images of the peripheral fundus. The multiline mode was used to scan each layer of the retina, and four quadrants of view centred on the nipple in the upper left, lower left, upper right, and lower right were recorded. In this study, the total retinal thickness was manually segmented into three parts and encompassed the entire thickness from the nerve fibre layer to the photoreceptor layer. The ganglion cell complex (GCC) was defined as the combined thickness of the retinal nerve fibre layer (RNFL), ganglion cell layer (GCL), and inner plexiform layer (IPL). The inner nuclear layer (INL) and the outer plexiform layer (OPL) were combined for thickness analysis due to the difficulty in distinguishing between these layers. The remaining retinal layers, including the outer nuclear layer (ONL), inner segment/outer segment (IS/OS), and retinal pigment epithelium (RPE) layers, were measured together. The thickness of the retina at 1.5 papillary diameters (PD), 3.0 PD, and 4.5 PD from the centre of the optic disc was measured using the Heidelberg measuring tool.

For FFA, the mice were anaesthetized, and their pupils were dilated. Subsequently, fundus angiography was performed immediately following an intraperitoneal injection of sodium fluorescein. Images were acquired alternately for both eyes within 5 seconds of the injection.

Hematoxylin and eosin (HE)

Mouse eyes were fixed in FAS eyeball fixation solution (Service-bio, Wuhan, China) for 48 hours, followed by dehydration and subsequent embedding in paraffin. The paraffin blocks were trimmed parallel to the optic nerve to obtain 3-4 μm thick sections where the optic nerve was located. Six paraffin sections from each eyeball were stained with HE. Images were captured using an Olympus fluorescence microscope. The retinal thickness within a range of 200-1100 μm from the optic disc was measured using Image-Pro Plus 6.0 software.

Staining of Retinal Sections

For immunofluorescence, mouse eyes were fixed in 4% PFA solution and subjected to gradient dehydration using 10%, 20%, and 30% sucrose solutions. The following day, the eyes were embedded in optimum cutting temperature (OCT) compound (SAKURA, USA) and frozen at -80°C. Several 14-μm-thick frozen sections were obtained from each eyeball through the use of a freezing microtome (Leica, Wetzlar, Germany). The section surface was parallel to the optic nerve, and 3-6 frozen sections were mounted on a single slide. The slides were stored at -20°C until use.

All sections were blocked with 5% BSA and 0.5% Triton X-100 in PBS for 2 hours. Primary antibodies, as listed in Table 1, were diluted in PBS containing 5% BSA and 0.5% Triton X-100 and incubated at 4°C overnight. The following day, after three rinses, the sections were incubated with secondary antibodies at RT for 2 hours in a cassette, followed by three washes. The sections were then stained with DAPI (Service-bio, Wuhan, China) for 15 minutes. After the final three rinses, the frozen sections were sealed. A Leica SP8 instrument equipped with a 40× objective was used to photograph frozen sections near the optic nerve head. The fluorescence intensity of all slices of each eye was determined using ImageJ software.

Surgical procedure for the UCCAO model and the HIOP model

Animals for the UCCAO model were prepared and managed according to the procedure described for the UPOAO model. After blunt dissection, the left CCA was ligated with a 6-0 suture. After 60 minutes of ligation, the suture was removed, and the skin wound was subsequently closed with sutures. Mice were routinely fed during the reperfusion period, and the retinas of UCCAO mice were harvested 7-days post-surgery.

The HIOP model was generated based on previously established methods [20]. Briefly, mice were anaesthetized by intraperitoneal injection of 1% sodium pentobarbital. Pupils were dilated using tropicamide phenylephrine eye drops (0.5% tropicamide and 0.5% deoxyadrenaline hydrochloride, Santen Pharmaceutical) applied 5 minutes in advance. An insulin hypodermic needle attached to a silicone elastomer tube was inserted into the anterior chamber of the unilateral experimental eye. The IOP was elevated for 60 minutes by raising the height of the saline solution storage bag to 162 cm, after which the needle was removed to immediately restore the IOP to baseline levels. Iris whitening and loss of retinal red-light reflexes indicated retinal ischemia. After surgery, levofloxacin hydrochloride ophthalmic gel was applied to the eyes of the mice. Retinas from HIOP mice were collected 7-days post-surgery.

Transcriptome sequence and analysis

Retinal samples from UPOAO, HIOP, and UCCAO mice were extracted and promptly frozen in liquid nitrogen within 10 minutes of cervical dislocation. Total RNA extraction was performed according to the protocol outlined in the TRIzol Reagent manual (Life Technologies, CA, USA). RNA integrity and concentration were assessed using an Agilent 2,100 Bioanalyzer (Agilent Technologies, Inc., Santa Clara, CA, USA). The resulting RNA samples were then pooled and subjected to sequencing on the Illumina HiSeq3000 platform in a 150-bp paired-end read format. The raw RNA-sequencing (RNA-seq) reads were preprocessed and quantified using the featureCounts function in the SubReads package version 1.5.3, with default parameters.

RNA-seq analysis was conducted on samples from 5 mice in the non-perfusion group, 4 mice in the 3-days perfusion group, and 5 mice in the 7-days perfusion group of the UPOAO model. Additionally, retinas from both the 7-days HIOP experimental eyes and bilateral UCCAO eyes were collected (n=5) for analysis.

Data normalization and subsequent processing utilized the ‘limma’ package in R software (version 4.2.3) [21]. Volcano plots and Venn diagrams were generated for visualization to identify significantly differentially expressed genes (DEGs) and perform Gene Ontology (GO) annotation analysis. DEGs with a |log2-fold change (FC)| > 1 and adjusted p value < 0.05 were significantly differentially expressed. Enrichment analysis focused on DEGs with an adjusted p < 0.05 and an enriched gene count > 5. To visualize the expression patterns of significant DEGs, a heatmap was generated using the ‘pheatmap’ package. Gene Set Enrichment Analysis (GSEA) was conducted using the GSEA program (version 4.3.2) [22], employing the default gene set m-subset (mh.all.v2023.1.mm.symbols.gmt) to explore significant functional and pathway differences. Enriched pathways were classified based on criteria such as adjusted p value (< 0.05), false discovery rate (FDR) q value (< 0.25), and normalized enrichment score (|NES| > 1). Furthermore, protein ‒ protein interaction (PPI) analysis was performed using the Search Tool for the Retrieval of Interacting Genes (STRING) database (https://string-db.org/) [23]. Genes with a greater degree of protein-level interactions with others were further analysed using Cytoscape software (version 3.8.2) to generate a downstream PPI map. PPI pairings with an interaction score > 0.7 were extracted and visualized using Cytoscape 3.9.0 [24].

Real-time Quantitative polymerase chain reaction (RT-qPCR)

RNA was extracted from fresh mouse retinal samples and subjected to reverse transcription using HiScript lll RT SuperMix for qPCR (Vazyme, China) following the manufacturer’s instructions. RT‒qPCR analysis was conducted on a CFX instrument (Bio-Rad) using AceQ qPCR SYBR Green Master Mix (Vazyme, China). The PCR program comprised 40 cycles of 10 seconds at 95°C and 30 seconds at 60°C. Assays were performed in triplicate, and Ct values were normalized to beta-actin levels. Relative quantification of target gene expression was calculated using the 2-ΔΔCt method. The primer sequences are provided in Table S1.

Statistical Analysis

All the data were analysed using GraphPad Prism version 9.0 (GraphPad Software, San Diego, CA, USA). Each experimental group included a minimum of three biological replicates. T tests, including RGCs and microglia counting, OPs analysis, retinal thickness measurements via HE staining, and fluorescence intensity quantification, were used to compare the two groups. Two-way ANOVA (two-way analysis of variance) test was used for more than two different experimental groups, including ERG waves and retinal thickness measurements in OCT. All data are presented as the mean ± standard error of the mean (SEM), and the statistical graphs are shown as scatter bar graphs and line charts. P < 0.05 was considered to indicate statistical significance.

Results

Silicone wire embolus insertion interrupts retinal blood flow

To reproduce the retinal ischemic process and potential IRI in RAO, we established a mouse model of UPOAO by combining silicone wire embolization with carotid artery ligation. Model validation was conducted through a series of experiments, with detailed descriptions provided in the Materials and Methods section (Fig. 1A-G). To confirm blood flow disruption, we performed cardiac perfusion with fluorescently labelled lectin and FFA.

Rhodamine-labelled canavalin A was used during in vivo perfusion with silicone wire embolus, and both eyes were evaluated. The sham eye exhibited normal blood perfusion in the retina, while the experimental eye showed no perfusion (Fig. 1H, I). FFA revealed delayed and limited perfusion in the experimental lateral retina, primarily near the optic disc (Fig. 1J, K). These findings indicate that the insertion of the silicone wire embolus effectively impaired blood flow to the retina.

60-minute ischemia in UPOAO damage retinal structure and function

To investigate the optimal ischemic duration, the mice were subjected to ischemic periods of 30 and 60 minutes, followed by reperfusion periods of 3-days and 7-days. Retinal structure and visual function were assessed through flat-mounted retina analysis and flash ERG.

In the 30-minute ischemia group, no significant RGCs death was observed after 3-days (Fig. 2A, B) or 7-days of reperfusion (Fig. 2C, D). However, after 60 minutes of ischemia followed by either 3-days or 7-days of reperfusion, reductions in RGCs density were evident (Fig. 2E-H). ERG results showed no statistical difference in amplitudes between bilateral eyes after 3- and 7-days of 30-minutes reperfusion (Fig. 3A-C). However, the b-wave amplitude notably decreased after 60 minutes of ischemia and 3-days of reperfusion (Fig. 3D). By 7-days of reperfusion, the b-wave amplitude had halved compared to that of the sham eyes (Fig. 3E). The appearance times of the a-waves and b-waves in the experimental and control eyes remained consistent across all four groups (Fig. S1). Additionally, OPs have also been used to sensitively monitor retinal ischemic effects by detecting changes before any alterations in b-waves occur [25–27]. The amplitudes of OPs in the experimental eyes, particularly in the 60-minute ischemia and 7-days reperfusion group, significantly decreased to less than 50% of those in the sham eyes (Fig. 3F, G). Based on RGCs survival and changes in ERG waves, we determined that a 60-minute ischemic duration is optimal for the UPOAO model.

Staining and Quantification of Retinal Ganglion Cells (RGCs) at Different Ischemia and Reperfusion Times.

Flat-mounted retina RGCs were labeled with Brn3a staining. (A, B) Representative pictures of the peripheral field (A), and quantification of surviving RGCs in all fields (B) in the 30-minute ischemia and 3-days reperfusion group. n = 3. (C, D) Representative pictures of the peripheral field (C), and quantification of surviving RGCs in all fields (D) in the 30-minute ischemia and 7-days reperfusion group. n = 3. (E, F) Representative pictures of the peripheral field (E), and quantification of surviving RGCs in all fields (F) in the 60-minute ischemia and 3-days reperfusion group. n = 5. (G, H) Representative pictures of the peripheral field (G), and quantification of surviving RGCs in all fields (H) in the 60-minute ischemia and 7-days reperfusion group. n = 5. The results showed a significant loss of RGCs after 60 minutes of ischemia. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t-test. Scale bar = 100 μm.

Comparison of Electroretinographic (ERG) Dark-Adapted Responses at Different Ischemia and Reperfusion Times.

Following the evaluation of surviving RGCs, visual function in sham and UPOAO experimental eyes at various ischemia and reperfusion times was assessed using ERG. (A) Representative waveforms in the four groups at the stimulus light intensities of 0.1, 1.0, and 10.0 cd.s/m2, respectively. (B) Quantification of a-wave and b-wave amplitudes in the 30-minute ischemia and 3-days reperfusion group. n = 5. (C) Quantification of a-wave and b-wave amplitudes in the 30-minute ischemia and 7-days reperfusion group. n = 5. (D) Quantification of a-wave and b-wave amplitudes in the 60-minute ischemia and 3-days reperfusion group. n = 7. (E) Quantification of a-wave and b-wave amplitudes in the 60-minute ischemia and 7-days reperfusion group. n = 8. Dark-adapted responses showed almost similar a-wave amplitudes but significantly decreased b-wave amplitudes in the 60-minute ischemia groups. The amplitudes of b-waves declined at 3-days and even more prominently at 7-days. (F, G) Representative OPs and quantification of amplitudes in the 60-minute ischemic groups. n = 7 in the 3-days reperfusion group; n = 8 in the 7-days reperfusion group. The amplitudes of OPs decreased significantly at 7-days reperfusion. The decline in b-waves and OPs along with the loss of RGCs, supports the selection of a 60-minute ischemic duration as an appropriate choice. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test for a-waves and b-waves; paired t-test for OPs.

Evaluation of retinal thickness in UPOAO through OCT and HE

To assess retinal thickness non-invasively in the UPOAO model, we conducted OCT imaging during 3-days and 7-days reperfusion periods. Following a 3-days reperfusion period, no significant changes were observed in the GCC, INL + OPL, and ONL + IS/OS + RPE layers at 1.5, 3, and 4.5 PD from the optic disc (Fig. 4B, Fig. S2A, B). However, the total retinal thickness decreased at 4.5 PD (Fig. 4C). After 7-days of reperfusion, the total retinal thickness decreased at 1.5 PD, 3 PD, and 4.5 PD, primarily due to inner retinal thinning (Fig. 4D-F, Fig. S2C, D).

Changes in Retina Morphology in the UPOAO Model.

(A) Representative OCT images of the mouse retina at 3-days. The Green lines indicate the OCT scan area starting from the optic disc. Local magnification and layering are annotated. GCC: ganglion cell complex, including RNFL, GCL and IPL layers. (B and C) Quantification of GCC and total retinal thickness at 3-days. The thickness of the GCC and the entire retina in OCT was measured and compared at distances of 1.5 PD, 3.0 PD and 4.5 PD from the optic disc, respectively. n = 5. (D) Representative OCT images of the mouse retina at 7-days. (E and F) Quantification of GCC and total retinal thickness at 7-days. n = 5. (G) Representative HE images of the mouse retina at 3-days. (H, I and J) Quantification of NFL + GCL, IPL, and INL thickness at 3-days. Retinal thickness in HE was measured near the optic nerve head and compared. n = 3. (K) Representative HE images of the mouse retina at 7-days. (L, M and N) Quantification of NFL + GCL, IPL, and INL thickness at 7-days. n = 3. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, two-way ANOVA test in OCT and paired t-test in HE. PD: papillary diameters; RNFL: retinal nerve fiber layer; GCL: ganglion cell layer; IPL: inner plexiform layer; INL: inner nuclear layer; OPL: outer plexiform layer; ONL: outer nuclear layer; IS: inner segment; OS outer segment; RPE: retinal pigment epithelium. Scale bar = 100 μm.

For a detailed analysis of inner retinal thickness, we extracted UPOAO mouse eyes and stained them with HE. The thickness of the RNFL + GCL increased at 3-days of reperfusion (Fig. 4H), followed by a decrease at 7-days of reperfusion (Fig. 4L). IPL thickness decreased at both 3-days and 7-days of reperfusion (Fig. 4I, M), while INL thickness decreased only at 7-days of reperfusion (Fig. 4N). We hypothesized that the initial swelling of the GCL in response to acute ischemia and hypoxia led to early thickening, followed by gradual thinning due to tissue dysfunction. This hypothesis was supported by the observation of RGCs loss and reduced GCC thickness evaluated by OCT at 7-days (Fig. 4E).

We observed by OCT that changes in retinal thickness became noticeable at 3-days post-UPOAO and exhibited a significant decrease at 7-days. HE results further revealed that alterations in inner retinal thickness constituted a substantial portion of the total retinal thickness during the early reperfusion stage after UPOAO.

Survival of retinal neural cells in UPOAO

Given that mouse retinal function heavily relies on rod cells and that scotopic (low light) vision is primarily governed by rod photoreceptors, the observed reductions in b-wave and OPs-wave amplitudes indicate impaired synaptic transmission within the inner retina [28]. Our OCT and HE findings suggested structural injuries within the inner retina. Notably, the thickness of the outer retinal layers in OCT remained unchanged. To further explore these observations, we conducted immunofluorescence staining to investigate alterations in major retinal cell types, primarily focusing on bipolar cells (BCs), photoreceptor cells, horizontal cells (HCs), and cholinergic amacrine cells.

BCs loss and photoreceptor cells survival

PKCα serves as a marker delineating BCs, with their cell bodies primarily located in the outermost part of the INL, axonal terminals extending into the innermost part of the IPL, and dendrites confined to the OPL (Fig. 5A, B). In the experimental UPOAO eyes, BCs did not show significant changes at the 3-days reperfusion mark but exhibited dramatic alterations by the 7-days reperfusion period (Fig. 5A2, B2). In particular, the immunostaining density of BCs dendritic and axonal arbors notably decreased at the 7-days reperfusion mark. BCs account for approximately 40% of INL cells in mice, and their somata and axonal processes form a substantial part of the INL and IPL, consistent with the thinning of the INL observed in both OCT and HE at 7-days (Fig. 4E, N). Notably, the decrease in the number of PKCα+ BCs at the 7-days reperfusion point coincided with the decrease in the b-waves amplitude in ERG (Fig. 3E), indicating functional interplay between BCs and other cells.

Changes in Bipolar Cells, Horizontal Cells, and Cholinergic Amacrine Cells in UPOAO Mice.

(A) Representative images of mouse retina co-stained with DAPI and PKCα at 3-days (B) Representative images of mouse retina co-stained with DAPI and PKCα at 7-days. (C) Quantification of PKCα fluorescence density at 3-days. (D) Quantification of PKCα fluorescence density at 7-days. n = 4. (E) Representative images of mouse retina co-stained with DAPI and Calbindin at 3-days. (F) Representative images of mouse retina co-stained with DAPI and Calbindin at 7-days. Horizontal cell somata are indicated by white asterisks. (G) Quantification of Calbindin fluorescence density at 3-days. (H) Quantification of Calbindin fluorescence density at 7-days. n = 3. (I) Representative images of mouse retina co-stained with DAPI and ChAT at 3-days. (J) Representative images of mouse retina co-stained with DAPI and ChAT at 7-days. Cholinergic amacrine cell somata are indicated by white asterisks. (K) Quantification of ChAT fluorescence density at 3-days. (L) Quantification of ChAT fluorescence density at 7-days. n = 3. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, t-test. Scale bar = 50 μm.

Recoverin marks the somata and outer segments of photoreceptor cells and is localized within the ONL (Fig. S3A, B). Interestingly, recoverin-positive photoreceptors remained relatively stable throughout the reperfusion periods, which aligns with the unchanged thickness of the outer retinal layers observed in OCT (Fig. S2C, D).

HCs and cholinergic amacrine cells loss

We evaluated HCs and cholinergic amacrine cells using calbindin and ChAT immunostaining, respectively. Calbindin immunostaining highlighted the somata of HCs within the INL, with terminal axon connections linearly positioned within the OPL (Fig. 5E, F). In the UPOAO experimental eyes, no significant change in the number of HCs was observed during the 3-days reperfusion period, while a notable reduction was observed after 7 days. By 7-days, a significant decrease in cell body numbers, a considerable reduction in axon density, and disrupted linear connections were observed (Fig. 5F2).

ChAT+ cell bodies were located in the GCL and INL, with dendrites forming two narrow stratified bands within the IPL. The immunofluorescence density of ChAT+ amacrine cells decreased notably after 3 days and even more prominently after 7 days (Fig. 5I, J). The fluorescence intensity of the two bands within the IPL markedly decreased and was nearly invisible after 7 days (Fig. 5J2).

Time course transcriptome analysis revealed features of different reperfusion periods in UPOAO

To explore the pathophysiological processes of UPOAO, we extracted retinas from both eyes for transcriptome sequencing. The samples included retinas from eyes subjected to 60 minutes of ischemia without reperfusion, 60 minutes of ischemia followed by 3-days of reperfusion, and 60 minutes of ischemia followed by 7-days of reperfusion, with the sham-treated eyes serving as controls.

In the non-perfusion group, 215 genes were upregulated and 204 genes were downregulated (Fig. 6A). GO enrichment analysis revealed that the DEGs were related to leukocyte migration, epidermis development, myeloid leukocyte migration and other cell migration pathways (Fig. 6B). The heatmap and box showed the upregulated and downregulated genes involved in immune cell migration-related pathways during the non-perfusion period in UPOAO (Fig. 6C). High-connectivity DEGs (‘hub genes’), such as Dusp1 and Fos, were also enriched in these pathways (Fig. S4A). GSEA also showed similar results (Fig. S4B-D). These results suggested that in the early stage of retinal ischemic injury, leukocytes from the microvasculature may infiltrate retinal tissue. More experimental validation will be performed to confirm this hypothesis.

Transcriptomic Features at Different Times of Reperfusion.

RNA-seq evaluation was performed at 0-day, 3-days, and 7-days reperfusion periods in UPOAO revealing enrichment in pathways related to immune cells migration, oxidative stress, and immune inflammation. (A, D, G) Volcano plots display differential expression genes (DEGs) between UPOAO and sham eyes in the no-perfusion group (A), 3-days perfusion group (D), and 7-days perfusion group (G), respectively. Red dots: significant upregulated genes, green and blue dots: significant downregulated genes, grey dots: stable expressed genes, adjusted P < 0.05. log2FC = 1. (B, E, H) Gene ontology (GO) analysis of differential genes in the non-perfusion group (B), 3-days group (E), and 7-days group (H). The DEGs between UPOAO and sham eyes were enriched in pathways associated with immune cells migration (0d), oxidative stress (3d), and immune inflammation (7d), respectively. (C, F, I) Heatmap displaying the top 100 downregulated and top 100 upregulated DEGs between UPOAO and sham at non-perfusion, 3-days, and 7-days reperfusion, respectively. The box represents the genes related to immune cells migration (C), oxidative stress (F), and immune inflammation (I). The ranking was determined by the magnitude of fold change. In each heatmap, the upper box represents the top 10 downregulated genes, while the lower box represents the top 10 upregulated genes.

In the 3-days reperfusion group, 372 genes were upregulated, and 170 genes were downregulated (Fig. 6D). GO enrichment analysis revealed that the DEGs at 3-days after reperfusion were related to energy metabolism, mitochondrial regulation, and oxidative stress pathways (Fig. 6E). The heatmap and box displayed the upregulated and downregulated genes involved in oxidative stress-related pathways at the 3-days reperfusion stage in UPOAO, including hub genes such as Mrpl18 and Mrps23 (Fig. 6F, Fig. S5A). GSEA revealed changes in pathways related to the binding membrane of organelles, the cell surface and the plasma membrane protein complex (Fig. S5B). Furthermore, 44 overlapping genes between the mitochondrial genes and DEGs at 3-days post-UPOAO were mainly related to mitochondrial transport and the mitochondrial tricarboxylic acid cycle (Fig. S5C, D). These results suggested that mitochondria and metabolism play significant roles in IRI.

In the 7-days reperfusion group, 429 genes were upregulated, and 310 genes were downregulated (Fig 6G). The DEGs at this stage were enriched in pathways related to immune effector processes, positive regulation of responses to external stimuli and inflammation (Fig. 6H). A heatmap was generated to show the upregulated and downregulated genes of immune inflammation-related pathways at 7-days after reperfusion in UPOAO, and most hub genes, such as Cd86, Cd48, Tlr4, and Tlr6 were also enriched in these pathways (Fig. 6I, Fig. S6A). GSEA also showed similar results (Fig S6B). Moreover, RT‒qPCR confirmed the upregulation of immune inflammation-related gene expression (Fig. S7). Significantly, 112 genes that overlapped between immune-related genes and DEGs at 7-days post-UPOAO were primarily associated with immune receptor activity and phagocytic vesicles (Fig. S6C, D). These results underscore the predominant role of immune responses during this stage.

To explore the associations between changes observed at 3-days and 7-days after reperfusion, we analyzed the DEGs and identified 17 overlapping genes (Fig. S8A), which were mainly associated with the nitric oxide biosynthetic process (Fig. S8B). Our findings suggest that the immune inflammatory response observed after 7-days of reperfusion may represent the cumulative effect of the acute oxidative stress response during the initial 3-days reperfusion period.

Leukocyte Infiltration and Microglial Activation

Time course transcriptome analysis underscored the predominant role of immune inflammatory responses during the reperfusion stage. Previous reports have indicated that the inflammatory response in retinal ischemia-reperfusion injury (RIRI) is orchestrated by peripheral immune cells and resident immune cells within the retina [29, 30]. In our study, we demonstrated the presence of leukocyte infiltration and microglial activation in the UPOAO model. Immunofluorescence staining for CD45 was performed on retinas at different time points post modelling to visualize the distribution and quantity of white blood cells (Fig. 7A-D). Laser confocal Z-plane projections confirmed minimal leukocyte infiltration in the retinas of sham eyes, which showed that CD45+ cells in the vascular lumen likely represented patrolling cells (Fig. 7A). A significant increase in CD45+ cells was observed in UPOAO model retinas after 1 day (Fig. 7E), with a progressive increase in quantity over time (Fig. 7F, G). Notably, the majority of CD45+ cells in UPOAO model retinas at 1-day displayed a morphology similar to that of vascular leukocytes (Fig. 7B), suggesting a large influx of peripheral white blood cells into the retinal tissue. In contrast, CD45+ cells at 3-days and 7-days of reperfusion exhibited a combination of amoeboid and branched morphologies (Fig. 7C, D), indicative of activated states.

Peripheral Leukocyte Infiltration and Retinal Resident Microglial Activation.

Rhodamine-labeled canavalin A was used for immediate cardiac perfusion to visualize blood vessels, followed by CD45 immunofluorescent staining to observe the relationship between blood vessels and CD45+ cells in sham (A), 1-day perfusion group (B), 3-days perfusion group (C), and 7-days perfusion group (D). The presence of CD45+ cells within blood vessels is indicated by white arrows. Scale bar = 50 μm. CD45+ cell counts were performed in whole retinas of 1-day (E), 3-days (F), 7-days (G). The cellular morphology and distribution of microglial cells in the superficial retina were assessed in 3-days (H) and 7-days (I). Activated microglial cells are indicated by white asterisks. Microglial cell counting was conducted in the superficial retina of 3-days (J) and 7-days (K). Data points for CD45+ cells were derived from four flat-mounted retinas, and data points for microglial cells were from five flat-mounted retinas. Data were presented as means ± s.e.m, *: p<0.05, **: p<0.01, ***: p<0.001, ****: p<0.0001, t-test. Scale bar = 50 μm.

Microglia, which are known to undergo activation and morphological changes in response to various insults, were examined in the superficial retinas of sham and UPOAO eyes. Immunofluorescence staining for Iba1 was conducted on flat-mounted retinas at 3-days and 7-days post-reperfusion. In sham eyes, Iba1+ microglia displayed small somas and elongated dendrites that were evenly distributed within the retina (Fig. 7H1, I1). Conversely, UPOAO retinas exhibited numerous activated Iba1+ microglia with enlarged somas, shortened dendrites, and an ameboid appearance (Fig. 7H2, I2). The number of Iba1+ microglia in UPOAO retinas was approximately five times greater than that in sham eyes at 3-days (Fig. 7J), with a further increase observed at 7-days (Fig. 7K), although this increase was less pronounced than that at the 3-days time point (Fig. 7J). These results indicated a potential reduction in immune-mediated inflammation over time.

RNA-seq comparison between UPOAO and extravascular occlusion models

To further study the characteristics of UPOAO, we analyzed the transcriptomes of two extravascular occlusion models: HIOP and UCCAO. The HIOP model induces ischemia through anterior chamber perfusion with normal saline (Fig. 8A), while the UCCAO model involves ligation of the unilateral common carotid artery (Fig. 8H). RNA-seq analysis of retinas subjected to 60 minutes of ischemia followed by 7-days of reperfusion in the HIOP model revealed enrichment of immune-related pathways, specifically highlighting increased activity in leukocyte-mediated immunity and regulation of immune effector processes (Fig. 8B). This finding was consistent with the results of GO analysis, which revealed associations of the overlapping DEGs between the HIOP and UPOAO models with toll-like receptor binding, T-cell receptor binding, and immune-related functions (Fig. 8C, D). To further explore these DEGs, we explored the remaining DEGs in the HIOP model in addition to the overlapping genes. GO analysis revealed that the remaining DEGs in the HIOP model were also enriched in immune responses, such as the T-cell receptor complex pathway, adaptive immune response, and B-cell-mediated immunity function (Fig. 8E). Remarkably, upon examining the remaining DEGs specific to the UPOAO model, GO analysis revealed DEGs distinctly related to lipid and steroid metabolic processes (Fig. 8F, G). This observation indicates that the UPOAO model, which is closely related to RAO, involves not only conventional immune responses but also unique regulation of lipid and steroid metabolism.

Transcriptomic Results Comparison between UPOAO, HIOP, and UCCAO Models.

RNA-seq comparison between UPOAO and extravascular occlusion models: HIOP and UCCAO. (A) Schematic illustration of the HIOP model. (B) GO analysis of differential expression genes (DEGs) in the HIOP model at 7-days perfusion. (C) Venn diagram indicating the overlapping DEGs (146 genes) between HIOP and UPOAO models, as well as the remaining DEGs specific to each group (593 genes in each group). (D) GO analysis of the overlapping DEGs between HIOP and UPOAO models. (E) GO analysis of the remaining DEGs in the HIOP model at 7-days perfusion, excluding the overlapping DEGs. (F) GO analysis of the remaining DEGs, excluding the overlapping DEGs. (G) Heatmap showing the inter-sample distribution of lipid and steroid-related DEGs from the analysis of the remaining DEGs in the UPOAO model. (H) Schematic illustration of the UCCAO model. (I) GO analysis of DEGs in the UCCAO model. (J) Hub genes identified through protein-protein interaction (PPI) network analysis of the DEGs in the UCCAO model. (K) Venn diagram indicating the overlapping DEGs between UCCAO and UPOAO models (15 genes). (L) Venn diagram indicating the presence of two overlapping DEGs between UCCAO DEGs and m6A-related genes.

Similarly, the UCCAO model exhibited associations with negative regulation of B-cell proliferation, lymphocyte activation, and related processes (Fig. 8I). Additionally, the hub genes identified through protein‒protein interaction (PPI) analysis are presented (Fig. 8J). However, the UCCAO model displayed minimal overlap with the UPOAO model, with only 15 overlapping DEGs identified in the Venn diagram analysis (Fig. 8K). Further analysis of the UCCAO DEGs revealed two genes (Mettl14 and YTHDF2) related to m6A modifications (Fig. 8L). Although arterial blood flow is occluded in both the UPOAO and UCCAO models, our results revealed that the pathophysiological processes of the two models were particularly different after 7-days of reperfusion.

Discussion

This study aimed to develop a more appropriate mouse model for investigating the ischemia-reperfusion process in RAO and unravelling the underlying pathophysiological mechanisms. We combined silicone wire embolization with carotid artery ligation to effectively block the blood supply to the retinal artery, closely mimicking the characteristics of the acute interruption of blood supply in RAO patients [31]. In the UPOAO model, a 60-minute ischemia period was sufficient to simulate the injury of major retinal neural cells, such as RGCs, BCs, HCs, and cholinergic amacrine cells, resulting in visual impairment. Moreover, histologic examination demonstrated thinning of the inner layer of the retina, especially the GCL. Time course transcriptome analysis revealed various pathophysiological processes related to immune cell migration, oxidative stress, and immune inflammation during non-reperfusion and reperfusion periods. The resident microglia within the retina and peripheral leukocytes that access the retina were markedly increased during reperfusion periods. Additionally, we compared the transcriptomic signatures of the UPOAO model with those of two commonly used ischemia-reperfusion mouse models (HIOP and UCCAO), highlighting the potential relevance of the UPOAO model to ocular vascular occlusive diseases.

The UPOAO mouse model effectively simulates the characteristic features of RAO observed in patients. RAO patients often exhibit distinctive patterns in dark-adapted ERGs, where b-waves are reduced while a-waves remain stable [32–34]. In our study, we observed almost the same phenomenon in ERG amplitude after 60 minutes of ischemia in the UPOAO model (Table S2). This observation was further confirmed by immunofluorescence staining of bipolar cells and photoreceptor cells. Importantly, a previous study linked the thinning of the inner retina with alterations in ERG in patients with RAO [35]. Our results showed that ischemic damage predominantly affected the inner retinal layers, while the outer layer was almost unaffected (Table S2). This distinction is attributed to the reliance of the inner layers on the retinal artery for blood supply, whereas the outer layers are supplied by the choroid. Additionally, histological examination revealed increased retinal NFL and GCL thickness and decreased IPL thickness after 3-days of reperfusion, indicating that ischemia-induced oedema had not completely subsided.

This study comprehensively explored the transcriptomic signature after ischemia and revealed that resident and peripheral immune cells may play major roles in pathological processes. In the non-reperfusion group, the transcriptional changes were primarily involved in the migration of immune cells such as leukocytes and neutrophils from blood vessels into ischemic retinal tissue. This process is closely linked to injury during ischemia and reperfusion, where leukocytes and neutrophils infiltrate neural tissue through the vascular endothelium [36, 37]. Our results revealed an increase in CD45+ leukocytes in the retina and retinal microvasculature, suggesting that these leukocytes were recruited from the bloodstream to the damaged retinal tissue following reperfusion. During leukocyte access to the retina, chemokines, chemokine receptors, adhesion molecules and cytoskeletal components play important roles in regulating endothelial permeability and facilitating leukocyte adhesion to endothelial cells [38, 39]. Our findings of related gene expression are consistent with this mechanism.

During the 3-days reperfusion period, we observed that an increased number of DEGs were significantly involved in the critical pathological response to oxidative stress in the context of IR tissues. The generation of reactive oxygen species (ROS) in mitochondria and subsequent oxidative stress are widely recognized as major causes of retinal cell damage induced by IR [40, 41]. This finding suggested that mitochondrial function, which is involved in oxidative stress processes, may play an important role in the pathophysiology of IRI during the early reperfusion stage. Oxidative stress can influence the permeability of the inner blood retina barrier (iBRB), allowing leukocytes to access retinal tissues [42]. The infiltration and activation of immune cells are recognized as the underlying causes of many retinal diseases, including ischemic ophthalmia [42, 43]. In the UPOAO model, the increase in CD45+ leukocytes and Iba1+ microglia in the retina after 3-days of reperfusion supports this viewpoint.

In the later stage of reperfusion (7-days), the DEGs were mainly enriched in immune regulation and inflammation. With more upregulated DEGs than downregulated ones, we hypothesize that the activation of the immune inflammatory response contributes to further IRI in retinal tissue. Previous studies have shown that the retina can elicit immunological responses during ischemia-reperfusion injury and that immune inflammation is an important phenomenon in the progression of this injury [30]. The excessive ROS generated by mitochondria in RGCs during the activation of inflammatory responses can damage cell structure and visual function [44]. Microglia, considered the primary resident immune cells, contribute to inflammatory responses and consequent neural damage [45]. Infiltrating leukocytes can activate resident microglia, forming a feedback loop that exacerbates inflammation [46]. Complex interactions between immune cells and retinal neurons after RIRI have been reported [47]. Interestingly, in our UPOAO model, the death of major retinal neural cells, such as RGCs, BCs, and HCs, was correlated with an increase in infiltrating leukocytes and resident microglia. Therefore, we considered that the immune response and neuroinflammation observed after 7-days of reperfusion may result from the cumulative effects of acute oxidative stress and the responses of resident and peripheral immune cells (Fig. 9, Table S2). In summary, we described the pathological processes at different time points and highlighted the important role of resident and peripheral immune cell responses in the UPOAO model. This finding offers valuable insights for subsequent screening of pathogenic genes and potential immunotherapy approaches.

The Characteristic Features of UPOAO Model during Ischemia-Reperfusion Periods.

Compared with the HIOP and UCCAO models, the UPOAO model is a more appropriate choice for studying retinal IRI in RAO. The HIOP model, widely used for investigating the pathogenesis of APACG, exhibits retinal degeneration features similar to those of the UPOAO model during both reperfusion periods [48]. In our UPOAO model, histologic and immunohistochemical analysis demonstrated that 60 minutes of ischemia followed by 3-days or 7-days of reperfusion can cause irreversible damage to retinal structure and visual function. Specifically, we observed distinct alterations, such as thinning of the inner retina, substantial RGCs apoptosis, and decreased b-wave amplitude. Multiple studies on the HIOP model have reported consistent results [45, 49, 50]. However, some typical phenotypes in the HIOP model, such as thinning of the outer nuclear layer and the entire neuroretina and decreased a-wave amplitude, were not observed in the UPOAO model. Moreover, our transcriptome sequencing revealed a specific subset of DEGs unique to the UPOAO model, including Apoe, Abca1, Ldlr, Cyp39a1, and Bmp6. Notably, these DEGs were significantly enriched in pathways associated with lipid and steroid synthesis. Steroid and lipid metabolism homeostasis is disrupted under conditions of oxidative stress and inflammation [51]. In the UPOAO model, the observed lipid and steroid biological processes may result from the cumulative pathological responses triggered by IRI, as observed in the HIOP model. In the UCCAO model, RNA-seq revealed that DEGs were enriched in two main pathways: 1) epigenetic modification-related pathways, including nucleobase-containing compound catabolic process, RNA catabolic process, mRNA catabolic process and N6-methyladenosine (m6A) modification; and 2) cell death pathways, including regulation of autophagosome assembly, negative regulation of neuron death, and negative regulation of the neuronal apoptotic process. Epigenetic mechanisms may play a key role in the pathophysiology of ocular disease [52]. m6A, one of the most common RNA modifications, has been reported to regulate cell death processes, including apoptosis and autophagy, in the pathological process of IRI [53]. These results suggest that epigenetic mechanisms may significantly influence cell death during the 7-days reperfusion period in the UCCAO model. Lee et al. evaluated the characteristics of UCCAO without reperfusion using visual, histological and immunohistochemical approaches. Their results revealed delayed perfusion of the ipsilateral retina, thinning of the inner retinal layer 10 weeks after surgery, and a dramatic decrease in the amplitudes of b-waves on day 14 after UCCAO [15, 54]. This finding suggested that UCCAO primarily represents a model of retinal hypoperfusion injury and may not effectively reflect acute ischemic-induced structural and functional damage in RAO (Fig. 9). These results led us to consider the UPOAO model an effective experimental model for studying the pathological processes underlying acute ischemia and IRI.

Our research has certain limitations that should be acknowledged. First, our study focused on the changes that occurred within 60 minutes of ischemia and within the first 7-days of reperfusion in the UPOAO model. Further exploration is needed to understand the changes induced by longer reperfusion periods. Additionally, we proposed possible pathological mechanisms, including iBRB damage, oxidative stress and immune inflammation, mainly based on the enrichment results of DEGs at three reperfusion time points. More molecular experimental validation is required to substantiate these proposed mechanisms.

In summary, by thoroughly exploring the injury to major retinal neural cells, visual impairment and pathophysiological changes in the UPOAO model, we confirmed that this model can effectively simulate the acute ischemia-reperfusion processes in RAO. It serves as an ideal mouse model for investigating the underlying pathological mechanisms of ischemia and reperfusion. Furthermore, our UPOAO model holds great promise as a novel model for studies on pathogenic genes and potential therapeutic interventions for RAO.

Supporting information